---
title: "scMut cluster size estimation"
author: "Qun Li"
date: "20 Mar 2025 ( 09:42:46 )"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  word_document:
    toc: true
    toc_depth: '3'
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# How to generate data
run 1_estimateClustersize.sh

# Data source
SomaMutDB 2.0: A comprehensive database for exploring somatic mutations and their functional impact in normal human tissues
Dataset1
    A body map of somatic mutagenesis in morphologically normal human tissues
Dataset2
    Single-cell whole-genome sequencing reveals the functional landscape of somatic mutations in B lymphocytes across the human lifespan
Dataset3
    Somatic genomic changes in single Alzheimer’s disease neurons


## Env settings
```{r envSettings}
rm(list = ls())
setwd("/Users/liqun/Desktop/scMutrace-tutorial/ReproducibilityAnalysis/Figure1/")
```

# libraries
```{r libraries}
library(ggplot2)
library(data.table)
```


```{r loadData}
meta_somatic <- readRDS("../Data/randomCluster/meta_somatic.rds")
df_somatic <- readRDS("../Data/randomCluster/df_somatic.rds")

#saveRDS(meta_somatic, "../Data/randomCluster/meta_somatic.rds")
#saveRDS(df_somatic, "../Data/randomCluster/df_somatic.rds")

colnames(df_somatic) <- c("sampleID","dist")
df_somatic_meta <- merge(df_somatic, meta_somatic, by = "sampleID")
df_somatic_meta_over10 <- df_somatic_meta[df_somatic_meta$dist >= 10,]
length(df_somatic_meta_over10$sample)/length(df_somatic_meta$sample)
ecdf(df_somatic_meta$dist)(10)

df_WGS <- readRDS("../Data/randomCluster/df_WGS.rds")
#saveRDS(df_WGS, "../Data/randomCluster/df_WGS.rds")
colnames(df_WGS) <- c("sample","dist")
df_WGS_over10 <- df_WGS[df_WGS$dist >= 10,]
length(df_WGS_over10$sample)/length(df_WGS$sample)
ecdf(df_WGS$dist)(10)
```

# plot
```{r plot}
threshold_somatic <- quantile(df_somatic_meta$dist, 0.005, na.rm = TRUE)
ecdf_somatic <- ecdf(df_somatic_meta$dist)
quantile_10 <- ecdf_somatic(10)

desired_order <- c("Bronchus", "Cardia", "Colon", "Duodenum", "Liver",
                   "Oesophagus", "Pancreas", "Rectum", "Stomach", "B",
                   "Neuron_MDA", "Neuron_PTA")

df_somatic_meta$Tissue <- factor(df_somatic_meta$Tissue, levels = desired_order)

ggplot(df_somatic_meta, aes(x = dist)) +
  geom_histogram(bins = 100, fill = "grey", alpha = 0.7) +
  scale_x_log10() +
  geom_vline(xintercept = 10,
             linetype = "dashed",
             color = "red",
             size = 1) +
  facet_wrap(~Tissue, scales = "free_y") +
  theme_classic(base_size = 14)

ggplot(df_somatic_meta, aes(x = dist)) +
  geom_histogram(bins = 100, fill = "grey", alpha = 0.7) +
  scale_x_log10() +
  geom_vline(xintercept = 10,
             linetype = "dashed",
             color = "red",
             size = 1) +
  theme_classic(base_size = 14)

my_colors <- c("#1f78b4", "#33a02c", "#e31a1c", "#ff7f00",
               "#6a3d9a", "#b15928", "#a6cee3", "#b2df8a",
               "#fb9a99", "#fdbf6f", "#cab2d6", "#ffff99")

p_somatic_numDis <- ggplot(df_somatic_meta, aes(x = log10(dist), color = Tissue)) +
  geom_density(aes(alpha = 0.3, size = 1, y = ..count..)) +
  #geom_density(aes(alpha = 0.4)) +
  geom_vline(xintercept = 1,          
             linetype = "dashed",
             color = "red",
             size = 1) +
  theme_classic(base_size = 14) +     
  xlab("Mutation distance (bp, log scale)") +
  ylab("Density") +
  ggtitle("Somatic mutation distance distribution by Tissue") +
  scale_color_manual(values = my_colors) 

threshold_wgs <- quantile(df_WGS$dist, 0.005, na.rm = TRUE)

threshold_wgs <- quantile(df_WGS$dist, 0.005, na.rm = TRUE)
ecdf_wgs <- ecdf(df_WGS$dist)
quantile_10 <- ecdf_wgs(10)

ggplot(df_WGS, aes(x = log10(dist))) +
  geom_histogram(bins = 100, fill = "grey", alpha = 0.7) +
  geom_vline(xintercept = 1,
             linetype = "dashed",
             color = "red",
             size = 1) +
  theme_classic(base_size = 14)

p_WGS_numDis <- ggplot(df_WGS, aes(x = log10(dist), color = sample)) +
  geom_density(aes(alpha = 0.3, size = 1, y = ..count..)) +    
  geom_vline(xintercept = 1,          
             linetype = "dashed",
             color = "red",
             size = 1) +
  theme_classic(base_size = 14) +     
  xlab("Mutation distance (bp, log scale)") +
  ylab("Density") +
  ggtitle("Somatic mutation distance distribution by Tissue") +
  scale_color_manual(values = my_colors)

p_somatic_numDis
p_WGS_numDis
#ggsave("p_somatic_numDis.pdf", p_somatic_numDis, width = 7.21, height = 3.67)
#ggsave("p_wgs_numDis.pdf", p_WGS_numDis, width = 7.21, height = 3.67)
```


# bin size
```{r binsize}
df_somatic_pos <- readRDS("../Data/randomCluster/df_somatic_pos.rds")
df_wgs_pos <- readRDS("../Data/randomCluster/df_wgs_pos.rds")

#saveRDS(df_somatic_pos, "../Data/randomCluster/df_somatic_pos.rds")
#saveRDS(df_wgs_pos, "../Data/randomCluster/df_wgs_pos.rds")

df_somatic_pos[, bin10 := floor(pos/10)*10]
df_wgs_pos[, bin10 := floor(pos/10)*10]

window_counts <- df_somatic_pos[, .N, by=.(sampleID, chr, bin10)]
window_counts_wgs <- df_wgs_pos[, .N, by=.(sampleID, chr, bin10)]

mutation_dist <- as.data.table(table(window_counts$N))
setnames(mutation_dist, c("N","N_windows"))
mutation_dist[, N := as.integer(as.character(N))] 

mutation_dist$mutation_number_cat <- ifelse(mutation_dist$N >= 3, ">=3", as.character(mutation_dist$N))
mutation_dist <- aggregate(N_windows ~ mutation_number_cat, data = mutation_dist, sum)
mutation_dist$mutation_number_cat <- factor(mutation_dist$mutation_number_cat, levels = c("0", "1", "2", ">=3"))

mutation_dist_wgs <- as.data.table(table(window_counts_wgs$N))
setnames(mutation_dist_wgs, c("N","N_windows"))
mutation_dist_wgs[, N := as.integer(as.character(N))] 

mutation_dist_wgs$mutation_number_cat <- ifelse(mutation_dist_wgs$N >= 3, ">=3", as.character(mutation_dist_wgs$N))
mutation_dist_wgs <- aggregate(N_windows ~ mutation_number_cat, data = mutation_dist_wgs, sum)
mutation_dist_wgs$mutation_number_cat <- factor(mutation_dist_wgs$mutation_number_cat, levels = c("0", "1", "2", ">=3"))

p_somatic_num10bp <- ggplot(mutation_dist, aes(x=mutation_number_cat, y=N_windows)) +
  geom_bar(stat="identity", fill="grey") +
  theme_classic() +
  xlab("Number of mutations per 10bp window") +
  ylab("Number of windows") +
  ggtitle("10bp window mutation count distribution")

p_wgs_num10bp <- ggplot(mutation_dist_wgs, aes(x=mutation_number_cat, y=N_windows)) +
  geom_bar(stat="identity", fill="grey") +
  theme_classic() +
  xlab("Number of mutations per 10bp window") +
  ylab("Number of windows") +
  ggtitle("10bp window mutation count distribution")

table(window_counts$N)
table(window_counts_wgs$N)

ecdf(window_counts$N)(2)
ecdf(window_counts_wgs$N)(2)

p_somatic_num10bp
p_wgs_num10bp

#ggsave("p_somatic_num10bp.pdf", p_somatic_num10bp, width = 7.21, height = 3.67)
#ggsave("p_wgs_num10bp.pdf", p_wgs_num10bp, width = 7.21, height = 3.67)
```


# r estimate
```{r estimate}
n <- 10
p <- 0.0075
k <- 0:n
prob <- dbinom(k, size = n, prob = p)

df_prob <- data.frame(
  mutation_number = k,
  Prob = prob
)

df_prob$mutation_number_cat <- ifelse(df_prob$mutation_number >= 3, ">=3", as.character(df_prob$mutation_number))
df_prob_plot <- aggregate(Prob ~ mutation_number_cat, data = df_prob, sum)

df_prob_plot$mutation_number_cat <- factor(df_prob_plot$mutation_number_cat, levels = c("0", "1", "2", ">=3"))

p_prob <- ggplot(df_prob_plot, aes(x = factor(mutation_number_cat), y = Prob)) +
  geom_bar(stat = "identity", fill = "grey") +
  labs(x = "Mutation number", y = "Probability",
       title = "10 bp mutation number distribution") +
  theme_classic()

p_prob

#ggsave("p_prob.pdf", p_prob, width = 7.21, height = 3.67)
```
