---
title: "accumulatedMutations_MEF"
author: "Qun Li"
date: "19 Aug 2025 at 10:18:05"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

# Env
```{r env}
rm(list = ls())
setwd("/Users/liqun/Desktop/scMutrace-tutorial/ReproducibilityAnalysis/Figure5/")
```

# Libraries
```{r libraries}
library(FSA)
library(ineq)
library(vegan)
library(dplyr)
library(tidyr)
library(ggpubr)
library(Seurat)
library(igraph)
library(ggraph)
library(webshot)
library(ggplot2)
library(ggpmisc)
library(stringr)
library(corrplot)
library(pheatmap)
library(reshape2)
library(tidygraph)
library(networkD3)
library(tidyverse)
library(patchwork)
```

# Def function
```{r function}
retriveCellsByLocation <- function(CSM, locList){
  CSM$CellTag <- rownames(CSM)
  CSM_tmp <- CSM[, c("CellTag",locList)]
  data_MutPerCell_CSM_tmp <- data.frame(
    CellTag = rownames(CSM_tmp),
    MutNum = rowSums(CSM_tmp == "Mut"),
    WTNum = rowSums(CSM_tmp == "WT")
  )
  cells_Total <- data_MutPerCell_CSM_tmp[rowSums(data_MutPerCell_CSM_tmp[,c("MutNum", "WTNum")]) > 0, ]$CellTag
  cells_WT <-    data_MutPerCell_CSM_tmp[data_MutPerCell_CSM_tmp[,"WTNum"] > 0,]$CellTag
  cells_Mut <-   data_MutPerCell_CSM_tmp[data_MutPerCell_CSM_tmp[,"MutNum"] > 0,]$CellTag
  cells <- list(Total = cells_Total, Mut = cells_Mut, WT = cells_WT)
  return(cells)
}

retriveLocsByCells <- function(CSM, cellList){
  CSM$CellTag <- rownames(CSM)
  CSM_tmp <- CSM[CSM$CellTag %in% cellList,]
  data_CellPerMut_CSM_tmp <- data.frame(
    Location = colnames(CSM_tmp),
    CellMutNum = colSums(CSM_tmp == "Mut"),
    CellWTNum = colSums(CSM_tmp == "WT")
  )
  pos_Total <- data_CellPerMut_CSM_tmp[rowSums(data_CellPerMut_CSM_tmp[, c("CellMutNum", "CellWTNum")]) > 0,]$Location
  pos_Mut <- data_CellPerMut_CSM_tmp[data_CellPerMut_CSM_tmp[, "CellMutNum"] > 0 ,]$Location
  pos_WT <- data_CellPerMut_CSM_tmp[data_CellPerMut_CSM_tmp[, "CellWTNum"] > 0 ,]$Location
  pos <- list(Total = pos_Total, Mut = pos_Mut, WT = pos_WT)
  return(pos)
}

cal_MultiSpe <- function(matrixData) {
  
  res <- data.frame(
    Shannon = vegan::diversity(matrixData, index = "shannon"),
    Simpson = vegan::diversity(matrixData, index = "simpson"),
    Gini = apply(matrixData, 1, function(x) ineq(x + 0.00001, type = "Gini")),
    sumMut = rowSums(matrixData),
    Tau = NA
  )
  
  # Tau = sum(1 - (x / max(x))) / (n - 1)
  res$Tau <- apply(matrixData, 1, function(x) {
    if (all(x == 0)) return(NA)
    x <- x / max(x)
    sum(1 - x) / (length(x) - 1)
  })
  
  rownames(res) <- rownames(matrixData)
  return(res)
}

cal_scMutScore <- function(data_cell_mut, data_mut_time){
  # data_cell_mut [cell * muts]
  # data_mut_time [muts * tissues/celltype/days]
  # data_cell_time_matrix
  replacement_map <- c("NoCall" = 0, "WT" = 0, "Mut" = 1)
  
  data_cell_mut_num <- as.data.frame(lapply(data_cell_mut, function(x) {
    ifelse(x %in% names(replacement_map), replacement_map[x], x)
  }))

  rownames(data_cell_mut_num) <- rownames(data_cell_mut)
  print(dim(data_cell_mut_num))
  print(dim(data_mut_time))
  
  data_cell_time_matrix <- as.matrix(data_cell_mut_num) %*% as.matrix(data_mut_time)
  rownames(data_cell_time_matrix) <- rownames(data_cell_mut)
  colnames(data_cell_time_matrix) <- colnames(data_mut_time)
  
  muts_total <- unique(rownames(data_mut_time))

  # get different tissues/time/celltype names
  col_names_data <- colnames(data_cell_time_matrix)
  
  col_lenghth <- length(col_names_data)
  
  col_coefficient_fix <- setNames(
    rep(1/col_lenghth, col_lenghth),
    col_names_data
  )

  data_cell_time_matrix <- data_cell_time_matrix[,col_names_data]
  
  # Gini
  res_index <- cal_MultiSpe(data_cell_time_matrix)
  
  res_diversity <- as.data.frame((data_cell_time_matrix > 1) * rep(col_coefficient_fix[colnames(data_cell_time_matrix)], each = nrow(data_cell_time_matrix)))
  res_diversity$Diversity <- rowSums(res_diversity)

  res_diversity$Diversity[is.na(res_diversity$Diversity)] <- 0.00001
  
  res_index$Gini[is.na(res_index$Gini)] <- 0.00001
  res_index$Tau[is.na(res_index$Tau)] <- 0.00001

  res_index$GiniTau <- res_index$Gini * res_index$Tau

  res_index$scMutScore <- 2 / (
    (1 / (res_index$GiniTau + 0.00001)) + 
    (1 / (res_diversity$Diversity + 0.00001))
  )
  
  res_index$Diversity <- res_diversity$Diversity
  
  res_index$scMutScore <- scales::rescale(res_index$scMutScore, to = c(0, 1))
  
  res <- list(diversity = res_diversity, score = res_index, matrixD = data_cell_time_matrix)
  return(res)
}

propagate_mutations <- function(data, time_points) {
  for (i in 1:nrow(data)) {
    mut_time <- which(data[i, time_points] == "Mut")
    if (length(mut_time) > 0) {
      first_mut <- min(mut_time)
      data[i, time_points[first_mut:length(time_points)]] <- "Mut"
    }
  }
  return(data)
}
```


# Load Data
```{r loadData}
MEF_CSM <- readRDS("../Data/MEF/MEF_CSM.rds")
MEF_CSM_hf1 <- MEF_CSM$hf1
MEF_CSM_hf2 <- MEF_CSM$hf2

MEF_SSM <- readRDS("../Data/MEF/MEF_SSM.rds")
MEF_SSM_hf1 <- MEF_SSM$hf1
MEF_SSM_hf2 <- MEF_SSM$hf2

MEF_Meta <- readRDS("../Data/MEF/MEF_Meta.rds")
MEF_Meta_hf1 <- MEF_Meta$hf1
MEF_Meta_hf2 <- MEF_Meta$hf2

# retrieved from MPG database, see https://github.com/QunATCG/scMutrace-tutorial/tree/main/AdvancedFunctions/MPDAPI
MPG_data <- read.csv("../Data/MEF/MPG_combined.csv")
```

# Observed Accumulated Mutation Num
```{r ObservedAccumulated Mutation Num}
Reps_sequencing <- c("Rep1_all", "Rep2_all")
Days_sequencing <- c("d6", "d9", "d12", "d15", "d21", "d28")

col_hf1_names <- c("hf1.d0", "hf1.d3", "hf1.d6", "hf1.d9", "hf1.d12", "hf1.d15", "hf1.d21", "hf1.d28")
col_hf2_names <- c("hf2.d0", "hf2.d3", "hf2.d6", "hf2.d9", "hf2.d12", "hf2.d15", "hf2.d21", "hf2.d28")

data_accMutNum_res <- as.data.frame(matrix(nrow = length(Days_sequencing), ncol = length(Reps_sequencing)))
colnames(data_accMutNum_res) <- Reps_sequencing
rownames(data_accMutNum_res) <- Days_sequencing

for (i in rownames(data_accMutNum_res)){
  data_accMutNum_res[i, "Rep1_all"] <- length(MEF_SSM_hf1[MEF_SSM_hf1[, paste0("hf1.", i)] == "Mut",]$mutPos)
  data_accMutNum_res[i, "Rep2_all"] <- length(MEF_SSM_hf2[MEF_SSM_hf2[, paste0("hf2.", i)] == "Mut",]$mutPos)
}

data_accMutNum_res$Days <- rownames(data_accMutNum_res)
data_accMutNum_res_plot <- melt(data_accMutNum_res)
colnames(data_accMutNum_res_plot) <- c("Days", "Group", "MutNum")

data_accMutNum_res_plot$Days <- factor(data_accMutNum_res_plot$Days, levels = c("d6", "d9", "d12", "d15", "d21", "d28"))

p_reps_somaticNum <- ggplot(data_accMutNum_res_plot, aes(x = Days, y = MutNum, colour = Group, fill = Group)) +
  geom_point(shape = 21, size = 3, stroke = 0.8) +
  ylab("Number of somatic mutations") +
  scale_colour_manual(values = c("#4d4d4d", "#b2182b")) +
  scale_fill_manual(values = c("#4d4d4d", "#b2182b")) +
  theme_classic()

p_reps_somaticNum

replacement_map <- c("NoCall" = 0, "WT" = 0, "Mut" = 1)

MEF_SSM_hf1_hf2 <- merge(MEF_SSM_hf1[,c("mutPos", col_hf1_names)], MEF_SSM_hf2[,c("mutPos", col_hf2_names)], by = "mutPos", all = T)
MEF_SSM_hf1_hf2[is.na(MEF_SSM_hf1_hf2)] <- "NoCall"

MEF_SSM_hf1_hf2_num <- as.data.frame(lapply(MEF_SSM_hf1_hf2, function(x) {
  ifelse(x %in% names(replacement_map), replacement_map[x], x)
}))

rownames(MEF_SSM_hf1_hf2_num) <- MEF_SSM_hf1_hf2$mutPos

# propagate_mutations
p_accSomaticMut_heatmap_observed <- pheatmap(MEF_SSM_hf1_hf2_num[,c(2:17)], show_rownames = F, color = c("#4d4d4d", "#b2182b"), cluster_cols = FALSE)
p_accSomaticMut_heatmap_observed
```

# Imputated Accumulated Mutation Num
```{r ImputatedAccumulatedMutationNum}
color_settings <- c(d0 = "#FFDC91FF", d3 = "#7876B1FF", d6 = "#4575b4",
                    d9 = "#91bfdb", d12 = "#e0f3f8", d15 = "#fee090",
                    d21 = "#fc8d59", d28 = "#BC3C29FF")

color_settings_2 <- c(Day_0 = "#FFDC91FF", Day_3 = "#7876B1FF", Day_6 = "#4575b4",
                    Day_9 = "#91bfdb", Day_12 = "#e0f3f8", Day_15 = "#fee090",
                    Day_21 = "#fc8d59", Day_28 = "#BC3C29FF")

Reps_sequencing <- c("Rep1_all", "Rep2_all")
Days_sequencing <- c("d0", "d3", "d6", "d9", "d12", "d15", "d21", "d28")

col_hf1_names <- c("hf1.d0", "hf1.d3", "hf1.d6", "hf1.d9", "hf1.d12", "hf1.d15", "hf1.d21", "hf1.d28")
col_hf2_names <- c("hf2.d0", "hf2.d3", "hf2.d6", "hf2.d9", "hf2.d12", "hf2.d15", "hf2.d21", "hf2.d28")

MEF_SSM_hf1_impute <- propagate_mutations(MEF_SSM_hf1, col_hf1_names)
MEF_SSM_hf2_impute <- propagate_mutations(MEF_SSM_hf2, col_hf2_names)

data_accMutNum_res <- as.data.frame(matrix(nrow = length(Days_sequencing), ncol = length(Reps_sequencing)))
colnames(data_accMutNum_res) <- Reps_sequencing
rownames(data_accMutNum_res) <- Days_sequencing

for (i in rownames(data_accMutNum_res)){
  data_accMutNum_res[i, "Rep1_all"] <- length(MEF_SSM_hf1_impute[MEF_SSM_hf1_impute[, paste0("hf1.", i)] == "Mut",]$mutPos)
  data_accMutNum_res[i, "Rep2_all"] <- length(MEF_SSM_hf2_impute[MEF_SSM_hf2_impute[, paste0("hf2.", i)] == "Mut",]$mutPos)
}

data_accMutNum_res$Days <- rownames(data_accMutNum_res)
data_accMutNum_res_plot <- melt(data_accMutNum_res)
colnames(data_accMutNum_res_plot) <- c("Days", "Group", "MutNum")

data_accMutNum_res_mean_plot <- aggregate(MutNum ~ Days, data = data_accMutNum_res_plot, FUN = mean)

data_accMutNum_res_mean_plot$Days <- factor(data_accMutNum_res_mean_plot$Days, levels = c("d0", "d3", "d6", "d9", "d12", "d15", "d21", "d28"))

p_reps_somaticNum <- ggplot(data_accMutNum_res_mean_plot, aes(x = Days, y = MutNum, fill = Days)) +
  geom_col(position = "dodge") +
  theme_minimal(base_size = 14) +
  labs(title = "Mean Somatic mutations at different days",
       x = "Day",
       y = "Number of Somatic mutations") +
  scale_fill_manual(values = color_settings) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_reps_somaticNum

replacement_map <- c("NoCall" = 0, "WT" = 0, "Mut" = 1)

MEF_SSM_hf1_hf2_impute <- merge(MEF_SSM_hf1_impute[,c("mutPos", col_hf1_names)], MEF_SSM_hf2_impute[,c("mutPos", col_hf2_names)], by = "mutPos", all = T)
MEF_SSM_hf1_hf2_impute[is.na(MEF_SSM_hf1_hf2_impute)] <- "NoCall"

MEF_SSM_hf1_hf2_impute_num <- as.data.frame(lapply(MEF_SSM_hf1_hf2_impute, function(x) {
  ifelse(x %in% names(replacement_map), replacement_map[x], x)
}))

rownames(MEF_SSM_hf1_hf2_impute_num) <- MEF_SSM_hf1_hf2_impute$mutPos

# propagate_mutations
p_accSomaticMut_heatmap_impute <- pheatmap(MEF_SSM_hf1_hf2_impute_num[,c(2:17)], show_rownames = F, color = c("#4d4d4d", "#b2182b"), cluster_cols = FALSE)
p_accSomaticMut_heatmap_impute


# cell number
hf1_cellNum <- as.data.frame(table(MEF_Meta_hf1$Day)) %>%
  mutate(Sample = "hf1")
hf2_cellNum <- as.data.frame(table(MEF_Meta_hf2$Day)) %>%
  mutate(Sample = "hf2")

hf_cellNum <- bind_rows(hf1_cellNum, hf2_cellNum)
colnames(hf_cellNum) <- c("Day", "Count", "Sample")

hf_cellNum_dayMean <- aggregate(Count ~ Day, data = hf_cellNum, FUN = mean)

days_order_row <- c("Day_0", "Day_3", "Day_6", "Day_9", "Day_12",
                    "Day_15", "Day_21", "Day_28")

hf_cellNum_dayMean <- hf_cellNum_dayMean %>% filter(Day %in% days_order_row)
hf_cellNum_dayMean$Day <- factor(hf_cellNum_dayMean$Day, levels = days_order_row)

color_settings_2 <- c(Day_0 = "#FFDC91FF", Day_3 = "#7876B1FF", Day_6 = "#4575b4",
                    Day_9 = "#91bfdb", Day_12 = "#e0f3f8", Day_15 = "#fee090",
                    Day_21 = "#fc8d59", Day_28 = "#BC3C29FF")

p_hf_cellNum <- ggplot(hf_cellNum_dayMean, aes(x = Day, y = Count, fill = Day)) +
  geom_col(position = "dodge") +
  theme_minimal(base_size = 14) +
  labs(title = "Mean Cell counts at different days",
       x = "Day",
       y = "Cell count") +
  scale_fill_manual(values = color_settings_2) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_hf_cellNum
```

# Accumulated Mutation Num normalized by cell number
```{r randomCells}
color_settings_randomCells <- c(hf1.d3 = "#7876B1FF", hf1.d6 = "#4575b4", hf1.d9 = "#91bfdb", hf1.d12 = "#e0f3f8", hf1.d15 = "#fee090",  hf1.d21 = "#fc8d59", hf1.d28 = "#BC3C29FF",
                                hf2.d3 = "#7876B1FF", hf2.d6 = "#4575b4", hf2.d9 = "#91bfdb", hf2.d12 = "#e0f3f8", hf2.d15 = "#fee090",  hf2.d21 = "#fc8d59", hf2.d28 = "#BC3C29FF")

color_settings_randomCells_day <- c(d6 = "#4575b4", d9 = "#91bfdb", d12 = "#e0f3f8", d15 = "#fee090",  d21 = "#fc8d59", d28 = "#BC3C29FF")

totalCells <- c(rownames(MEF_CSM_hf1), rownames(MEF_CSM_hf2))

selectRandomNum <- c(10, 50, 100, 200, 300, 400, 500, 1000, 2000, 3000, 4000, 5000, 6000, 8000)

dataSelect <- c("hf1.d6", "hf1.d9", "hf1.d12", "hf1.d15", "hf1.d21", "hf1.d28", 
                "hf2.d6", "hf2.d9", "hf2.d12", "hf2.d15", "hf2.d21", "hf2.d28")

days <- c("d6", "d9", "d12", "d15", "d21", "d28")
reps <- 20
data_random_counts <- as.data.frame(matrix(nrow = (length(selectRandomNum) * length(dataSelect) * reps), ncol = 4))
colnames(data_random_counts) <- c("ID", "reps", "selectNum", "MutNum")

set.seed(2025)
t = 0
for (i in dataSelect){
  dataCells <- totalCells[grep(i, totalCells)]
  selectNum <- selectRandomNum
  for (j in 1:reps){
    for (k in selectNum){
      dataCells_random <- sample(dataCells, k, replace = TRUE)
      hf1_selected <- dataCells_random[dataCells_random %in% rownames(MEF_CSM_hf1)]
      hf2_selected <- dataCells_random[dataCells_random %in% rownames(MEF_CSM_hf2)]
      
      data_random_hf1 <- MEF_CSM_hf1[hf1_selected,]
      data_random_hf2 <- MEF_CSM_hf2[hf2_selected,]
      
      num_hf1 <- sum(colSums(data_random_hf1 == "Mut") > 0)
      num_hf2 <- sum(colSums(data_random_hf2 == "Mut") > 0)
      t = t + 1
      if(t %% 400 == 0){
        print(paste0("processed ", t, " iterations"))
      }
      data_random_counts[t,]$ID <- i
      data_random_counts[t,]$reps <- j
      data_random_counts[t,]$selectNum <- k
      data_random_counts[t,]$MutNum <- num_hf1 + num_hf2
    }
  }
}

data_random_counts <- separate(data_random_counts,col = "ID",into = c("Timecourse", "Day"), sep = "\\.")
data_random_counts$ID <- paste(data_random_counts$Timecourse, data_random_counts$Day, sep = ".")
data_random_counts$Day <- factor(data_random_counts$Day, levels = days)
data_random_counts$ID <- factor(data_random_counts$ID, levels = dataSelect)
data_random_counts_plot <- data_random_counts[data_random_counts$selectNum %in% c(10, 50, 100, 200, 300, 400, 500, 1000, 2000, 3000, 4000, 5000),]

p_random_somaticNum <- ggplot(data_random_counts_plot,
       aes(x = Day, y = MutNum, color = factor(selectNum))) + 
  geom_jitter(width = 0.2, alpha = 0.5, size = 1) +   
  facet_wrap(~ Timecourse, ncol = 1, scales = "free_y") +
  ylab("Number of Somatic Mutations") +
  xlab("Day") +
  ylim (0, 250) +
  labs(color = "SelectNum") +
  scale_color_manual(values = c(
    "10" = "#1a1a1a",
    "50" = "#4d4d4d",
    "100" = "#878787",
    "200" = "#fff5f0",
    "300" = "#fee0d2",
    "400" = "#fcbba1",
    "500" = "#fc9272",
    "1000" = "#fb6a4a",
    "2000" = "#ef3b2c",
    "3000" = "#cb181d",
    "4000" = "#a50f15",
    "5000" = "#67000d"
  )) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_random_somaticNum
```

# MPGgenotypecheck
```{r checkMPGGenotype}
# check position
MPG_data <- read.csv("../Data/MEF/MPG_combined.csv")
MEF_detected_pos <- unique(union(MEF_SSM_hf1$position_cor, MEF_SSM_hf2$position_cor))
MEF_detected_pos[MEF_detected_pos %in% MPG_data$position_cor]

MPG_data <- MPG_data[,c(3,8:44)]
MPG_data[MPG_data == ""] <- NA

MEF_SSM_refAlt <- rbind(MEF_SSM_hf1[,c("position_cor", "ref", "alt")], MEF_SSM_hf2[,c("position_cor", "ref", "alt")])
MPG_data_refAlt <- merge(MEF_SSM_refAlt, MPG_data, by = "position_cor", all.x = T)
MPG_data_refAlt[is.na(MPG_data_refAlt)] <- 0

MPG_data_refAlt[ , 4:ncol(MPG_data_refAlt)] <- apply(MPG_data_refAlt[ , 4:ncol(MPG_data_refAlt)], 2, function(col) {
  refCheck <- MPG_data_refAlt[,"ref"]
  ifelse(col != 0 & col != refCheck, 1, 0)
})

MPG_data_refAlt <- MPG_data_refAlt[!duplicated(MPG_data_refAlt$position_cor),]
rownames(MPG_data_refAlt) <- MPG_data_refAlt$position_cor

MPG_data_refAlt$MutNum <- rowSums(MPG_data_refAlt[,4:40])
MPG_data_refAlt_sorted <- MPG_data_refAlt[order(-MPG_data_refAlt$MutNum), ]

colnames_order <- c("C57BL_6J", "C57BL_10J", "C57BL_6NJ", "C57BR_cdJ", "C57L_J",
                    "AKR_J", "A_J", "BALB_cJ", "BTBRT_Itpr3_J", "BUB_BnJ", 
                    "C3H_HeH", "C3H_HeJ", "C58_J", "CAST_EiJ", "CBA_J", 
                    "DBA_1J", "DBA_2J", "FVB_NJ", "I_LnJ", "KK_HlJ", 
                    "LEWES_EiJ", "LP_J", "MOLF_EiJ", 
                    "NOD_ShiLtJ", "NZB_BlNJ", "NZO_HlLtJ", "NZW_LacJ", 
                    "PWK_PhJ", "RF_J", "SEA_GnJ", "SPRET_EiJ", "ST_bJ", "WSB_EiJ", 
                    "ZALENDE_EiJ", "X129P2_OlaHsd", "X129S1_SvImJ", "X129S5SvEv_Brd")

colnames_order[!colnames_order %in% colnames(MPG_data_refAlt_sorted)]

heatmap_data <- MPG_data_refAlt_sorted[, colnames_order]

custom_colors <- c("#4d4d4d", "#b2182b")

p_heatmapMPG <- pheatmap(heatmap_data, show_rownames = FALSE, cluster_rows = FALSE, cluster_cols = FALSE, color = custom_colors)
p_heatmapMPG
```


# Calculationg mutations of each cell and cells of each mutation 
```{r calculateMutPerCell}
data_MutPerCell_hf1 <- data.frame(
  CellTag = rownames(MEF_CSM_hf1),
  MutNum_somatic = rowSums(MEF_CSM_hf1 == "Mut"),
  WTNum_somatic = rowSums(MEF_CSM_hf1 == "WT")
)

data_CellPerMut_hf1 <- data.frame(
  Location = colnames(MEF_CSM_hf1),
  CellMutNum = colSums(MEF_CSM_hf1 == "Mut"),
  CellWTNum = colSums(MEF_CSM_hf1 == "WT")
)

data_MutPerCell_hf2 <- data.frame(
  CellTag = rownames(MEF_CSM_hf2),
  MutNum_somatic = rowSums(MEF_CSM_hf2 == "Mut"),
  WTNum_somatic = rowSums(MEF_CSM_hf2 == "WT")
)

data_CellPerMut_hf2 <- data.frame(
  Location = colnames(MEF_CSM_hf2),
  CellMutNum = colSums(MEF_CSM_hf2 == "Mut"),
  CellWTNum = colSums(MEF_CSM_hf2 == "WT")
)
```

# calMulti
```{r calMulti}
# [Cell * Time] = [Cell * Mut] * [Mut * Time]
time_types <- c("Day_6", "Day_9", "Day_12", "Day_15", "Day_21", "Day_28")
cell_num = 1
mut_num = 1

# _hf1
cells_candidate_hf1 <- data_MutPerCell_hf1[data_MutPerCell_hf1$MutNum_somatic >= cell_num,]$CellTag
cells_candidate_hf1 <- cells_candidate_hf1[cells_candidate_hf1 %in% MEF_Meta_hf1$Cellbarcode]
muts_candidate_hf1  <- data_CellPerMut_hf1[data_CellPerMut_hf1$CellMutNum >= mut_num,]$Location

# data_Cell_Mut
data_cell_mut_raw_hf1 <- MEF_CSM_hf1[cells_candidate_hf1, muts_candidate_hf1]

# data_Mut_Time
data_mut_time_hf1 <- as.data.frame(matrix(nrow = length(muts_candidate_hf1), ncol = length(time_types)))
rownames(data_mut_time_hf1) <- muts_candidate_hf1
colnames(data_mut_time_hf1) <- time_types

for (i in 1:nrow(data_mut_time_hf1)){
  mut_tmp <- rownames(data_mut_time_hf1)[i]
  cell_tmp <- retriveCellsByLocation(data_cell_mut_raw_hf1, mut_tmp)
  selected_meta_tags <- MEF_Meta_hf1[MEF_Meta_hf1$Cellbarcode %in% cell_tmp$Mut,]$Day
  data_mut_time_hf1[i,]$Day_6 <- sum(str_count(selected_meta_tags, "Day_6"))
  data_mut_time_hf1[i,]$Day_9 <- sum(str_count(selected_meta_tags, "Day_9"))
  data_mut_time_hf1[i,]$Day_12 <- sum(str_count(selected_meta_tags, "Day_12"))
  data_mut_time_hf1[i,]$Day_15 <- sum(str_count(selected_meta_tags, "Day_15"))
  data_mut_time_hf1[i,]$Day_21 <- sum(str_count(selected_meta_tags, "Day_21"))
  data_mut_time_hf1[i,]$Day_28 <- sum(str_count(selected_meta_tags, "Day_28"))
}

# _hf2
cells_candidate_hf2 <- data_MutPerCell_hf2[data_MutPerCell_hf2$MutNum_somatic >= cell_num,]$CellTag
cells_candidate_hf2 <- cells_candidate_hf2[cells_candidate_hf2 %in% MEF_Meta_hf2$Cellbarcode]
muts_candidate_hf2  <- data_CellPerMut_hf2[data_CellPerMut_hf2$CellMutNum >= mut_num,]$Location

# data_Cell_Mut
data_cell_mut_raw_hf2 <- MEF_CSM_hf2[cells_candidate_hf2, muts_candidate_hf2]

# data_Mut_Time
data_mut_time_hf2 <- as.data.frame(matrix(nrow = length(muts_candidate_hf2), ncol = length(time_types)))
rownames(data_mut_time_hf2) <- muts_candidate_hf2
colnames(data_mut_time_hf2) <- time_types

for (i in 1:nrow(data_mut_time_hf2)){
  mut_tmp <- rownames(data_mut_time_hf2)[i]
  cell_tmp <- retriveCellsByLocation(data_cell_mut_raw_hf2, mut_tmp)
  selected_meta_tags <- MEF_Meta_hf2[MEF_Meta_hf2$Cellbarcode %in% cell_tmp$Mut,]$Day
  data_mut_time_hf2[i,]$Day_6 <- sum(str_count(selected_meta_tags, "Day_6"))
  data_mut_time_hf2[i,]$Day_9 <- sum(str_count(selected_meta_tags, "Day_9"))
  data_mut_time_hf2[i,]$Day_12 <- sum(str_count(selected_meta_tags, "Day_12"))
  data_mut_time_hf2[i,]$Day_15 <- sum(str_count(selected_meta_tags, "Day_15"))
  data_mut_time_hf2[i,]$Day_21 <- sum(str_count(selected_meta_tags, "Day_21"))
  data_mut_time_hf2[i,]$Day_28 <- sum(str_count(selected_meta_tags, "Day_28"))
}
```

```{r computescMutScore}
res_index_hf1 <- cal_scMutScore(data_cell_mut_raw_hf1, data_mut_time_hf1)
res_index_hf2 <- cal_scMutScore(data_cell_mut_raw_hf2, data_mut_time_hf2)
res_index <- rbind(res_index_hf1$score, res_index_hf2$score)

res_index$cb <- rownames(res_index)
res_index <- separate(res_index, col = "cb", into = c("celltype", "cell"), sep = "_")
res_index$Tag <- res_index$celltype
res_index <- separate(res_index, col = "celltype", into = c("rep", "Day"), sep = "\\.")

res_index$Tag <- factor(res_index$Tag, levels = c("hf1.d0", "hf1.d3", "hf1.d6", "hf1.d9", "hf1.d12", "hf1.d15", "hf1.d21", "hf1.d28",
                                                  "hf2.d0", "hf2.d3", "hf2.d6", "hf2.d9", "hf2.d12", "hf2.d15", "hf2.d21", "hf2.d28"))

res_index$Day <- factor(res_index$Day, levels = c("d0", "d3", "d6", "d9", "d12", "d15", "d21", "d28"))

res_index <- res_index[res_index$Diversity > 0, ]
```

```{r checkIndex}
color_settings <- c(d0 = "#FFDC91FF", d3 = "#7876B1FF", d6 = "#4575b4",
                    d9 = "#91bfdb", d12 = "#e0f3f8", d15 = "#fee090",
                    d21 = "#fc8d59", d28 = "#BC3C29FF")

color_settings_2 <- c(Day_0 = "#FFDC91FF", Day_3 = "#7876B1FF", Day_6 = "#4575b4",
                    Day_9 = "#91bfdb", Day_12 = "#e0f3f8", Day_15 = "#fee090",
                    Day_21 = "#fc8d59", Day_28 = "#BC3C29FF")

res_index <- res_index[!is.na(res_index$scMutScore),]

res_index <- res_index[res_index$Day %in% c("d6", "d9", "d12", "d15", "d21", "d28"), ]

ggplot(data = res_index, mapping = aes(x = Day, y = Gini, fill = Day)) +
  geom_violin(trim = TRUE) + 
  geom_boxplot(width = 0.1, fill = "white", outliers = FALSE) +
  labs(x = "", y = "Gini") +
  scale_fill_manual(values = color_settings) +
  theme_classic()

ggplot(data = res_index, mapping = aes(x = Day, y = Tau, fill = Day)) +
  geom_violin(trim = TRUE) + 
  geom_boxplot(width = 0.1, fill = "white", outliers = FALSE) +
  labs(x = "", y = "Tau") +
  scale_fill_manual(values = color_settings) +
  theme_classic()

ggplot(data = res_index, mapping = aes(x = Day, y = GiniTau, fill = Day)) +
  geom_violin(trim = TRUE) + 
  geom_boxplot(width = 0.1, fill = "white", outliers = FALSE) +
  labs(x = "", y = "GiniTau") +
  scale_fill_manual(values = color_settings) +
  theme_classic()

ggplot(data = res_index, mapping = aes(x = Day, y = scMutScore, fill = Day)) +
  geom_violin(trim = TRUE) + 
  geom_boxplot(width = 0.1, fill = "white", outliers = FALSE) +
  labs(x = "", y = "scMutraceScore") +
  scale_fill_manual(values = color_settings) +
  theme_classic()

ggplot(data = res_index, mapping = aes(x = Day, y = scMutScore, fill = Day)) +
  #geom_violin(trim = TRUE) + 
  geom_boxplot(outliers = FALSE) +
  labs(x = "", y = "scMutraceScore") +
  scale_fill_manual(values = color_settings) +
  theme_classic()

res_index_long <- melt(res_index[,c("Gini","Tau", "scMutScore")])

ggplot(res_index_long, aes(x=value, fill=variable)) +
  geom_density(alpha=0.4) +
  theme_classic() +
  labs(x="Value", y="Density", title="Diversity Indices Distribution") +
  scale_fill_brewer(palette="Set3")

corrplot(cor(res_index[,c("Gini","Tau", "Shannon", "Simpson", "Diversity", "scMutScore", "sumMut")], use="pairwise.complete.obs"), 
         method="circle", col = colorRampPalette(c("black", "white", "red"))(200), 
         addCoef.col = "white", tl.col = "black", number.cex = 0.8)
```

# scMutScore Model Validation
```{r scMutScoreModelValidation}
set.seed(2025)

# re2 cells number lower than 100
# d12  d15  d21  d28   d3   d6   d9 
# 166  190  954 1215    2   18   82

# 1. init
data_check_CM_hf1 <- data_cell_mut_raw_hf1
data_check_Meta_hf1 <- MEF_Meta_hf1[MEF_Meta_hf1$Cellbarcode %in% rownames(data_check_CM_hf1),]
col_check_names_hf1 <- c("hf1.d0", "hf1.d3", "hf1.d6", "hf1.d9", "hf1.d12", "hf1.d15", "hf1.d21", "hf1.d28")
#time_types_hf1 <- c("Day_6", "Day_9", "Day_12", "Day_15", "Day_21", "Day_28")
time_types_hf1 <- c("d6", "d9", "d12", "d15", "d21", "d28")

col_check_names <- c(col_check_names_hf1)
cell_num <- 1
mut_num <- 1
mut_loss_ratio <- 0.5
cell_loss_ratio <- 0.5
iteration_times <- 100

# list for save data
data_scMutModel_list_hf1 <- vector("list", iteration_times)

num_cells_hf1 <- nrow(data_check_CM_hf1)
num_mutations_hf1 <- ncol(data_check_CM_hf1)
cells_all_hf1 <- rownames(data_check_CM_hf1)
muts_all_hf1 <- colnames(data_check_CM_hf1)

rowname_day_hf1 <- data.frame(rowname = rownames(data_check_CM_hf1), day = sub(".*\\.(d\\d+)_.*", "\\1", rownames(data_check_CM_hf1)), stringsAsFactors = FALSE)
num_randomSelected_forEachTime_hf1 <- ceiling(min(table(rowname_day_hf1$day)) * cell_loss_ratio)

# 2. loop
## ---------------------------------------------------------------
## Precompute balanced sampling plan: every cell will be sampled
## before any cell is repeatedly sampled within each day.
## ---------------------------------------------------------------
cells_by_day <- split(seq_len(nrow(rowname_day_hf1)), rowname_day_hf1$day)

make_sampling_plan <- function(cells_by_day, n_per_day) {
  lapply(cells_by_day, function(idx) {
    idx_shuffled <- sample(idx)
    split(
      idx_shuffled,
      ceiling(seq_along(idx_shuffled) / n_per_day)
    )
  })
}

sampling_plan <- make_sampling_plan(
  cells_by_day,
  num_randomSelected_forEachTime_hf1
)

max_batches <- max(sapply(sampling_plan, length))
cell_day_map_hf1 <- setNames(rowname_day_hf1$day, rowname_day_hf1$rowname)

for (iter in seq_len(iteration_times)) {

  # 2.1 randomly loss info
  # cell idx
  ## Reshuffle after each full cycle
  
  if (iter > 1 && (iter - 1) %% max_batches == 0) {
    sampling_plan <- make_sampling_plan(
      cells_by_day,
      num_randomSelected_forEachTime_hf1
    )
    max_batches <- max(sapply(sampling_plan, length))
  }

  ## Balanced cell sampling across days
  loss_cell_indices <- unlist(
    lapply(sampling_plan, function(day_batches) {
      batch_id <- ((iter - 1) %% length(day_batches)) + 1
      day_batches[[batch_id]]
    }),
    use.names = FALSE
  )

  loss_mut_indices <- sample(seq_len(num_mutations_hf1), ceiling(mut_loss_ratio * num_mutations_hf1))

  C_partial_missing <- data_check_CM_hf1
  # new data
  if(length(loss_cell_indices) > 0 && length(loss_mut_indices) > 0){
    C_partial_missing[loss_cell_indices, loss_mut_indices] <- "NoCall"
  }

  # 2.2 count cell/mut number
  MutPerCell <- rowSums(C_partial_missing == "Mut")
  CellPerMut <- colSums(C_partial_missing == "Mut")

  cells_candidate <- cells_all_hf1[MutPerCell >= cell_num & cells_all_hf1 %in% data_check_Meta_hf1$Cellbarcode]
  muts_candidate <- muts_all_hf1[CellPerMut >= mut_num]

  message(paste0("Iteration: ", iter, " loss_cell_indices: ", length(loss_cell_indices), " loss_mut_indices: ", length(loss_mut_indices)))
  message(paste0("Iteration: ", iter, " cells_candidate: ", length(cells_candidate), " muts_candidate: ", length(muts_candidate)))

  data_cell_mut_raw_check <- C_partial_missing[cells_candidate, muts_candidate, drop = TRUE]

  # 2.3 init Mut × Time
  M_partial_missing <- matrix(0, nrow = length(muts_candidate), ncol = length(time_types_hf1),
                              dimnames = list(muts_candidate, time_types_hf1))

  for (mut in muts_candidate) {
    cells_with_mut <- rownames(data_cell_mut_raw_check)[data_cell_mut_raw_check[, mut] == "Mut"]
    cells_with_mut_clean <- sub("\\.[0-9]+$", "", cells_with_mut)
    days <- cell_day_map_hf1[cells_with_mut_clean]
    M_partial_missing[mut, ] <- table(factor(days, levels = time_types_hf1))
  }
  
  M_partial_missing_norm <- M_partial_missing

  # 2.5 scMutScore
  res_index_tmp <- cal_scMutScore(data_cell_mut_raw_check, M_partial_missing_norm)$score
  res_index_tmp$iteration <- paste0("iter", iter)
  res_index_tmp <- res_index_tmp[res_index_tmp$Diversity > 0, ]
  res_index_tmp$cb <- rownames(res_index_tmp)
  res_index_tmp$iteration <- paste0("iter", iter)
  rownames(res_index_tmp) <- NULL
  data_scMutModel_list_hf1[[iter]] <- res_index_tmp
}


# 3. merge res
data_scMutModel_res <- do.call(rbind, data_scMutModel_list_hf1)

# 4. column
data_scMutModel_res <- separate(data_scMutModel_res, col = "cb", into = c("celltype", "cell"), sep = "_")
data_scMutModel_res$Tag <- data_scMutModel_res$celltype
data_scMutModel_res <- separate(data_scMutModel_res, col = "celltype", into = c("rep", "Day"), sep = "\\.")

data_scMutModel_res <- data_scMutModel_res[data_scMutModel_res$Day %in% c("d6", "d9", "d12", "d15", "d21", "d28"),]

data_scMutModel_res$Tag <- factor(data_scMutModel_res$Tag, levels = col_check_names)
data_scMutModel_res$Day <- factor(data_scMutModel_res$Day, levels = c("d6", "d9", "d12", "d15", "d21", "d28"))

# 5. plot
data_scMutModel_res_summary <- data_scMutModel_res %>%
  dplyr::group_by(iteration, Day) %>%
  dplyr::summarise(mean_scMutScore = mean(scMutScore, na.rm = TRUE)) %>%
  dplyr::ungroup()

data_scMutModel_res_summary <- data_scMutModel_res_summary %>%
  dplyr::mutate(
    mean_scMutScore_scaled = scales::rescale(mean_scMutScore, to = c(0, 1))
  )

p_iter_scMutraceScore <- ggplot(data = data_scMutModel_res_summary, mapping = aes(x = Day, y = mean_scMutScore_scaled, fill = Day)) +
  geom_boxplot(outliers = FALSE) +
  labs(x = "", y = "scMutraceScore") +
  scale_fill_manual(values = color_settings) +
  theme_classic()

p_iter_scMutraceScore
```



# compareClonesScore
```{r compareCellClonesScore}
# generate data for comparision
MEF_meta_hf1_hf2 <- rbind(MEF_Meta_hf1, MEF_Meta_hf2)
res_index_scMut <- data.frame(Cellbarcode = rownames(res_index), scMutScore = res_index$scMutScore, Tag = res_index$Tag)

res_index_scMut_meta <- merge(res_index_scMut, MEF_meta_hf1_hf2, by = "Cellbarcode")
res_index_scMut_meta$Day <- factor(res_index_scMut_meta$Day, levels = c("Day_0", "Day_3", "Day_6", "Day_9", "Day_12", "Day_15", "Day_21", "Day_28"))
res_index_scMut_meta <- res_index_scMut_meta[res_index_scMut_meta$Day %in% c("Day_6", "Day_9", "Day_12", "Day_15", "Day_21", "Day_28"),]

colNames_data_clone_score <- c("D0_tag", "D3_tag", "D13_tag", "Day", "Mean_Day_score", "Mean_Day_pseudotime", "ID")
data_clone_score <- as.data.frame(matrix(ncol = length(colNames_data_clone_score)))
colnames(data_clone_score) <- colNames_data_clone_score
d0_tags <- unique(MEF_meta_hf1_hf2$CellTagD0_48k)

count_line = 0

for (i in 1:length(d0_tags)){
  d0_tag_tmp <- d0_tags[i]
  d3_tags <- unique(MEF_meta_hf1_hf2[MEF_meta_hf1_hf2$CellTagD0_48k == d0_tag_tmp,]$CellTagD3_48k)
  days_reprogramming <- unique(MEF_meta_hf1_hf2[MEF_meta_hf1_hf2$CellTagD0_48k == d0_tag_tmp,]$Day)
  for (day in 1:length(days_reprogramming)){
    if (length(d3_tags)  >= 1){
    for (j in 1:length(d3_tags)){
      d3_tag_tmp <- d3_tags[j]
      d13_tags <- unique(MEF_meta_hf1_hf2[MEF_meta_hf1_hf2$CellTagD0_48k == d0_tag_tmp & MEF_meta_hf1_hf2$CellTagD3_48k == d3_tag_tmp,]$CellTagD13_48k)
      if(length(d13_tags) >= 1){
          for (k in 1:length(d13_tags)){
            d13_tag_tmp <- d13_tags[k]
            count_line = count_line + 1
            data_clone_score[count_line,]$Day = days_reprogramming[day]
            data_clone_score[count_line,]$D0_tag = d0_tag_tmp
            data_clone_score[count_line,]$D3_tag = d3_tag_tmp
            data_clone_score[count_line,]$D13_tag = d13_tag_tmp
        }
      }else{
        count_line = count_line + 1
        data_clone_score[count_line,]$Day = days_reprogramming[day]
        data_clone_score[count_line,]$D0_tag = d0_tag_tmp
        data_clone_score[count_line,]$D3_tag = d3_tag_tmp
        data_clone_score[count_line,]$D13_tag = "NotAva"
      }
    }
  }else{
    count_line = count_line + 1
    data_clone_score[count_line,]$Day = days_reprogramming[day]
    data_clone_score[count_line,]$D0_tag = d0_tag_tmp
    data_clone_score[count_line,]$D3_tag = "NotAva"
    data_clone_score[count_line,]$D13_tag = "NotAva"
  }
  }
}


data_clone_score$ID <- paste(data_clone_score$D0_tag, data_clone_score$D3_tag, data_clone_score$D13_tag, sep = "_")
data_clone_score <- data_clone_score[data_clone_score$ID != "NotAva_NotAva_NotAva",]

for (i in 1:nrow(data_clone_score)){
  
  d0_tag_tmp <- data_clone_score[i,]$D0_tag
  d3_tag_tmp <- data_clone_score[i,]$D3_tag
  d13_tag_tmp <- data_clone_score[i,]$D13_tag
  day_tmp <- data_clone_score[i,]$Day
  
  data_clone_score[i,]$Mean_Day_score <- "NotAva"
  data_clone_score[i,]$Mean_Day_pseudotime <- "NotAva"
  
  id_check <- paste(d0_tag_tmp, d3_tag_tmp, d13_tag_tmp, sep = "_")
  data_clone_score[i,]$ID <- id_check
  
  scMutscore_tmp <- res_index_scMut_meta[res_index_scMut_meta$cloneTimeTag == id_check & res_index_scMut_meta$Day == day_tmp,]$scMutScore
  pseudotime_tmp <- res_index_scMut_meta[res_index_scMut_meta$cloneTimeTag == id_check & res_index_scMut_meta$Day == day_tmp,]$pseudotime
  
  scMutscore_tmp <- as.numeric(scMutscore_tmp[scMutscore_tmp != "NotAva"])
  pseudotime_tmp <- as.numeric(pseudotime_tmp[pseudotime_tmp!= "NotAva"])
  
  data_clone_score[i,]$Mean_Day_score <- mean(scMutscore_tmp)
  data_clone_score[i,]$Mean_Day_pseudotime <- mean(pseudotime_tmp)
}

# filter
# clonetagDay
data_clone_score_clonetagDay <- data_clone_score[data_clone_score$Mean_Day_score != "NaN" & data_clone_score$Mean_Day_pseudotime != "NaN", ]
ID_counts <- table(data_clone_score_clonetagDay$ID)
valid_IDs <- names(ID_counts[ID_counts >= 2])
data_clone_score_clonetagDay <- subset(data_clone_score_clonetagDay, ID %in% valid_IDs)

data_clone_score_clonetagDay$Mean_Day_score <- as.numeric(data_clone_score_clonetagDay$Mean_Day_score)
data_clone_score_clonetagDay$Mean_Day_pseudotime <- as.numeric(data_clone_score_clonetagDay$Mean_Day_pseudotime)

data_clone_score_clonetagDay$Day <- factor(data_clone_score_clonetagDay$Day, levels = c("Day_0", "Day_3", "Day_6", "Day_9", "Day_12", "Day_15", "Day_21", "Day_28"))


# reprogramming Day
p_clone_scMutraceScore <- ggplot(data_clone_score_clonetagDay, mapping = aes(x = as.factor(Day), y = Mean_Day_score, fill = Day)) +
  geom_boxplot(outliers = TRUE) +
  geom_line(aes(group = ID), color = "gray", alpha = 0.1) +
  scale_fill_manual(values = color_settings_2) +
  ylim(0,1) +
  labs(x = "", y = "scMutraceScore") +
  theme_classic()

p_clone_pseudotime <- ggplot(data_clone_score_clonetagDay, aes(x = as.factor(Day), y = Mean_Day_pseudotime, fill = Day)) +
  geom_boxplot(outliers = TRUE) +
  geom_line(aes(group = ID), color = "gray", alpha = 0.1) +
  scale_fill_manual(values = color_settings_2) +
  labs(x = "", y = "Pseudotime") +
  theme_classic()

p_cell_scMutraceScore <- ggplot(data = res_index_scMut_meta, mapping = aes(x = Day, y = scMutScore, fill = Day)) +
  geom_boxplot(outliers = FALSE) +
  ylim(0,1) +
  scale_fill_manual(values = color_settings_2) +
  labs(x = "", y = "scMutraceScore") +
  theme_classic()

p_cell_pseudotime <- ggplot(data = res_index_scMut_meta[res_index_scMut_meta$pseudotime != "NotAva",], mapping = aes(x = Day, y = as.numeric(pseudotime), fill = Day)) +
  geom_boxplot(outliers = FALSE) +
  scale_fill_manual(values = color_settings_2) +
  labs(x = "", y = "pseudotime") +
  theme_classic()

res_index_scMut_meta_success <- res_index_scMut_meta[res_index_scMut_meta$Outcome == "Succesful",][,c("scMutScore", "pseudotime", "Day", "Outcome")]

res_index_scMut_meta_deadend <- res_index_scMut_meta[res_index_scMut_meta$Outcome == "DeadEnd",][,c("scMutScore", "pseudotime", "Day", "Outcome")]

res_index_scMut_meta_success_deadend <- rbind(res_index_scMut_meta_success, res_index_scMut_meta_deadend)

# Outcome

p_reprogramming_scMutraceScore <- ggplot(res_index_scMut_meta[res_index_scMut_meta$reprogramming %in%c("F_reprogramming", "T_reprogramming"),], aes(x = as.factor(reprogramming), y = scMutScore, fill = reprogramming)) +
  labs(x = "", y = "scMutScore") +
  geom_boxplot(outliers = FALSE) +
  stat_compare_means(method = "wilcox.test", method.args = list(alternative = "two.sided")) +
  scale_fill_manual(values = c("#7876B1FF", "#FFDC91FF")) +
  theme_classic()

p_reprogramming_pseudotime <- ggplot(res_index_scMut_meta[res_index_scMut_meta$reprogramming %in%c("F_reprogramming", "T_reprogramming"),], aes(x = as.factor(reprogramming), y = as.numeric(pseudotime), fill = reprogramming)) +
  labs(x = "", y = "pseudotime") +
  geom_boxplot(outliers = FALSE) +
  stat_compare_means(method = "wilcox.test", method.args = list(alternative = "two.sided")) +
  scale_fill_manual(values = c("#7876B1FF", "#FFDC91FF")) +
  theme_classic()

timepoint_color <- c(Day_6 = '#7D5188', Day_9 = '#7D71A5',  Day_12 = '#738DAF',  Day_15 = '#6EA5B0', Day_21 = '#70BDAA',  Day_28 = '#FDEF77')
cluster_color <- c(c0 = '#D5393E', c1 = '#5A6B93',  c2 = '#519376',  c3 = '#6E8371', c4 = '#A55F7F',  c5 = '#E48648', c6 = "#F3E95B", c7 = "#BE9644", c8 = "#B26966", c9 = "#080908")
outcom_color <- c(BAB = "lightgray", DeadEnd = "#7876B1FF", Succesful = "#FFDC91FF")
reprogramming_colors <- c(F_reprogramming = '#7876B1FF', T_reprogramming = '#FFDC91FF', NotAva = "#d9d9d9")

p_umap_reprogramming <- ggplot(data = res_index_scMut_meta[res_index_scMut_meta$reprogramming %in% c("F_reprogramming", "T_reprogramming"), ]) + geom_point(mapping =  aes(x = TSNE_1, y = TSNE_2, color = reprogramming), size = 2, show.legend = TRUE) +
  scale_color_manual(values = reprogramming_colors) +
  xlab("") + ylab("") +
  theme_classic()

p_umap_outcome <- ggplot(data = res_index_scMut_meta[res_index_scMut_meta$Outcome %in% c("DeadEnd", "Succesful"), ]) + geom_point(mapping =  aes(x = TSNE_1, y = TSNE_2, color = Outcome), size = 2, show.legend = TRUE) +
  scale_color_manual(values = outcom_color) +
  xlab("") + ylab("") +
  theme_classic()

p_umap_cell_day <- ggplot(data = res_index_scMut_meta) + geom_point(mapping =  aes(x = TSNE_1, y = TSNE_2, color = Day), size = 1.5, alpha = 0.7, show.legend = TRUE) +
  scale_color_manual(values = color_settings_2) +
  xlab("") + ylab("") +
  theme_classic()

res_index_scMut_meta$cluster <- paste0("c", res_index_scMut_meta$cluster)
p_umap_cell_cluster <- ggplot(data = res_index_scMut_meta) + geom_point(mapping =  aes(x = TSNE_1, y = TSNE_2, color = cluster), size = 1.5, show.legend = TRUE) +
  scale_color_manual(values = cluster_color) +
  xlab("") + ylab("") +
  theme_classic()

p_umap_cell_Pseudotime <- ggplot(res_index_scMut_meta[res_index_scMut_meta$pseudotime != "NotAva",], aes(x = TSNE_1, y = TSNE_2, fill = as.numeric(pseudotime))) + 
  geom_point(shape = 21, size = 2, alpha = 0.3) +
  scale_fill_gradientn(colours = c("#313695", "#4575b4", "#74add1", "#abd9e9", "#e0f3f8", "#ffffbf", "#fee090", "#fdae61", "#f46d43", "#d73027", "#a50026")) +
  theme_classic() +
  labs(fill = "Pseudotime")

p_umap_cell_scMutScore <- ggplot(res_index_scMut_meta[res_index_scMut_meta$scMutScore != "NotAva",], aes(x = TSNE_1, y = TSNE_2, fill =  as.numeric(scMutScore))) + 
  geom_point(shape = 21, size = 2, alpha = 0.3) +
  scale_fill_gradientn(colours = c("#313695", "#4575b4", "#74add1", "#abd9e9", "#e0f3f8", "#ffffbf", "#fee090", "#fdae61", "#f46d43", "#d73027", "#a50026")) + 
  theme_classic() +
  labs(fill = "scMutScore")

data_scMutScore_pseudotime <- res_index_scMut_meta[,c("scMutScore", "pseudotime")]
data_scMutScore_pseudotime <- data_scMutScore_pseudotime[data_scMutScore_pseudotime$scMutScore != "NotAva" & data_scMutScore_pseudotime$pseudotime != "NotAva",]
data_scMutScore_pseudotime$scMutScore <- as.numeric(data_scMutScore_pseudotime$scMutScore)
data_scMutScore_pseudotime$pseudotime <- as.numeric(data_scMutScore_pseudotime$pseudotime)
data_scMutScore_pseudotime$pseudotime_scale <- (data_scMutScore_pseudotime$pseudotime - min(data_scMutScore_pseudotime$pseudotime)) / (max(data_scMutScore_pseudotime$pseudotime) - min(data_scMutScore_pseudotime$pseudotime))

p_cor_scMutScore_pseudotime <- ggplot(data_scMutScore_pseudotime, aes(x = scMutScore, y = pseudotime_scale)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") + 
  stat_cor(method = "pearson", label.x = 0.5, label.y = 1) +
  labs(x = "scMutScore", y = "Pseudotime") + 
  theme_classic()

p_cor_scMutScore_sumMut <- ggplot(res_index, aes(x = scMutScore, y = log2(sumMut))) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") + 
  stat_cor(method = "pearson", label.x = 0.5, label.y = 1) +
  labs(x = "scMutScore", y = "sum of matrix") + 
  theme_classic()

cor.test(data_scMutScore_pseudotime$scMutScore, data_scMutScore_pseudotime$pseudotime_scale)
cor.test(res_index$scMutScore, res_index$sumMut)
```

# compareClonesScore figures
```{r figures}
p_cell_scMutraceScore
p_cell_pseudotime

p_clone_scMutraceScore
p_clone_pseudotime

p_reprogramming_scMutraceScore
p_reprogramming_pseudotime

p_umap_reprogramming
p_umap_outcome

p_umap_cell_day
p_umap_cell_cluster

p_umap_cell_Pseudotime
p_umap_cell_scMutScore

p_cor_scMutScore_pseudotime
p_cor_scMutScore_sumMut
```

# save figures
```{r savefigures, eval=FALSE} 
# save data
pdf("../Figure/MEF/p_cell_scMutraceScore.pdf", width = 5.36, height = 3.42)
p_cell_scMutraceScore
dev.off()

pdf("../Figure/MEF/p_cell_pseudotime.pdf", width = 5.36, height = 3.42)
p_cell_pseudotime
dev.off()

pdf("../Figure/MEF/p_clone_scMutraceScore.pdf", width = 5.36, height = 3.42)
p_clone_scMutraceScore
dev.off()

pdf("../Figure/MEF/p_clone_pseudotime.pdf", width = 5.36, height = 3.42)
p_clone_pseudotime
dev.off()

pdf("../Figure/MEF/p_umap_cell_day.pdf", width = 5.36, height = 3.42)
p_umap_cell_day
dev.off()

pdf("../Figure/MEF/p_umap_cell_scMutScore.pdf", width = 5.36, height = 3.42)
p_umap_cell_scMutScore
dev.off()

pdf("../Figure/MEF/p_umap_cell_Pseudotime.pdf", width = 5.36, height = 3.42)
p_umap_cell_Pseudotime
dev.off()

pdf("../Figure/MEF/p_success_faile_scMutraceScore.pdf", width = 5.36, height = 3.42)
p_reprogramming_scMutraceScore
dev.off()
```


# check randomly clone scMutScore
```{r checkrandomlyclonescMutScore}
selected_Days <- c("Day_6", "Day_9", "Day_12", "Day_15", "Day_21", "Day_28")
set.seed(2025)
data_random_cells <- res_index_scMut_meta[sample(nrow(res_index_scMut_meta), 1000, replace = TRUE), c("scMutScore", "Day")]
data_random_cells$Day <- "Day_6"

for (i in 2:length(selected_Days)){
  random_rows_1 <- res_index_scMut_meta[sample(nrow(res_index_scMut_meta), 50), c("scMutScore", "Day")]
  random_rows_1$Day <- selected_Days[i]
  data_random_cells <- rbind(data_random_cells, random_rows_1)
}

data_random_cells$Day <- factor(data_random_cells$Day, levels = c("Day_6", "Day_9", "Day_12", "Day_15", "Day_21", "Day_28"))

p_random_cell_days <- ggplot(data = data_random_cells, mapping = aes(x = Day, y = scMutScore, fill = Day)) +
  geom_boxplot(outliers = FALSE) +
  scale_fill_manual(values = color_settings_2) +
  ylim(0,1) +
  labs(x = "", y = "scMutraceScore") +
  theme_classic()

data_random_clones <- data_clone_score_clonetagDay[sample(nrow(data_clone_score_clonetagDay), 400, replace = TRUE), c("Mean_Day_score", "Day", "ID")]
data_random_clones$Day <- "Day_6"

for (i in 2:length(selected_Days)){
  random_rows_1 <- data_clone_score_clonetagDay[sample(nrow(data_clone_score_clonetagDay), 50), c("Mean_Day_score", "Day", "ID")]
  random_rows_1$Day <- selected_Days[i]
  data_random_clones <- rbind(data_random_clones, random_rows_1)
}

data_random_clones$Day <- factor(data_random_clones$Day, levels = c("Day_6", "Day_9", "Day_12", "Day_15", "Day_21", "Day_28"))

p_random_clone_days <- ggplot(data_random_clones, mapping = aes(x = as.factor(Day), y = Mean_Day_score, fill = Day)) +
  geom_boxplot(outliers = TRUE) +
  geom_line(aes(group = ID), color = "gray", alpha = 0.1) +
  scale_fill_manual(values = color_settings_2) +
  ylim(0,1) +
  labs(x = "", y = "scMutraceScore") +
  theme_classic()

p_random_cell_days
p_random_clone_days

library(clinfun)
# cells scMutscore
res_index_selected <- res_index_scMut_meta[res_index_scMut_meta$Day %in% c('Day_6', 'Day_9', 'Day_12', 'Day_15', 'Day_21', 'Day_28'),]
res_index_selected$Day <- factor(res_index_selected$Day , levels = c('Day_6', 'Day_9', 'Day_12', 'Day_15', 'Day_21', 'Day_28'), ordered = TRUE)
jonckheere.test(x = res_index_selected$scMutScore, g = res_index_selected$Day, alternative = "increasing")
  # JT = 9075317, p-value < 2.2e-16
  # alternative hypothesis: increasing

res_index_selected <- res_index_scMut_meta[res_index_scMut_meta$Day %in% c('Day_6', 'Day_9', 'Day_12', 'Day_15', 'Day_21', 'Day_28'),]
res_index_selected <- res_index_scMut_meta[res_index_scMut_meta$pseudotime != "NotAva",]
res_index_selected$Day <- factor(res_index_selected$Day , levels = c('Day_6', 'Day_9', 'Day_12', 'Day_15', 'Day_21', 'Day_28'), ordered = TRUE)
jonckheere.test(x = as.numeric(res_index_selected$pseudotime), g = res_index_selected$Day, alternative = "increasing")
  #JT = 6857494, p-value < 2.2e-16
  #alternative hypothesis: increasing

# clone scMutscore
data_clone_score_clonetagDay_selected <- data_clone_score_clonetagDay[data_clone_score_clonetagDay$Day %in% c('Day_6', 'Day_9', 'Day_12', 'Day_15', 'Day_21', 'Day_28'),]
data_clone_score_clonetagDay_selected$Day <- factor(data_clone_score_clonetagDay_selected$Day , levels = c('Day_6', 'Day_9', 'Day_12', 'Day_15', 'Day_21', 'Day_28'), ordered = TRUE)
jonckheere.test(x = data_clone_score_clonetagDay_selected$Mean_Day_score, g = data_clone_score_clonetagDay_selected$Day, alternative = "increasing")
  #JT = 26956, p-value = 0.000107
  #alternative hypothesis: increasing
jonckheere.test(x = as.numeric(data_clone_score_clonetagDay_selected$Mean_Day_pseudotime), g = data_clone_score_clonetagDay_selected$Day, alternative = "increasing")
  #JT = 37857, p-value < 2.2e-16
  #alternative hypothesis: increasing

# data_random_cells
data_random_cells$Day <- factor(data_random_cells$Day , levels = c('Day_6', 'Day_9', 'Day_12', 'Day_15', 'Day_21', 'Day_28'), ordered = TRUE)
jonckheere.test(x = data_random_cells$scMutScore, g = data_random_cells$Day, alternative = "increasing")
  #JT = 135420, p-value = 0.657
  #alternative hypothesis: increasing

#data_random_clones
data_random_clones$Day <- factor(data_random_clones$Day , levels = c('Day_6', 'Day_9', 'Day_12', 'Day_15', 'Day_21', 'Day_28'), ordered = TRUE)
jonckheere.test(x = data_random_clones$Mean_Day_score, g = data_random_clones$Day, alternative = "increasing")
  #JT = 60830, p-value = 0.7551
  #alternative hypothesis: increasing
```

# tracing allClones
```{r tracingallClones}
res_index_scMut_meta_MEF <- res_index_scMut_meta[res_index_scMut_meta$CellTagD0_48k != "NotAva" & res_index_scMut_meta$CellTagD3_48k == "NotAva" & res_index_scMut_meta$CellTagD13_48k == "NotAva",]

res_index_scMut_meta_D3  <- res_index_scMut_meta[res_index_scMut_meta$CellTagD3_48k != "NotAva" & res_index_scMut_meta$CellTagD13_48k == "NotAva",]

res_index_scMut_meta_D13  <- res_index_scMut_meta[res_index_scMut_meta$CellTagD13_48k != "NotAva",]


p_cell_MEF <- ggplot(data = res_index_scMut_meta_MEF, mapping = aes(x = Day, y = scMutScore, fill = Day)) +
  geom_violin(trim = TRUE) + coord_cartesian(ylim = c(0, 1)) +
  geom_boxplot(width = 0.1, fill = "white", outliers = FALSE) +
  scale_fill_manual(values = color_settings_2) +
  labs(x = "", y = "scMutraceScore") +
  theme_classic()

p_cell_D3 <- ggplot(data = res_index_scMut_meta_D3, mapping = aes(x = Day, y = scMutScore, fill = Day)) +
  geom_violin(trim = TRUE) + coord_cartesian(ylim = c(0, 1)) +
  geom_boxplot(width = 0.1, fill = "white", outliers = FALSE) +
  scale_fill_manual(values = color_settings_2) +
  labs(x = "", y = "scMutraceScore") +
  theme_classic()

p_cell_D13 <- ggplot(data = res_index_scMut_meta_D13, mapping = aes(x = Day, y = scMutScore, fill = Day)) +
  geom_violin(trim = TRUE) + coord_cartesian(ylim = c(0, 1)) +
  geom_boxplot(width = 0.1, fill = "white", outliers = FALSE) +
  #stat_summary(fun.data="mean_sdl", fun.args = list(mult=1),geom="pointrange", color = "black") +
  scale_fill_manual(values = color_settings_2) +
  labs(x = "", y = "scMutraceScore") +
  theme_classic()

p_cell_MEF
p_cell_D3
p_cell_D13
```

# randomlySelectMut
```{r randomlySelectMut}
color_settings <- c(d0 = "#FFDC91FF", d3 = "#7876B1FF", d6 = "#4575b4",
                    d9 = "#91bfdb", d12 = "#e0f3f8", d15 = "#fee090",
                    d21 = "#fc8d59", d28 = "#BC3C29FF")

color_settings_2 <- c(Day_0 = "#FFDC91FF", Day_3 = "#7876B1FF", Day_6 = "#4575b4",
                    Day_9 = "#91bfdb", Day_12 = "#e0f3f8", Day_15 = "#fee090",
                    Day_21 = "#fc8d59", Day_28 = "#BC3C29FF")

col_names_days <- c("Day_6", "Day_9", "Day_12", "Day_15", "Day_21", "Day_28", "sampleSize")
reps_random <- 100
row_names_reps <- paste("iter", c(1:reps_random), sep = "_")

sample_num <- c(10, 30, 50, 100, 150)

data_random_muts2cell_hf1 <- as.data.frame(matrix(nrow = reps_random * length(sample_num), ncol = length(col_names_days)))
data_random_muts2cell_hf2 <- as.data.frame(matrix(nrow = reps_random * length(sample_num), ncol = length(col_names_days)))

colnames(data_random_muts2cell_hf1) <- col_names_days
colnames(data_random_muts2cell_hf2) <- col_names_days

#rownames(data_random_muts2cell_hf1) <- row_names_reps
#rownames(data_random_muts2cell_hf2) <- row_names_reps

hf1_muts <- data_CellPerMut_hf1$Location
hf2_muts <- data_CellPerMut_hf2$Location

set.seed(2025)
line_count <- 0
for (i in 1:reps_random){
  
  for (j in sample_num){
    
    line_count <- line_count + 1
    hf1_muts_tmp <- sample(hf1_muts, j, replace = TRUE)
    hf2_muts_tmp <- sample(hf2_muts, j, replace = TRUE)
    
    hf1_cells <- retriveCellsByLocation(MEF_CSM_hf1, hf1_muts_tmp)$Mut
    hf2_cells <- retriveCellsByLocation(MEF_CSM_hf2, hf2_muts_tmp)$Mut
    
    data_random_muts2cell_hf1[line_count, "Day_6"] <- sum(str_count(hf1_cells, "hf1.d6"))
    data_random_muts2cell_hf1[line_count, "Day_9"] <- sum(str_count(hf1_cells, "hf1.d9"))
    data_random_muts2cell_hf1[line_count, "Day_12"] <- sum(str_count(hf1_cells, "hf1.d12"))
    data_random_muts2cell_hf1[line_count, "Day_15"] <- sum(str_count(hf1_cells, "hf1.d15"))
    data_random_muts2cell_hf1[line_count, "Day_21"] <- sum(str_count(hf1_cells, "hf1.d21"))
    data_random_muts2cell_hf1[line_count, "Day_28"] <- sum(str_count(hf1_cells, "hf1.d28"))
    data_random_muts2cell_hf1[line_count, "sampleSize"] <- j
    
    data_random_muts2cell_hf2[line_count, "Day_6"] <- sum(str_count(hf2_cells, "hf2.d6"))
    data_random_muts2cell_hf2[line_count, "Day_9"] <- sum(str_count(hf2_cells, "hf2.d9"))
    data_random_muts2cell_hf2[line_count, "Day_12"] <- sum(str_count(hf2_cells, "hf2.d12"))
    data_random_muts2cell_hf2[line_count, "Day_15"] <- sum(str_count(hf2_cells, "hf2.d15"))
    data_random_muts2cell_hf2[line_count, "Day_21"] <- sum(str_count(hf2_cells, "hf2.d21"))
    data_random_muts2cell_hf2[line_count, "Day_28"] <- sum(str_count(hf2_cells, "hf2.d28"))
    data_random_muts2cell_hf2[line_count, "sampleSize"] <- j
  }
}

data_random_muts2cell_hf1$iteration <- rownames(data_random_muts2cell_hf1)
data_random_muts2cell_hf2$iteration <- rownames(data_random_muts2cell_hf2)

data_random_muts2cell_hf1_long <- data_random_muts2cell_hf1 %>%
  pivot_longer(
    cols = starts_with("Day"),
    names_to = "Day",         
    values_to = "CellNum"
  )

data_random_muts2cell_hf2_long <- data_random_muts2cell_hf2 %>%
  pivot_longer(
    cols = starts_with("Day"),
    names_to = "Day",         
    values_to = "CellNum"
  )

data_random_muts2cell_hf1_long$Day <- factor(data_random_muts2cell_hf1_long$Day, levels = c("Day_6", "Day_9", "Day_12", "Day_15", "Day_21", "Day_28"))
data_random_muts2cell_hf2_long$Day <- factor(data_random_muts2cell_hf2_long$Day, levels = c("Day_6", "Day_9", "Day_12", "Day_15", "Day_21", "Day_28"))


p_Mut2Cell_hf1 <- ggplot(data_random_muts2cell_hf1_long,
       aes(x = Day, y = CellNum, color = factor(sampleSize))) + 
  geom_jitter(width = 0.2, alpha = 0.5, size = 1) +   
  ylab("Number of cells") +
  xlab("Day") +
  labs(color = "SelectNum") +
   scale_color_manual(values = c(
    "10" = "#4d4d4d",
    "30" = "#bababa",
    "50" = "#fddbc7",
    "100" = "#f4a582",
    "150" = "#b2182b"
  )) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_Mut2Cell_hf2 <- ggplot(data_random_muts2cell_hf2_long,
       aes(x = Day, y = CellNum, color = factor(sampleSize))) + 
  geom_jitter(width = 0.2, alpha = 0.5, size = 1) +   
  ylab("Number of cells") +
  xlab("Day") +
  labs(color = "SelectNum") +
   scale_color_manual(values = c(
    "10" = "#4d4d4d",
    "30" = "#bababa",
    "50" = "#fddbc7",
    "100" = "#f4a582",
    "150" = "#b2182b"
  )) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_Mut2Cell_hf1
p_Mut2Cell_hf2
```

# lineageTracingExample
```{r lineageTracingExample}
selected_start_cloneID <- 2352
data_clone_score_selected <- res_index_scMut_meta[res_index_scMut_meta$CellTagD0_48k == selected_start_cloneID,][,c("Cellbarcode", "scMutScore", "CellTagD0_48k", "CellTagD3_48k", "CellTagD13_48k")]
colnames(data_clone_score_selected) <- c("Cellbarcode", "scMutScore", "D0_tag", "D3_tag", "D13_tag")


# Categorization of all situations
data_clone_score_selected_long <- bind_rows(
  # C1: D3 == NotAva & D13 != NotAva → D0 → D13
  data_clone_score_selected %>%
    dplyr::filter(D3_tag == "NotAva" & D13_tag != "NotAva") %>%
    dplyr::select(source = D0_tag, target = D13_tag),
  
  # C2: D3 == NotAva & D13 == NotAva → D0 → D0_self
  data_clone_score_selected %>%
    dplyr::filter(D3_tag == "NotAva" & D13_tag == "NotAva") %>%
    dplyr::mutate(target = paste0(D0_tag, "_self")) %>%
    dplyr::select(source = D0_tag, target),
  
  # C3: D3 != NotAva & D13 != NotAva → D0→D3
  data_clone_score_selected %>%
    dplyr::filter(D3_tag != "NotAva" & D13_tag != "NotAva") %>%
    dplyr::select(source = D0_tag, target = D3_tag),
  
  # C4: D3 != NotAva & D13 != NotAva → D3→D13
  data_clone_score_selected %>%
    dplyr::filter(D3_tag != "NotAva" & D13_tag != "NotAva") %>%
    dplyr::select(source = D3_tag, target = D13_tag),
  
  # C5: D3 != NotAva & D13 == NotAva → D0→D3
  data_clone_score_selected %>%
    dplyr::filter(D3_tag != "NotAva" & D13_tag == "NotAva") %>%
    dplyr::select(source = D0_tag, target = D3_tag),
  
  # C6: D3 != NotAva & D13 == NotAva → D3→D3_self
  data_clone_score_selected %>%
    dplyr::filter(D3_tag != "NotAva" & D13_tag == "NotAva") %>%
    dplyr::mutate(target = paste0(D3_tag, "_self")) %>%
    dplyr::select(source = D3_tag, target)
)

# Counting the number of pairs
clone_links <- data_clone_score_selected_long %>%
  dplyr::group_by(source, target) %>%
  dplyr::summarise(value = n()) %>%
  dplyr::arrange(desc(value))

clone_links <- as.data.frame(clone_links)

clone_nodes <- data.frame(
  name=c(as.character(clone_links$source), 
  as.character(clone_links$target)) %>% unique()
)

clone_nodes$group <- ""

for (i in 1:nrow(clone_nodes)){
  checkNode <- as.character(clone_nodes[i,"name"])
  if (checkNode %in% as.character(data_clone_score_selected$D0_tag)){
    clone_nodes[i,"group"] <- "MEF"
  }
  
  if (checkNode %in% as.character(data_clone_score_selected$D3_tag)){
    clone_nodes[i,"group"] <- "D3"
  }
  
  if (checkNode %in% as.character(data_clone_score_selected$D13_tag)){
    clone_nodes[i,"group"] <- "D13"
  }
}

clone_links$IDsource <- match(clone_links$source, clone_nodes$name)-1 
clone_links$IDtarget <- match(clone_links$target, clone_nodes$name)-1

p_sankeyNetwork <- sankeyNetwork(Links = clone_links, Nodes = clone_nodes,
              Source = "IDsource", Target = "IDtarget",
              Value = "value", NodeID = "name", 
              sinksRight=FALSE,
              colourScale = JS("d3.scaleOrdinal()
                                  .domain(['MEF', 'D3', 'D13'])
                                  .range(['#1f77b4', '#ff7f0e', '#2ca02c'])"),
              NodeGroup = "group")

p_sankeyNetwork
```

# lineageTracingExample with scMutScore
```{r network}
selected_start_cloneID <- c(124, 487, 2352)

data_clone_score_selected <- res_index_scMut_meta[res_index_scMut_meta$CellTagD0_48k %in% selected_start_cloneID,][,c("Cellbarcode", "scMutScore", "CellTagD0_48k", "CellTagD3_48k", "CellTagD13_48k")]
colnames(data_clone_score_selected) <- c("Cellbarcode", "scMutScore", "D0_tag", "D3_tag", "D13_tag")

# Categorization of all situations
data_clone_score_selected_long <- bind_rows(
  # C1: D3 == NotAva & D13 != NotAva → D0 → D13
  data_clone_score_selected %>%
    dplyr::filter(D3_tag == "NotAva" & D13_tag != "NotAva") %>%
    dplyr::select(source = D0_tag, target = D13_tag),
  
  # C2: D3 == NotAva & D13 == NotAva → D0 → D0_self
  data_clone_score_selected %>%
    dplyr::filter(D3_tag == "NotAva" & D13_tag == "NotAva") %>%
    dplyr::mutate(target = paste0(D0_tag, "_self")) %>%
    dplyr::select(source = D0_tag, target),
  
  # C3: D3 != NotAva & D13 != NotAva → D0→D3
  data_clone_score_selected %>%
    dplyr::filter(D3_tag != "NotAva" & D13_tag != "NotAva") %>%
    dplyr::select(source = D0_tag, target = D3_tag),
  
  # C4: D3 != NotAva & D13 != NotAva → D3→D13
  data_clone_score_selected %>%
    dplyr::filter(D3_tag != "NotAva" & D13_tag != "NotAva") %>%
    dplyr::select(source = D3_tag, target = D13_tag),
  
  # C5: D3 != NotAva & D13 == NotAva → D0→D3
  data_clone_score_selected %>%
    dplyr::filter(D3_tag != "NotAva" & D13_tag == "NotAva") %>%
    dplyr::select(source = D0_tag, target = D3_tag),
  
  # C6: D3 != NotAva & D13 == NotAva → D3→D3_self
  data_clone_score_selected %>%
    dplyr::filter(D3_tag != "NotAva" & D13_tag == "NotAva") %>%
    dplyr::mutate(target = paste0(D3_tag, "_self")) %>%
    dplyr::select(source = D3_tag, target)
)

# Counting the number of pairs
clone_links <- data_clone_score_selected_long %>%
  dplyr::group_by(source, target) %>%
  dplyr::summarise(value = n()) %>%
  dplyr::arrange(desc(value))

clone_links <- as.data.frame(clone_links)

clone_links <- clone_links[!grepl("_self", clone_links$target), ]

#clone_links <- clone_links[clone_links$value > 1,]

clone_nodes <- data.frame(
  name=c(as.character(clone_links$source), 
  as.character(clone_links$target)) %>% unique()
)

clone_nodes$group <- ""

for (i in 1:nrow(clone_nodes)){
  checkNode <- as.character(clone_nodes[i,"name"])
  if (checkNode %in% as.character(data_clone_score_selected$D0_tag)){
    clone_nodes[i,"group"] <- "MEF"
  }
  
  if (checkNode %in% as.character(data_clone_score_selected$D3_tag)){
    clone_nodes[i,"group"] <- "D3"
  }
  
  if (checkNode %in% as.character(data_clone_score_selected$D13_tag)){
    clone_nodes[i,"group"] <- "D13"
  }
}

# cal node scMutScore
clone_nodes$scMutScore <- 0

data_MEF_D3_D13 <- data.frame()

for (i in 1:nrow(clone_nodes)){
  group_tmp <- clone_nodes[i, "group"]
  node_tmp <- clone_nodes[i, "name"]
  print(node_tmp)
  print(group_tmp)
  if (group_tmp == "MEF"){
    data_mef_tmp <- res_index_scMut_meta[res_index_scMut_meta$CellTagD0_48k %in% selected_start_cloneID & res_index_scMut_meta$CellTagD3_48k == "NotAva" & res_index_scMut_meta$CellTagD13_48k == "NotAva", c("scMutScore", "Day")]
    if(nrow(data_mef_tmp) == 0){
      data_mef_tmp_tmp <- res_index_scMut_meta[res_index_scMut_meta$CellTagD0_48k == selected_start_cloneID, c("scMutScore", "Day")]
      data_mef_tmp_tmp$Tag <- "MEF"
      clone_nodes[i, "scMutScore"] <- min(data_mef_tmp_tmp$scMutScore)
      data_MEF_D3_D13 <- rbind(data_MEF_D3_D13, data_mef_tmp_tmp)
    }else{
      data_mef_tmp$Tag <- "MEF"
      clone_nodes[i, "scMutScore"] <- min(data_mef_tmp$scMutScore)
      data_MEF_D3_D13 <- rbind(data_MEF_D3_D13, data_mef_tmp)
    }
  }
  if (group_tmp == "D3"){
    data_d3_tmp <- res_index_scMut_meta[res_index_scMut_meta$CellTagD0_48k %in% selected_start_cloneID & res_index_scMut_meta$CellTagD3_48k == node_tmp & res_index_scMut_meta$CellTagD13_48k == "NotAva", c("scMutScore", "Day")]
    if(nrow(data_d3_tmp) == 0){
      data_d3_tmp_tmp <- res_index_scMut_meta[res_index_scMut_meta$CellTagD0_48k == selected_start_cloneID & res_index_scMut_meta$CellTagD3_48k == node_tmp, c("scMutScore", "Day")]
      data_d3_tmp_tmp$Tag <- "D3"
      clone_nodes[i, "scMutScore"] <- min(data_d3_tmp_tmp$scMutScore)
      data_MEF_D3_D13 <- rbind(data_MEF_D3_D13, data_d3_tmp_tmp)
    }else{
      data_d3_tmp$Tag <- "D3"
      clone_nodes[i, "scMutScore"] <- min(data_d3_tmp$scMutScore)
      data_MEF_D3_D13 <- rbind(data_MEF_D3_D13, data_d3_tmp)
    }
  }
  if (group_tmp == "D13"){
    data_d13_tmp <- res_index_scMut_meta[res_index_scMut_meta$CellTagD0_48k %in% selected_start_cloneID & res_index_scMut_meta$CellTagD13_48k == node_tmp, c("scMutScore", "Day")]
    data_d13_tmp$Tag <- "D13"
    clone_nodes[i, "scMutScore"] <- min(data_d13_tmp$scMutScore)
    data_MEF_D3_D13 <- rbind(data_MEF_D3_D13, data_d13_tmp)
  }
}

data_MEF_D3_D13$Tag <- factor(data_MEF_D3_D13$Tag, levels = c("MEF", "D3", "D13"))
clone_nodes$group <- factor(clone_nodes$group, levels = c("MEF", "D3", "D13"))
data_MEF_D3_D13$Day <- factor(data_MEF_D3_D13$Day, levels = c("Day_0", "Day_3", "Day_6", "Day_9", "Day_12", "Day_15", "Day_21", "Day_28"))

network <- graph_from_data_frame(d=clone_links, vertices=clone_nodes, directed=T)

color_settings_network_1 <- c( D13 = "#e41a1c", D3 = "#377eb8", MEF = "#4daf4a")

color_settings_network_2 <- colorRampPalette(c("#fee8c8", "#fdbb84", "#e34a33"))

colors <- color_settings_network_2(5)

breaks <- unique(quantile(V(network)$scMutScore, probs = seq(0, 1, length.out = 6)))

value_group <- cut(V(network)$scMutScore, breaks = breaks, include.lowest = TRUE, labels = FALSE)

my_color_2 <- colors[value_group]

edge_width <- log2(E(network)$value + 0.1)

vertex_frame_color <- color_settings_network_1[V(network)$group]

p_selected_clone_MEF <- ggplot(data = data_MEF_D3_D13[data_MEF_D3_D13$Tag == "MEF",], mapping = aes(x = Day, y = scMutScore, fill = Day)) +
  geom_violin(trim = FALSE) + coord_cartesian(ylim = c(0, 1)) +
  geom_boxplot(width = 0.05, fill = "white", outliers = FALSE) +
  #stat_summary(fun.data="mean_sdl", fun.args = list(mult=1),geom="pointrange", color = "black") +
  scale_fill_manual(values = color_settings_2) +
  labs(x = "", y = paste0("scMutraceScore_", selected_start_cloneID)) +
  theme_classic()

p_selected_clone_D3 <- ggplot(data = data_MEF_D3_D13[data_MEF_D3_D13$Tag == "D3",], mapping = aes(x = Day, y = scMutScore, fill = Day)) +
  geom_violin(trim = FALSE) + coord_cartesian(ylim = c(0, 1)) +
  geom_boxplot(width = 0.05, fill = "white", outliers = FALSE) +
  #stat_summary(fun.data="mean_sdl", fun.args = list(mult=1),geom="pointrange", color = "black") +
  scale_fill_manual(values = color_settings_2) +
  labs(x = "", y = paste0("scMutraceScore_", selected_start_cloneID)) +
  theme_classic()

p_selected_clone_D13 <- ggplot(data = data_MEF_D3_D13[data_MEF_D3_D13$Tag == "D13",], mapping = aes(x = Day, y = scMutScore, fill = Day)) +
  geom_violin(trim = FALSE) + coord_cartesian(ylim = c(0, 1)) +
  geom_boxplot(width = 0.05, fill = "white", outliers = FALSE) +
  #stat_summary(fun.data="mean_sdl", fun.args = list(mult=1),geom="pointrange", color = "black") +
  scale_fill_manual(values = color_settings_2) +
  labs(x = "", y = paste0("scMutraceScore_", selected_start_cloneID)) +
  theme_classic()

plot(network, 
     vertex.color=my_color_2,
     vertex.label.color = "black",
     vertex.label.cex = 0.8,
     vertex.size = 15,
     #vertex.frame.width = 3,
     #vertex.frame.color = vertex_frame_color,
     edge.color = "black",
     edge.arrow.size = 0.4,
     edge.width = edge_width)

col_fun <- colorRampPalette(c("#fee8c8", "#fdbb84", "#e34a33"))
col_seq <- col_fun(length(my_color_2))
image(z = matrix(1:length(my_color_2), ncol = 1), col = col_seq,
      axes = FALSE, main = "Gradient Legend")
axis(4, at = seq(0, 1, length.out = 5), labels = round(seq(0, 1, length.out = 5), 2))
```

# ModelFigure
```{r testModel}
m1 <- c(1, 1, 1, 1, 1, 1, 1, 1)
m2 <- c(0, 1, 1, 1, 1, 1, 1, 1)
m3 <- c(0, 0, 1, 0, 0, 0, 0, 0)
m4 <- c(0, 0, 0, 1, 1, 1, 1, 1)
m5 <- c(0, 0, 0, 0, 1, 1, 1, 1)
m6 <- c(0, 0, 0, 0, 0, 1, 0, 0)
m7 <- c(0, 0, 0, 0, 0, 0, 1, 0)
m8 <- c(0, 0, 0, 0, 0, 0, 0, 1)


t1 <- c(1, 0, 0, 0, 0, 0, 0, 0)
t2 <- c(1, 1, 0, 0, 0, 0, 0, 0)
t3 <- c(1, 1, 1, 1, 0, 0, 0, 0)
t4 <- c(1, 1, 1, 1, 0, 0, 0, 0)
t5 <- c(1, 1, 0, 1, 1, 0, 0, 0)
t6 <- c(1, 1, 0, 1, 1, 1, 1, 1)

test_Matrix_cm <- data.frame(m1 = m1, m2 = m2, m3 = m3, m4 = m4, m5 = m5, m6 = m6, m7 = m7, m8 = m8)
test_Matrix_mt <- data.frame(t1 = t1, t2 = t2, t3 = t3, t4 =t4, t5 = t5, t6 = t6)
rownames(test_Matrix_cm) <- c("c1", "c2", "c3", "c4", "c5", "c6", "c7", "c8")

test_Matrix_ct <- as.matrix(test_Matrix_cm) %*% as.matrix(test_Matrix_mt)
rownames(test_Matrix_ct) <- c("c1", "c2", "c3", "c4", "c5", "c6", "c7", "c8")

res <- cal_MultiSpe(test_Matrix_ct)

cal_scMutScore(test_Matrix_cm, test_Matrix_mt)

my_colors <- c("#f7f7f7", "#d9d9d9", "#bdbdbd", "#969696", "#636363", "#252525")
breaks <- c(0, 0.5, 1.5, 2.5, 3.5, 4.5, 5.5)

p_Matrix_cm <- pheatmap(test_Matrix_cm, cluster_rows = F, cluster_cols = F, color = my_colors, breaks = breaks, border_color = "black", cellwidth = 10, cellheight = 10)
p_Matrix_mt <- pheatmap(test_Matrix_mt, cluster_rows = F, cluster_cols = F, color = my_colors, breaks = breaks, border_color = "black", cellwidth = 10, cellheight = 10)
p_Matrix_ct <- pheatmap(test_Matrix_ct, cluster_rows = F, cluster_cols = F, color = my_colors, breaks = breaks, border_color = "black", cellwidth = 10, cellheight = 10)

p_Matrix_cm
p_Matrix_mt
p_Matrix_ct
```



# sessionInfo
```{r sessionInfo}
sessionInfo()
```










