---
title: "Figure 4 pAML reproducibility similarity analysis"
author: "Qun Li"
date: "22 Aug 2025 at 10:12:37"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  word_document:
    toc: true
    toc_depth: '3'
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

# Env
```{r envs}
rm(list = ls())
setwd("/Users/liqun/Desktop/scMutrace-tutorial/ReproducibilityAnalysis/Figure4/")
```

# Load libraries
```{r loadLibs}
library(FSA)
library(ineq)
library(grid)
library(vegan)
library(ArchR)
library(dplyr)
library(GO.db)
library(tidyr)
library(ggpubr)
library(Seurat)
library(ggpmisc)
library(stringr)
library(ggplot2)
library(circlize)
library(pheatmap)
library(corrplot)
library(bedtoolsr)
library(tidyverse)
library(patchwork)
library(gridExtra)
library(data.table)
library(UpSetR)
library(ComplexUpset)
library(ggsankey)
library(RColorBrewer)
library(org.Hs.eg.db)
library(GenomicRanges)
library(ComplexHeatmap)
library(clusterProfiler)
library(networkD3)
library(htmlwidgets)
```

# define functions
```{r defineFunctions}
cosine_similarity <- function(a, b, na.rm = TRUE) {
  a <- as.numeric(a)
  b <- as.numeric(b)
  
  if (na.rm) {
    keep <- !(is.na(a) | is.na(b))
    a <- a[keep]
    b <- b[keep]
  }
  
  denom <- sqrt(sum(a^2)) * sqrt(sum(b^2))
  if (denom == 0) return(as.numeric(NA)) 
  
  sim <- sum(a * b) / denom
  
  sim <- max(min(sim, 1), -1)
  
  return(as.numeric(sim))
}

retriveCellsByLocation <- function(CSM, locList){
  CSM$CellTag <- rownames(CSM)
  CSM_tmp <- CSM[, c("CellTag",locList)]
  data_MutPerCell_CSM_tmp <- data.frame(
    CellTag = rownames(CSM_tmp),
    MutNum = rowSums(CSM_tmp == "Mut"),
    WTNum = rowSums(CSM_tmp == "WT")
  )
  cells_Total <- data_MutPerCell_CSM_tmp[rowSums(data_MutPerCell_CSM_tmp[,c("MutNum", "WTNum")]) > 0, ]$CellTag
  cells_WT <-    data_MutPerCell_CSM_tmp[data_MutPerCell_CSM_tmp[,"WTNum"] > 0,]$CellTag
  cells_Mut <-   data_MutPerCell_CSM_tmp[data_MutPerCell_CSM_tmp[,"MutNum"] > 0,]$CellTag
  cells <- list(Total = cells_Total, Mut = cells_Mut, WT = cells_WT)
  return(cells)
}

retriveLocsByCells <- function(CSM, cellList){
  CSM$CellTag <- rownames(CSM)
  CSM_tmp <- CSM[CSM$CellTag %in% cellList,]
  data_CellPerMut_CSM_tmp <- data.frame(
    Location = colnames(CSM_tmp),
    CellMutNum = colSums(CSM_tmp == "Mut"),
    CellWTNum = colSums(CSM_tmp == "WT")
  )
  pos_Total <- data_CellPerMut_CSM_tmp[rowSums(data_CellPerMut_CSM_tmp[, c("CellMutNum", "CellWTNum")]) > 0,]$Location
  pos_Mut <- data_CellPerMut_CSM_tmp[data_CellPerMut_CSM_tmp[, "CellMutNum"] > 0 ,]$Location
  pos_WT <- data_CellPerMut_CSM_tmp[data_CellPerMut_CSM_tmp[, "CellWTNum"] > 0 ,]$Location
  pos <- list(Total = pos_Total, Mut = pos_Mut, WT = pos_WT)
  return(pos)
}

get_mut_cols <- function(df, loc) {
  row <- df[df$Location == loc, ]       
  mut_cols <- names(row)[row == "Mut"] 
  mut_cols[mut_cols != "Location"]     
}
```

*AML_census.rds* and *AML_tiergenes.txt* were downloaded from the COSMIC database and curated for downstream analyses

*AML_SSM.rds* and *AML_ATA_CSM_list.rds*, *AML_RNA_CSM_list.rds* were generated using scMutrace as described in the Methods section and `0_pipelineFunctions.Rmd`(pipelines for generating single-cell somatic mutation data list).

*AML_scRNA_meta.rds* and *AML_scATA_meta.rds* contain meta data for cell type annotation similarity analysis and were generate from `0_pipelineFunctions.Rmd`(pipelines for porcessing scRNA-seq data and pipelines for processing scATAC-seq data).

*sampleInfo.txt* contains sample information including cell numbers and patient clinical data for scRNA-seq and scATAC-seq data and was downloaded from original paper.

*EGA_scATA.hg38_multianno.txt* and *EGA_scRNA.hg38_multianno.txt* contain mutation annotation information generated using ANNOVAR.

*Mutational_Matrix_SBS.txt* and *COSMIC_SBS96_Activities.txt* were generated using SigProfiler suite as described in the Methods section.


# load data
```{r loadData}
# load AML census data
AML_census <- readRDS("../Data/HumanpAML/AML_census.rds")
AML_genes_tiers <- read.table("../Data/HumanpAML/AML_tiergenes.txt", header = T, sep = "\t")

# load SSM data
AML_SSM <- readRDS("../Data/HumanpAML/AML_SSM.rds")
AML_SSM_scRNA <- AML_SSM$scRNA
AML_SSM_scATA <- AML_SSM$scATA

# CSM
AML_ATA_CSM_list <- readRDS("../Data/HumanpAML/AML_ATA_CSM_list.rds")
AML_RNA_CSM_list <- readRDS("../Data/HumanpAML/AML_RNA_CSM_list.rds")

# meta data for celltype annotation similarity
AML_scRNA_meta_list <- readRDS("../Data/HumanpAML/AML_scRNA_meta.rds")
# metaQun
AML_scATA_meta_list <- readRDS("../Data/HumanpAML/AML_scATA_meta.rds")

AML_scRNA_meta <- do.call(rbind, AML_scRNA_meta_list)
rownames(AML_scRNA_meta) <- sub("\\.", "#", rownames(AML_scRNA_meta))

AML_scATA_meta <- do.call(rbind, AML_scATA_meta_list)
AML_scATA_meta$SampleID <- sub("_.*", "", AML_scATA_meta$Sample)

rownames(AML_scATA_meta) <- sub("^AML[0-9]+\\.", "", rownames(AML_scATA_meta))

cellTypes <- unique(union(AML_scRNA_meta$Classified_Celltype, AML_scATA_meta$predictedGroup_Un))

# extract mutations and get bed format with census tag
AML_SSM_scRNA_bed <- separate(AML_SSM_scRNA, Location, into = c("chr", "pos", "ref", "alt"), sep = "_", remove = FALSE)[,1:5]
AML_SSM_scRNA_bed$end <- AML_SSM_scRNA_bed$pos
AML_SSM_scRNA_bed <- AML_SSM_scRNA_bed[,c("chr", "pos", "end", "ref", "alt", "Location")]
AML_SSM_scRNA_bed$Tag <- ifelse(AML_SSM_scRNA_bed$Location %in% AML_census$mutation$Location, "Census", "NotCensus")
AML_scRNA_NotCensus_locs <- AML_SSM_scRNA_bed %>% dplyr::filter(Tag == "NotCensus") %>% pull(Location)

AML_SSM_scATA_bed <- separate(AML_SSM_scATA, Location, into = c("chr", "pos", "ref", "alt"), sep = "_", remove = FALSE)[,1:5]
AML_SSM_scATA_bed$end <- AML_SSM_scATA_bed$pos
AML_SSM_scATA_bed <- AML_SSM_scATA_bed[,c("chr", "pos", "end", "ref", "alt", "Location")]
AML_SSM_scATA_bed$Tag <- ifelse(AML_SSM_scATA_bed$Location %in% AML_census$mutation$Location, "Census", "NotCensus")
AML_scATA_NotCensus_locs <- AML_SSM_scATA_bed %>% dplyr::filter(Tag == "NotCensus") %>% pull(Location)

# load paired sample info data
AML_sampleInfo <- read.table("../Data/HumanpAML/sampleInfo.txt", header = T, sep = "\t")
AML_sampleInfo[is.na(AML_sampleInfo)] <- 0

# AML ann
AML_scATA_ann <- read.table("../Data/HumanpAML/EGA_scATA.hg38_multianno.txt", header = T, sep = "\t")
AML_scRNA_ann <- read.table("../Data/HumanpAML/EGA_scRNA.hg38_multianno.txt", header = T, sep = "\t")

col_freq_databases <- c("gnomad40_exome_AF", "gnomad40_exome_AF_raw", "gnomad40_exome_AF_XX", 
                        "gnomad40_exome_AF_XY", "gnomad40_exome_AF_grpmax", "gnomad40_exome_faf95", 
                        "gnomad40_exome_faf99", "gnomad40_exome_fafmax_faf95_max", "gnomad40_exome_fafmax_faf99_max", 
                        "gnomad40_exome_AF_afr", "gnomad40_exome_AF_amr", "gnomad40_exome_AF_asj", 
                        "gnomad40_exome_AF_eas", "gnomad40_exome_AF_fin", "gnomad40_exome_AF_mid", 
                        "gnomad40_exome_AF_nfe", "gnomad40_exome_AF_remaining", "gnomad40_exome_AF_sas", 
                        "gnomad40_genome_AF", "gnomad40_genome_AF_raw", "gnomad40_genome_AF_XX", 
                        "gnomad40_genome_AF_XY", "gnomad40_genome_AF_grpmax", "gnomad40_genome_faf95", 
                        "gnomad40_genome_faf99", "gnomad40_genome_fafmax_faf95_max", 
                        "gnomad40_genome_fafmax_faf99_max", "gnomad40_genome_AF_afr",
                        "gnomad40_genome_AF_ami", "gnomad40_genome_AF_amr", 
                        "gnomad40_genome_AF_asj", "gnomad40_genome_AF_eas", 
                        "gnomad40_genome_AF_fin", "gnomad40_genome_AF_mid", 
                        "gnomad40_genome_AF_nfe", "gnomad40_genome_AF_remaining", "gnomad40_genome_AF_sas")

col_freq_databases %in% colnames(AML_scATA_ann)
col_freq_databases %in% colnames(AML_scRNA_ann)

AML_scATA_ann <- AML_scATA_ann %>%
  mutate(across(all_of(col_freq_databases), 
                ~ as.numeric(replace(., . == ".", 0))))

AML_scRNA_ann <- AML_scRNA_ann %>%
  mutate(across(all_of(col_freq_databases), 
                ~ as.numeric(replace(., . == ".", 0))))

# Annotate bed census and nonCensus
# scATA
AML_SSM_scATA_bed <- AML_SSM_scATA_bed %>%
  left_join(
    AML_scATA_ann %>%
      dplyr::select(Location, all_of(col_freq_databases)),
    by = "Location"
  ) %>%
  mutate(
    freq = if_else(
      if_any(all_of(col_freq_databases), ~ . > 0.0001),
      "high", "rare"
    )
  ) %>%
  dplyr::select(names(AML_SSM_scATA_bed), freq)

# scRNA
AML_SSM_scRNA_bed <- AML_SSM_scRNA_bed %>%
  left_join(
    AML_scRNA_ann %>%
      dplyr::select(Location, all_of(col_freq_databases)),
    by = "Location"
  ) %>%
  mutate(
    freq = if_else(
      if_any(all_of(col_freq_databases), ~ . > 0.0001),
      "high", "rare"
    )
  ) %>%
  dplyr::select(names(AML_SSM_scRNA_bed), freq)

# load mutation features data
# generate using sigProfiler suite
AML_mutation_features <- read.table("../Data/HumanpAML/Mutational_Matrix_SBS.txt", header = T, sep = "\t", row.names = 1)
AML_mutation_features_contribution <- read.table("../Data/HumanpAML/COSMIC_SBS96_Activities.txt", header = T, sep = "\t")
```
# mutation Similarity DXREL for lineage switching (downsample)
```{r mutationSimilarityDXREL_Data}
AML_scRNA_CSM_list <- AML_RNA_CSM_list
AML_scATA_CSM_list <- AML_ATA_CSM_list

# scRNA
# AML_CSM_DX_list
AML_scRNA_CSM_DX_list <- lapply(AML_scRNA_CSM_list, function(df) {
  df[grep("DX", rownames(df)), , drop = FALSE]
})
AML_scRNA_CSM_DX_list <- Filter(function(x) nrow(x) > 0, AML_scRNA_CSM_DX_list)

# AML_CSM_REM_list
AML_scRNA_CSM_REM_list <- lapply(AML_scRNA_CSM_list, function(df) {
  df[grep("REM", rownames(df)), , drop = FALSE]
})
AML_scRNA_CSM_REM_list <- Filter(function(x) nrow(x) > 0, AML_scRNA_CSM_REM_list)

# AML_CSM_REL_list
AML_scRNA_CSM_REL_list <- lapply(AML_scRNA_CSM_list, function(df) {
  df[grep("REL", rownames(df)), , drop = FALSE]
})
AML_scRNA_CSM_REL_list <- Filter(function(x) nrow(x) > 0, AML_scRNA_CSM_REL_list)

# scATA
# AML_CSM_DX_list
AML_scATA_CSM_DX_list <- lapply(AML_scATA_CSM_list, function(df) {
  df[grep("DX", rownames(df)), , drop = FALSE]
})
AML_scATA_CSM_DX_list <- Filter(function(x) nrow(x) > 0, AML_scATA_CSM_DX_list)

# AML_CSM_REM_list
AML_scATA_CSM_REM_list <- lapply(AML_scATA_CSM_list, function(df) {
  df[grep("REM", rownames(df)), , drop = FALSE]
})
AML_scATA_CSM_REM_list <- Filter(function(x) nrow(x) > 0, AML_scATA_CSM_REM_list)

# AML_CSM_REL_list
AML_scATA_CSM_REL_list <- lapply(AML_scATA_CSM_list, function(df) {
  df[grep("REL", rownames(df)), , drop = FALSE]
})
AML_scATA_CSM_REL_list <- Filter(function(x) nrow(x) > 0, AML_scATA_CSM_REL_list)

HSPC <- c("HSC", "Progenitor")
Myeloid <- c("GMP", "Monocytes", "CD16.Monocytes", "cDC", "pDC")
Lymphoid <- c("CLP", "Pre.B.Cell", "B.Cell", "CD4.Naive", "CD4.Memory", "CD8.Naive", "CD8.Memory", "CD8.Effector", "NK")
Erythroid_Basophil <- c("Early.Erythrocyte", "Late.Erythrocyte")
Other <- c("Early.Basophil", "Plasma")

AML_scRNA_meta <- AML_scRNA_meta %>%
  dplyr::mutate(
    maincelltype = case_when(
      Classified_Celltype %in% HSPC ~ "HSPC",
      Classified_Celltype %in% Myeloid ~ "Myeloid",
      Classified_Celltype %in% Lymphoid ~ "Lymphoid",
      Classified_Celltype %in% Erythroid_Basophil ~ "Erythroid",
      Classified_Celltype %in% Other ~ "Other",
      TRUE ~ NA_character_
    )
  )

AML_scATA_meta <- AML_scATA_meta %>%
  dplyr::mutate(
    maincelltype = case_when(
      predictedGroup_Un %in% HSPC ~ "HSPC",
      predictedGroup_Un %in% Myeloid ~ "Myeloid",
      predictedGroup_Un %in% Lymphoid ~ "Lymphoid",
      predictedGroup_Un %in% Erythroid_Basophil ~ "Erythroid",
      predictedGroup_Un %in% Other ~ "Other",
      TRUE ~ NA_character_
    )
  )

AML_scATA_meta$Classified_Celltype <- AML_scATA_meta$predictedGroup_Un

maincelltypes <- c("HSPC", "Myeloid", "Lymphoid", "Erythroid", "Other")
subcelltypes <- c(HSPC, Myeloid, Lymphoid, Erythroid_Basophil, Other)

AML_scRNA_meta_DX <- AML_scRNA_meta %>%
  dplyr::filter(grepl("DX", Library_ID))

AML_scRNA_meta_REL <- AML_scRNA_meta %>%
  dplyr::filter(grepl("REL", Library_ID))

AML_scATA_meta_DX <- AML_scATA_meta %>%
  dplyr::filter(grepl("DX", Sample))

AML_scATA_meta_REL <- AML_scATA_meta %>%
  dplyr::filter(grepl("REL", Sample))

selected_samples_id <- c("AML4", "AML5", "AML6", "AML8",  "AML10", "AML12", "AML16", "AML18", "AML21")
selected_samples_id <- c("AML2", "AML4", "AML5", "AML6", "AML8", "AML10", "AML11", "AML12",
                         "AML13", "AML15", "AML16", "AML17", "AML18", "AML19", "AML21",
                         "AML22", "AML24", "AML25", "AML26", "AML27")

AML_scRNA_CSM_DX_select_list <- AML_scRNA_CSM_DX_list[names(AML_scRNA_CSM_DX_list) %in% selected_samples_id]
names(AML_scRNA_CSM_DX_select_list) <- paste(selected_samples_id, "DX", sep = "_")
AML_scRNA_CSM_REL_select_list <- AML_scRNA_CSM_REL_list[names(AML_scRNA_CSM_REL_list) %in% selected_samples_id]
names(AML_scRNA_CSM_REL_select_list) <- paste(selected_samples_id, "REL", sep = "_")
AML_scRNA_CSM_DX_REL_select_list <- c(AML_scRNA_CSM_DX_select_list, AML_scRNA_CSM_REL_select_list)

AML_scRNA_meta_DX$Classified_CelltypeQ <- paste(AML_scRNA_meta_DX$Lambo_et_al_ID, "DX", AML_scRNA_meta_DX$Classified_Celltype, sep = "_")
AML_scRNA_meta_REL$Classified_CelltypeQ <- paste(AML_scRNA_meta_REL$Lambo_et_al_ID, "REL", AML_scRNA_meta_REL$Classified_Celltype, sep = "_")
AML_scRNA_meta_DX$maincelltypeQ <- paste(AML_scRNA_meta_DX$Lambo_et_al_ID, "DX", AML_scRNA_meta_DX$maincelltype, sep = "_")
AML_scRNA_meta_REL$maincelltypeQ <- paste(AML_scRNA_meta_REL$Lambo_et_al_ID, "REL", AML_scRNA_meta_REL$maincelltype, sep = "_")

AML_scATA_CSM_DX_select_list <- AML_scATA_CSM_DX_list[names(AML_scATA_CSM_DX_list) %in% selected_samples_id]
names(AML_scATA_CSM_DX_select_list) <- paste(selected_samples_id, "DX", sep = "_")
AML_scATA_CSM_REL_select_list <- AML_scATA_CSM_REL_list[names(AML_scATA_CSM_REL_list) %in% selected_samples_id]
names(AML_scATA_CSM_REL_select_list) <- paste(selected_samples_id, "REL", sep = "_")
AML_scATA_CSM_DX_REL_select_list <- c(AML_scATA_CSM_DX_select_list, AML_scATA_CSM_REL_select_list)
AML_scATA_meta_DX$Classified_CelltypeQ <- paste(AML_scATA_meta_DX$SampleID, "DX", AML_scATA_meta_DX$Classified_Celltype, sep = "_")
AML_scATA_meta_REL$Classified_CelltypeQ <- paste(AML_scATA_meta_REL$SampleID, "REL", AML_scATA_meta_REL$Classified_Celltype, sep = "_")

AML_scRNA_meta_DX_REL <- rbind(AML_scRNA_meta_DX, AML_scRNA_meta_REL)
AML_scATA_meta_DX_REL <- rbind(AML_scATA_meta_DX, AML_scATA_meta_REL)

# samples for lineage switching

AML_scRNA_meta_DX_REL <- AML_scRNA_meta_DX_REL[AML_scRNA_meta_DX_REL$Lambo_et_al_ID %in% selected_samples_id, ]
AML_scRNA_meta_DX_REL$CellTag <- rownames(AML_scRNA_meta_DX_REL) %>%
                                    sub("-1$", "", .) %>%
                                    gsub("#", "_", .)
AML_scRNA_meta$CellTag <- rownames(AML_scRNA_meta) %>%
                                    sub("-1$", "", .) %>%
                                    gsub("#", "_", .)

AML_scRNA_meta_DX_REL$SampleID <- AML_scRNA_meta_DX_REL$Lambo_et_al_ID
AML_scATA_meta_DX_REL <- AML_scATA_meta_DX_REL[AML_scATA_meta_DX_REL$SampleID %in% selected_samples_id,]
AML_scATA_meta_DX_REL$CellTag <- rownames(AML_scATA_meta_DX_REL) %>%
                                    sub("-1$", "", .) %>%
                                    gsub("#", "_", .)
AML_scATA_meta$CellTag <- rownames(AML_scATA_meta) %>%
                                    sub("-1$", "", .) %>%
                                    gsub("#", "_", .)

AML_scATA_meta_DX_REL$maincelltypeQ <- paste(AML_scATA_meta_DX_REL$Sample, AML_scATA_meta_DX_REL$maincelltype, sep = "_")

cells_limits <- 0

celltypes_scRNA <- names(table(AML_scRNA_meta_DX_REL$Classified_CelltypeQ)[table(AML_scRNA_meta_DX_REL$Classified_CelltypeQ) >= cells_limits])
celltypes_scATA <- names(table(AML_scATA_meta_DX_REL$Classified_CelltypeQ)[table(AML_scATA_meta_DX_REL$Classified_CelltypeQ) >= cells_limits])

celltypes_scRNA_main <- names(table(AML_scRNA_meta_DX_REL$maincelltypeQ)[table(AML_scRNA_meta_DX_REL$maincelltypeQ) >= cells_limits])
celltypes_scATA_main <- names(table(AML_scATA_meta_DX_REL$maincelltypeQ)[table(AML_scATA_meta_DX_REL$maincelltypeQ) >= cells_limits])
```

# calSimilarity downsampling
```{r test}
iteration_times <- 1:50
sample_size_f <- 0.30
sample_size <- 200

set.seed(2025)

#selected_samples_id <- c("AML4", "AML6", "AML8", "AML10", "AML16", "AML27")
#selected_samples_id <- c("AML4", "AML5", "AML6", "AML8",  "AML10",  "AML12", "AML16", "AML18", "AML21")
#selected_samples_id <- c("AML13")
selected_samples_id <- c("AML2", "AML4", "AML5", "AML6", "AML8", "AML10", "AML11", "AML12",
                         "AML13", "AML15", "AML16", "AML17", "AML18", "AML19", "AML21",
                         "AML22", "AML24", "AML25", "AML26", "AML27")
stage_dx_rel <- c("DX", "REL")
celltypes_all <- c(HSPC, Myeloid, Lymphoid, Erythroid_Basophil)

df_combination <- expand.grid(
  sample = selected_samples_id,
  stage  = stage_dx_rel,
  celltype = celltypes_all,
  stringsAsFactors = FALSE
)

result_all <- paste(df_combination$sample, df_combination$stage, df_combination$celltype, sep = "_")
celltypes_scRNA <- result_all
celltypes_scATA <- result_all

mut_cellTypeAnn_similarity_list_RNA <- vector("list", length(selected_samples_id))
names(mut_cellTypeAnn_similarity_list_RNA) <- selected_samples_id

mut_cellTypeAnn_similarity_list_RNA_main <- vector("list", length(selected_samples_id))
names(mut_cellTypeAnn_similarity_list_RNA_main) <- selected_samples_id

mut_cellTypeAnn_similarity_list_ATA <- vector("list", length(selected_samples_id))
names(mut_cellTypeAnn_similarity_list_ATA) <- selected_samples_id

mut_cellTypeAnn_similarity_list_ATA_main <- vector("list", length(selected_samples_id))
names(mut_cellTypeAnn_similarity_list_ATA_main) <- selected_samples_id

DX_HSPC <- paste0("DX_", HSPC)
DX_myeloid <- paste0("DX_", Myeloid)
DX_lymphoid <- paste0("DX_", Lymphoid)
DX_Erythroid <- paste0("DX_", Erythroid_Basophil)
REL_HSPC <- paste0("REL_", HSPC)
REL_myeloid <- paste0("REL_", Myeloid)
REL_lymphoid <- paste0("REL_", Lymphoid)
REL_Erythroid <- paste0("REL_", Erythroid_Basophil)

# RNA
for (sample_iter in selected_samples_id) {
  message("process.. ", sample_iter)

  mut_cellTypeAnn_similarity_list_sample <- vector("list", length(iteration_times))
  names(mut_cellTypeAnn_similarity_list_sample) <- paste0("iter_", iteration_times)

  celltypes_scRNA_sample <- grep(paste0("^", sample_iter, "(_|$)"), celltypes_scRNA, value = TRUE)

  for (iter_i in iteration_times) {
    message("  process.. ", iter_i)

    mut_cellTypeAnn_similarity_tmp <- matrix(
      0,
      nrow = length(celltypes_scRNA_sample),
      ncol = length(celltypes_scRNA_sample),
      dimnames = list(celltypes_scRNA_sample, celltypes_scRNA_sample)
    )

    # cache mut sites per celltype (IMPORTANT SPEED UP)
    mut_cache <- list()

    for (ct in celltypes_scRNA_sample) {
      cells_ct <- AML_scRNA_meta_DX_REL |>
        dplyr::filter(Classified_CelltypeQ == ct) |>
        dplyr::pull(CellTag)
      
      #sample_ct <- sub("_.*", "", ct)
      
      #cells_ct <- AML_scRNA_meta_DX_REL |>
      #  dplyr::filter(SampleID == sample_ct) |>
      #  dplyr::pull(CellTag)
      
      if (length(cells_ct) == 0) {
        mut_cache[[ct]] <- character(0)
        next
      }
      # num
      #cells_ct <- sample(cells_ct, min(length(cells_ct), sample_size))
      # frac
      cells_ct <- sample(cells_ct, max(1, floor(length(cells_ct) * sample_size_f)))

      sample_id <- sub("^([^_]+_[^_]+)_.*", "\\1", ct)
      
      # modify mapping
      cells_csm_raw <- rownames(AML_scRNA_CSM_DX_REL_select_list[[sample_id]])
      cells_csm <- gsub("_L[0-9]+_", "_", cells_csm_raw)
      df_map <- data.frame(raw = cells_csm_raw, clean = cells_csm, stringsAsFactors = FALSE)
      cells_raw_selected <- df_map$raw[df_map$clean %in% cells_ct]
      
      mut_cache[[ct]] <- retriveLocsByCells(
        AML_scRNA_CSM_DX_REL_select_list[[sample_id]],
        cells_raw_selected
      )$Mut
    }

    # compute only upper triangle
    for (i in seq_along(celltypes_scRNA_sample)) {
      for (j in i:length(celltypes_scRNA_sample)) {

        ct_row <- celltypes_scRNA_sample[i]
        ct_col <- celltypes_scRNA_sample[j]

        m1 <- mut_cache[[ct_row]]
        m2 <- mut_cache[[ct_col]]

        if (length(m1) == 0 || length(m2) == 0) {
          score <- 0
        } else {
          u <- union(m1, m2)
          score <- length(intersect(m1, m2)) / length(u)
        }

        mut_cellTypeAnn_similarity_tmp[i, j] <- score
        mut_cellTypeAnn_similarity_tmp[j, i] <- score
      }
    }

    mut_cellTypeAnn_similarity_list_sample[[paste0("iter_", iter_i)]] <-
      mut_cellTypeAnn_similarity_tmp
  }

  mut_cellTypeAnn_similarity_list_RNA[[sample_iter]] <- mut_cellTypeAnn_similarity_list_sample
}

# RNA main
for (sample_iter in selected_samples_id) {
  message("process.. ", sample_iter)

  mut_cellTypeAnn_similarity_list_sample <- vector("list", length(iteration_times))
  names(mut_cellTypeAnn_similarity_list_sample) <- paste0("iter_", iteration_times)

  celltypes_scRNA_sample <- grep(paste0("^", sample_iter, "(_|$)"), celltypes_scRNA_main, value = TRUE)

  for (iter_i in iteration_times) {
    message("  process.. ", iter_i)

    mut_cellTypeAnn_similarity_tmp <- matrix(
      0,
      nrow = length(celltypes_scRNA_sample),
      ncol = length(celltypes_scRNA_sample),
      dimnames = list(celltypes_scRNA_sample, celltypes_scRNA_sample)
    )

    # cache mut sites per celltype (IMPORTANT SPEED UP)
    mut_cache <- list()

    for (ct in celltypes_scRNA_sample) {
      cells_ct <- AML_scRNA_meta_DX_REL |>
        dplyr::filter(maincelltypeQ == ct) |>
        dplyr::pull(CellTag)
      
      #sample_ct <- sub("_.*", "", ct)
      
      #cells_ct <- AML_scRNA_meta_DX_REL |>
      #  dplyr::filter(SampleID == sample_ct) |>
      #  dplyr::pull(CellTag)
      
      if (length(cells_ct) == 0) {
        mut_cache[[ct]] <- character(0)
        next
      }
      # num
      #cells_ct <- sample(cells_ct, min(length(cells_ct), sample_size))
      # frac
      cells_ct <- sample(cells_ct, max(1, floor(length(cells_ct) * sample_size_f)))

      sample_id <- sub("^([^_]+_[^_]+)_.*", "\\1", ct)
      
      # modify mapping
      cells_csm_raw <- rownames(AML_scRNA_CSM_DX_REL_select_list[[sample_id]])
      cells_csm <- gsub("_L[0-9]+_", "_", cells_csm_raw)
      df_map <- data.frame(raw = cells_csm_raw, clean = cells_csm, stringsAsFactors = FALSE)
      cells_raw_selected <- df_map$raw[df_map$clean %in% cells_ct]
      
      mut_cache[[ct]] <- retriveLocsByCells(
        AML_scRNA_CSM_DX_REL_select_list[[sample_id]],
        cells_raw_selected
      )$Mut
    }

    # compute only upper triangle
    for (i in seq_along(celltypes_scRNA_sample)) {
      for (j in i:length(celltypes_scRNA_sample)) {

        ct_row <- celltypes_scRNA_sample[i]
        ct_col <- celltypes_scRNA_sample[j]

        m1 <- mut_cache[[ct_row]]
        m2 <- mut_cache[[ct_col]]

        if (length(m1) == 0 || length(m2) == 0) {
          score <- 0
        } else {
          u <- union(m1, m2)
          score <- length(intersect(m1, m2)) / length(u)
        }

        mut_cellTypeAnn_similarity_tmp[i, j] <- score
        mut_cellTypeAnn_similarity_tmp[j, i] <- score
      }
    }

    mut_cellTypeAnn_similarity_list_sample[[paste0("iter_", iter_i)]] <-
      mut_cellTypeAnn_similarity_tmp
  }

  mut_cellTypeAnn_similarity_list_RNA_main[[sample_iter]] <- mut_cellTypeAnn_similarity_list_sample
}

# ATA
for (sample_iter in selected_samples_id) {
  message("process.. ", sample_iter)

  mut_cellTypeAnn_similarity_list_sample <- vector("list", length(iteration_times))
  names(mut_cellTypeAnn_similarity_list_sample) <- paste0("iter_", iteration_times)

  celltypes_scATA_sample <- grep(paste0("^", sample_iter, "(_|$)"), celltypes_scATA, value = TRUE)

  for (iter_i in iteration_times) {
    message("  process.. ", iter_i)

    mut_cellTypeAnn_similarity_tmp <- matrix(
      0,
      nrow = length(celltypes_scATA_sample),
      ncol = length(celltypes_scATA_sample),
      dimnames = list(celltypes_scATA_sample, celltypes_scATA_sample)
    )

    # cache mut sites per celltype (IMPORTANT SPEED UP)
    mut_cache <- list()

    for (ct in celltypes_scATA_sample) {
      cells_ct <- AML_scATA_meta_DX_REL |>
        dplyr::filter(Classified_CelltypeQ == ct) |>
        dplyr::pull(CellTag)
      
      #sample_ct <- sub("_.*", "", ct)
      
      if (length(cells_ct) == 0) {
        mut_cache[[ct]] <- character(0)
        next
      }

      #cells_ct <- sample(cells_ct, min(length(cells_ct), sample_size))
      cells_ct <- sample(cells_ct, max(1, floor(length(cells_ct) * sample_size_f)))

      sample_id <- sub("^([^_]+_[^_]+)_.*", "\\1", ct)

      mut_cache[[ct]] <- retriveLocsByCells(
        AML_scATA_CSM_DX_REL_select_list[[sample_id]],
        cells_ct
      )$Mut
    }

    # compute only upper triangle
    for (i in seq_along(celltypes_scATA_sample)) {
      for (j in i:length(celltypes_scATA_sample)) {

        ct_row <- celltypes_scATA_sample[i]
        ct_col <- celltypes_scATA_sample[j]

        m1 <- mut_cache[[ct_row]]
        m2 <- mut_cache[[ct_col]]

        if (length(m1) == 0 || length(m2) == 0) {
          score <- 0
        } else {
          u <- union(m1, m2)
          score <- length(intersect(m1, m2)) / length(u)
        }

        mut_cellTypeAnn_similarity_tmp[i, j] <- score
        mut_cellTypeAnn_similarity_tmp[j, i] <- score
      }
    }

    mut_cellTypeAnn_similarity_list_sample[[paste0("iter_", iter_i)]] <-
      mut_cellTypeAnn_similarity_tmp
  }

  mut_cellTypeAnn_similarity_list_ATA[[sample_iter]] <- mut_cellTypeAnn_similarity_list_sample
}

# ATA main
for (sample_iter in selected_samples_id) {
  message("process.. ", sample_iter)

  mut_cellTypeAnn_similarity_list_sample <- vector("list", length(iteration_times))
  names(mut_cellTypeAnn_similarity_list_sample) <- paste0("iter_", iteration_times)

  celltypes_scATA_sample <- grep(paste0("^", sample_iter, "(_|$)"), celltypes_scATA_main, value = TRUE)

  for (iter_i in iteration_times) {
    message("  process.. ", iter_i)

    mut_cellTypeAnn_similarity_tmp <- matrix(
      0,
      nrow = length(celltypes_scATA_sample),
      ncol = length(celltypes_scATA_sample),
      dimnames = list(celltypes_scATA_sample, celltypes_scATA_sample)
    )

    # cache mut sites per celltype (IMPORTANT SPEED UP)
    mut_cache <- list()

    for (ct in celltypes_scATA_sample) {
      cells_ct <- AML_scATA_meta_DX_REL |>
        dplyr::filter(maincelltypeQ == ct) |>
        dplyr::pull(CellTag)
      
      #sample_ct <- sub("_.*", "", ct)
      
      if (length(cells_ct) == 0) {
        mut_cache[[ct]] <- character(0)
        next
      }

      #cells_ct <- sample(cells_ct, min(length(cells_ct), sample_size))
      cells_ct <- sample(cells_ct, max(1, floor(length(cells_ct) * sample_size_f)))

      sample_id <- sub("^([^_]+_[^_]+)_.*", "\\1", ct)

      mut_cache[[ct]] <- retriveLocsByCells(
        AML_scATA_CSM_DX_REL_select_list[[sample_id]],
        cells_ct
      )$Mut
    }

    # compute only upper triangle
    for (i in seq_along(celltypes_scATA_sample)) {
      for (j in i:length(celltypes_scATA_sample)) {

        ct_row <- celltypes_scATA_sample[i]
        ct_col <- celltypes_scATA_sample[j]

        m1 <- mut_cache[[ct_row]]
        m2 <- mut_cache[[ct_col]]

        if (length(m1) == 0 || length(m2) == 0) {
          score <- 0
        } else {
          u <- union(m1, m2)
          score <- length(intersect(m1, m2)) / length(u)
        }

        mut_cellTypeAnn_similarity_tmp[i, j] <- score
        mut_cellTypeAnn_similarity_tmp[j, i] <- score
      }
    }

    mut_cellTypeAnn_similarity_list_sample[[paste0("iter_", iter_i)]] <-
      mut_cellTypeAnn_similarity_tmp
  }

  mut_cellTypeAnn_similarity_list_ATA_main[[sample_iter]] <- mut_cellTypeAnn_similarity_list_sample
}

mut_cellTypeAnn_similarity_mean_RNA <- lapply(mut_cellTypeAnn_similarity_list_RNA, function(sample_list) {
  Reduce("+", sample_list) / length(sample_list)
})

mut_cellTypeAnn_similarity_mean_RNA_main <- lapply(mut_cellTypeAnn_similarity_list_RNA_main, function(sample_list) {
  Reduce("+", sample_list) / length(sample_list)
})

mut_cellTypeAnn_similarity_mean_ATA <- lapply(mut_cellTypeAnn_similarity_list_ATA, function(sample_list) {
  Reduce("+", sample_list) / length(sample_list)
})

mut_cellTypeAnn_similarity_mean_ATA_main <- lapply(mut_cellTypeAnn_similarity_list_ATA_main, function(sample_list) {
  Reduce("+", sample_list) / length(sample_list)
})

sort_similarity_matrix <- function(mat,
                                   HSPC,
                                   Myeloid,
                                   Lymphoid,
                                   Erythroid_Basophil,
                                   Other) {

  ct_names <- rownames(mat)

  patient  <- sub("^(AML\\d+)_.*", "\\1", ct_names)
  stage    <- sub("^AML\\d+_(DX|REL)_.*", "\\1", ct_names)
  celltype <- sub("^AML\\d+_(DX|REL)_", "", ct_names)

  celltype2lineage <- c(
    setNames(rep("HSPC", length(HSPC)), HSPC),
    setNames(rep("Myeloid", length(Myeloid)), Myeloid),
    setNames(rep("Lymphoid", length(Lymphoid)), Lymphoid),
    setNames(rep("Erythroid_Basophil", length(Erythroid_Basophil)), Erythroid_Basophil),
    setNames(rep("Other", length(Other)), Other)
  )

  lineage <- celltype2lineage[celltype]

  stage <- factor(stage, levels = c("DX", "REL"))

  lineage_order <- c("HSPC", "Myeloid", "Lymphoid", "Erythroid_Basophil", "Other")
  lineage <- factor(lineage, levels = lineage_order)

  celltype_levels <- c(HSPC, Myeloid, Lymphoid, Erythroid_Basophil, Other)
  celltype <- factor(celltype, levels = celltype_levels)

  ord <- order(patient, stage, lineage, celltype)

  mat_sorted <- mat[ord, ord]

  # -------- annotation --------
  annotation <- data.frame(
    Patient = patient,
    Stage = stage,
    Lineage = lineage,
    Celltype = celltype
  )
  rownames(annotation) <- ct_names
  annotation <- annotation[ord, , drop = FALSE]

  return(list(
    mat = mat_sorted,
    annotation = annotation,
    order = ord
  ))
}

saveRDS(mut_cellTypeAnn_similarity_list_RNA, "../Data/HumanpAML/mut_cellTypeAnn_similarity_RNA_subcell_byfrac30.rds")
saveRDS(mut_cellTypeAnn_similarity_list_ATA, "../Data/HumanpAML/mut_cellTypeAnn_similarity_ATA_subcell_byfrac30.rds")
saveRDS(mut_cellTypeAnn_similarity_list_RNA_main, "../Data/HumanpAML/mut_cellTypeAnn_similarity_RNA_maincell_byfrac30.rds")
saveRDS(mut_cellTypeAnn_similarity_list_ATA_main, "../Data/HumanpAML/mut_cellTypeAnn_similarity_ATA_maincell_byfrac30.rds")

# RNA
res_AML4_RNA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_RNA$AML4,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML5_RNA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_RNA$AML5,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML6_RNA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_RNA$AML6,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML8_RNA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_RNA$AML8,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML10_RNA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_RNA$AML10,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML12_RNA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_RNA$AML12,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML16_RNA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_RNA$AML16,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML18_RNA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_RNA$AML18,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML21_RNA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_RNA$AML21,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

# ATA
res_AML4_ATA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_ATA$AML4,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML5_ATA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_ATA$AML5,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML6_ATA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_ATA$AML6,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML8_ATA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_ATA$AML8,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML10_ATA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_ATA$AML10,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML12_ATA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_ATA$AML12,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML16_ATA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_ATA$AML16,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML18_ATA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_ATA$AML18,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

res_AML21_ATA <- sort_similarity_matrix(
  mat = mut_cellTypeAnn_similarity_mean_ATA$AML21,
  HSPC = HSPC,
  Myeloid = Myeloid,
  Lymphoid = Lymphoid,
  Erythroid_Basophil = Erythroid_Basophil,
  Other = Other
)

mat_sorted_AML04_RNA <- res_AML4_RNA$mat
mat_sorted_AML16_RNA <- res_AML16_RNA$mat
mat_sorted_AML05_RNA <- res_AML5_RNA$mat
mat_sorted_AML06_RNA <- res_AML6_RNA$mat
mat_sorted_AML08_RNA <- res_AML8_RNA$mat
mat_sorted_AML10_RNA <- res_AML10_RNA$mat
mat_sorted_AML12_RNA <- res_AML12_RNA$mat
mat_sorted_AML18_RNA <- res_AML18_RNA$mat
mat_sorted_AML21_RNA <- res_AML21_RNA$mat

mat_sorted_AML04_ATA <- res_AML4_ATA$mat
mat_sorted_AML16_ATA <- res_AML16_ATA$mat
mat_sorted_AML05_ATA <- res_AML5_ATA$mat
mat_sorted_AML06_ATA <- res_AML6_ATA$mat
mat_sorted_AML08_ATA <- res_AML8_ATA$mat
mat_sorted_AML10_ATA <- res_AML10_ATA$mat
mat_sorted_AML12_ATA <- res_AML12_ATA$mat
mat_sorted_AML18_ATA <- res_AML18_ATA$mat
mat_sorted_AML21_ATA <- res_AML21_ATA$mat

celltype_comm_AML04 <- intersect(colnames(mat_sorted_AML04_ATA), colnames(mat_sorted_AML04_RNA))
celltype_comm_AML05 <- intersect(colnames(mat_sorted_AML05_ATA), colnames(mat_sorted_AML05_RNA))
celltype_comm_AML06 <- intersect(colnames(mat_sorted_AML06_ATA), colnames(mat_sorted_AML06_RNA))
celltype_comm_AML08 <- intersect(colnames(mat_sorted_AML08_ATA), colnames(mat_sorted_AML08_RNA))
celltype_comm_AML10 <- intersect(colnames(mat_sorted_AML10_ATA), colnames(mat_sorted_AML10_RNA))
celltype_comm_AML12 <- intersect(colnames(mat_sorted_AML12_ATA), colnames(mat_sorted_AML12_RNA))
celltype_comm_AML16 <- intersect(colnames(mat_sorted_AML16_ATA), colnames(mat_sorted_AML16_RNA))
celltype_comm_AML18 <- intersect(colnames(mat_sorted_AML18_ATA), colnames(mat_sorted_AML18_RNA))
celltype_comm_AML21 <- intersect(colnames(mat_sorted_AML21_ATA), colnames(mat_sorted_AML21_RNA))

# RNA
corrplot(mat_sorted_AML04_RNA, method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML16_RNA, method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML05_RNA, method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML06_RNA, method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML08_RNA, method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML10_RNA, method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML12_RNA, method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML18_RNA, method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")

# ATA
corrplot(mat_sorted_AML04_ATA, method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML16_ATA, method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML05_ATA, method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML06_ATA, method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML08_ATA, method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML10_ATA, method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML12_ATA, method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML18_ATA, method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
```

```{r saveplot}
# RNA
corrplot(mat_sorted_AML04_RNA[celltype_comm_AML04, celltype_comm_AML04], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML16_RNA[celltype_comm_AML16, celltype_comm_AML16], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML05_RNA[celltype_comm_AML05, celltype_comm_AML05], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML06_RNA[celltype_comm_AML06, celltype_comm_AML06], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML08_RNA[celltype_comm_AML08, celltype_comm_AML08], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML10_RNA[celltype_comm_AML10, celltype_comm_AML10], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML18_RNA[celltype_comm_AML18, celltype_comm_AML18], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")

# ATA
corrplot(mat_sorted_AML04_ATA[celltype_comm_AML04, celltype_comm_AML04], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML16_ATA[celltype_comm_AML16, celltype_comm_AML16], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML05_ATA[celltype_comm_AML05, celltype_comm_AML05], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML06_ATA[celltype_comm_AML06, celltype_comm_AML06], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML08_ATA[celltype_comm_AML08, celltype_comm_AML08], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML10_ATA[celltype_comm_AML10, celltype_comm_AML10], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
corrplot(mat_sorted_AML18_ATA[celltype_comm_AML18, celltype_comm_AML18], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")


# save
pdf("../corrplot_AML04_RNA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML04_RNA[celltype_comm_AML04, celltype_comm_AML04], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML16_RNA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML16_RNA[celltype_comm_AML16, celltype_comm_AML16], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML05_RNA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML05_RNA[celltype_comm_AML05, celltype_comm_AML05], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML06_RNA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML06_RNA[celltype_comm_AML06, celltype_comm_AML06], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML08_RNA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML08_RNA[celltype_comm_AML08, celltype_comm_AML08], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML10_RNA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML10_RNA[celltype_comm_AML10, celltype_comm_AML10], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML12_RNA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML12_RNA[celltype_comm_AML12, celltype_comm_AML12], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML18_RNA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML18_RNA[celltype_comm_AML18, celltype_comm_AML18], method = 'color', 
         type = 'lower', col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML04_ATA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML04_ATA[celltype_comm_AML04, celltype_comm_AML04], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML16_ATA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML16_ATA[celltype_comm_AML16, celltype_comm_AML16], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML05_ATA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML05_ATA[celltype_comm_AML05, celltype_comm_AML05], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML06_ATA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML06_ATA[celltype_comm_AML06, celltype_comm_AML06], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML08_ATA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML08_ATA[celltype_comm_AML08, celltype_comm_AML08], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML10_ATA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML10_ATA[celltype_comm_AML10, celltype_comm_AML10], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML12_ATA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML12_ATA[celltype_comm_AML12, celltype_comm_AML12], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
dev.off()

pdf("../corrplot_AML18_ATA.pdf", width = 8, height = 8)
corrplot(mat_sorted_AML18_ATA[celltype_comm_AML18, celltype_comm_AML18], method = 'color', 
         type = 'upper', col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), order = "original")
dev.off()
```

# permutation test
```{r pertest}
iteration_times <- 1:50
sample_size_f <- 0.50
sample_size <- 400

mut_cellTypeAnn_similarity_list_RNA_main <- readRDS("../Data/HumanpAML/mut_cellTypeAnn_similarity_RNA_maincell_byfrac30.rds")
mut_cellTypeAnn_similarity_list_ATA_main <- readRDS("../Data/HumanpAML/mut_cellTypeAnn_similarity_ATA_maincell_byfrac30.rds")

paired_perm_test_viz <- function(x, y, n_perm = 1000) {

  diff <- x - y
  obs <- mean(diff)

  # permutation null
  perm <- replicate(n_perm, {
    mean(diff * sample(c(-1, 1), length(diff), replace = TRUE))
  })

  p <- mean(abs(perm) >= abs(obs))

  # CI from null distribution
  ci <- quantile(perm, c(0.025, 0.975))

  # return structured object for plotting
  list(
    obs = obs,
    p_value = p,
    ci = ci,
    perm_distribution = perm,
    diff_values = diff
  )
}

unpaired_perm_test_viz <- function(x, y, n_perm = 1000) {

  obs <- mean(x) - mean(y)
  combined <- c(x, y)
  n_x <- length(x)

  perm <- replicate(n_perm, {
    idx <- sample(length(combined))
    mean(combined[idx[1:n_x]]) - mean(combined[idx[-(1:n_x)]])
  })

  p <- mean(abs(perm) >= abs(obs))
  ci <- quantile(perm, c(0.025, 0.975))

  list(
    obs = obs,
    p_value = p,
    ci = ci,
    perm_distribution = perm
  )
}

plot_perm_test <- function(res, title = "Permutation Test") {

  df <- data.frame(perm = res$perm_distribution)

  ggplot2::ggplot(df, ggplot2::aes(x = perm)) +
    ggplot2::geom_histogram(bins = 50, fill = "grey80", color = "black") +
    ggplot2::geom_vline(xintercept = res$obs, color = "red", size = 1) +
    ggplot2::geom_vline(xintercept = res$ci, linetype = "dashed", color = "blue") +
    ggplot2::labs(
      title = title,
      subtitle = paste0("p = ", signif(res$p_value, 3)),
      x = "Permutation mean difference",
      y = "Frequency"
    ) +
    ggplot2::theme_classic()
}



cal_sigValSimilarity <- function(mut_list, iteration_times, sampleid){
  # 1
  vals_RelLym_DxMye <- sapply(
    paste0("iter_", iteration_times),
    function(x) mut_list[[sampleid]][[x]][paste(sampleid, "REL_Lymphoid", sep = "_"), paste(sampleid, "DX_Myeloid", sep = "_")]
  )
  
  # 2
  vals_RelMye_DxMye <- sapply(
    paste0("iter_", iteration_times),
    function(x) mut_list[[sampleid]][[x]][paste(sampleid, "REL_Myeloid", sep = "_"), paste(sampleid, "DX_Myeloid", sep = "_")]
  )
  
  # 3
  vals_DxLym_RelMye <- sapply(
    paste0("iter_", iteration_times),
    function(x) mut_list[[sampleid]][[x]][paste(sampleid, "DX_Lymphoid", sep = "_"), paste(sampleid, "REL_Myeloid", sep = "_")]
  )
  
  # 4
  vals_DxLym_RelLym <- sapply(
    paste0("iter_", iteration_times),
    function(x) mut_list[[sampleid]][[x]][paste(sampleid, "DX_Lymphoid", sep = "_"), paste(sampleid, "REL_Lymphoid", sep = "_")]
  )
  
  # 5
  vals_DxMye_DxLym <- sapply(
    paste0("iter_", iteration_times),
    function(x) mut_list[[sampleid]][[x]][paste(sampleid, "DX_Myeloid", sep = "_"), paste(sampleid, "DX_Lymphoid", sep = "_")]
  )
  
  # 6
  vals_DxMye_DxLym <- sapply(
    paste0("iter_", iteration_times),
    function(x) mut_list[[sampleid]][[x]][paste(sampleid, "REL_Myeloid", sep = "_"), paste(sampleid, "REL_Lymphoid", sep = "_")]
  )
  
  set.seed(2025)
  
  # permutation test
  message(" 12")
  pair_12 <- unpaired_perm_test_viz(c(vals_RelLym_DxMye), c(vals_RelMye_DxMye))
  pair_12_p <- pair_12$p_value
  pair_12_c <- pair_12$obs - median(pair_12$perm_distribution)
  
  message(" 13")
  pair_13 <- unpaired_perm_test_viz(c(vals_RelLym_DxMye), c(vals_DxLym_RelMye))
  pair_13_p <- pair_13$p_value
  pair_13_c <- pair_13$obs - median(pair_13$perm_distribution)
  
  message(" 14")
  pair_14 <- unpaired_perm_test_viz(c(vals_RelLym_DxMye), c(vals_DxLym_RelLym))
  pair_14_p <- pair_14$p_value
  pair_14_c <- pair_14$obs - median(pair_14$perm_distribution)
  
  message(" 15")
  pair_15 <- unpaired_perm_test_viz(c(vals_RelLym_DxMye), c(vals_DxMye_DxLym))
  pair_15_p <- pair_15$p_value
  pair_15_c <- pair_15$obs - median(pair_15$perm_distribution)
  
  message(" 16")
  pair_16 <- unpaired_perm_test_viz(c(vals_RelLym_DxMye), c(vals_DxMye_DxLym))
  pair_16_p <- pair_16$p_value
  pair_16_c <- pair_16$obs - median(pair_16$perm_distribution)
  
  result <- list(
    pair_12_p = pair_12_p,
    pair_12_c = pair_12_c,
    pair_13_p = pair_13_p,
    pair_13_c = pair_13_c,
    pair_14_p = pair_14_p,
    pair_14_c = pair_14_c,
    pair_15_p = pair_15_p,
    pair_15_c = pair_15_c,
    pair_16_p = pair_16_p,
    pair_16_c = pair_16_c
  )
  
  return(result)
}

selected_samples_id <- c("AML2", "AML4", "AML5", "AML6", "AML8", "AML10", "AML11", "AML12",
                         "AML13", "AML15", "AML16", "AML17", "AML18", "AML19", "AML21",
                         "AML22", "AML24", "AML25", "AML26", "AML27")
# RNA
data_scRNA_sigSimilarity <- as.data.frame(matrix(data = 0, nrow = length(selected_samples_id), ncol = 10))
rownames(data_scRNA_sigSimilarity) <- selected_samples_id
colnames(data_scRNA_sigSimilarity) <- paste(c("pair_12_p", "pair_12_c", "pair_13_p", "pair_13_c",
                                        "pair_14_p", "pair_14_c", "pair_15_p", "pair_15_c",
                                        "pair_16_p", "pair_16_c"), "RNA", sep = "_")

# ATA
data_scATA_sigSimilarity <- as.data.frame(matrix(data = 0, nrow = length(selected_samples_id), ncol = 10))
rownames(data_scATA_sigSimilarity) <- selected_samples_id
colnames(data_scATA_sigSimilarity) <- paste(c("pair_12_p", "pair_12_c", "pair_13_p", "pair_13_c",
                                        "pair_14_p", "pair_14_c", "pair_15_p", "pair_15_c",
                                        "pair_16_p", "pair_16_c"), "ATA", sep = "_")

for (sampleid in selected_samples_id){
  message(paste0(sampleid, "-scRNA"))
  r_scRNA <- cal_sigValSimilarity(mut_cellTypeAnn_similarity_list_RNA_main, iteration_times, sampleid)
  message(paste0(sampleid, "-scATA"))
  r_scATA <- cal_sigValSimilarity(mut_cellTypeAnn_similarity_list_ATA_main, iteration_times, sampleid)
  data_scRNA_sigSimilarity[sampleid, "pair_12_p_RNA"] <- r_scRNA$pair_12_p
  data_scRNA_sigSimilarity[sampleid, "pair_12_c_RNA"] <- r_scRNA$pair_12_c
  data_scRNA_sigSimilarity[sampleid, "pair_13_p_RNA"] <- r_scRNA$pair_13_p
  data_scRNA_sigSimilarity[sampleid, "pair_13_c_RNA"] <- r_scRNA$pair_13_c
  data_scRNA_sigSimilarity[sampleid, "pair_14_p_RNA"] <- r_scRNA$pair_14_p
  data_scRNA_sigSimilarity[sampleid, "pair_14_c_RNA"] <- r_scRNA$pair_14_c
  data_scRNA_sigSimilarity[sampleid, "pair_15_p_RNA"] <- r_scRNA$pair_15_p
  data_scRNA_sigSimilarity[sampleid, "pair_15_c_RNA"] <- r_scRNA$pair_15_c
  data_scRNA_sigSimilarity[sampleid, "pair_16_p_RNA"] <- r_scRNA$pair_16_p
  data_scRNA_sigSimilarity[sampleid, "pair_16_c_RNA"] <- r_scRNA$pair_16_c
  
  
  data_scATA_sigSimilarity[sampleid, "pair_12_p_ATA"] <- r_scATA$pair_12_p
  data_scATA_sigSimilarity[sampleid, "pair_12_c_ATA"] <- r_scATA$pair_12_c
  data_scATA_sigSimilarity[sampleid, "pair_13_p_ATA"] <- r_scATA$pair_13_p
  data_scATA_sigSimilarity[sampleid, "pair_13_c_ATA"] <- r_scATA$pair_13_c
  data_scATA_sigSimilarity[sampleid, "pair_14_p_ATA"] <- r_scATA$pair_14_p
  data_scATA_sigSimilarity[sampleid, "pair_14_c_ATA"] <- r_scATA$pair_14_c
  data_scATA_sigSimilarity[sampleid, "pair_15_p_ATA"] <- r_scATA$pair_15_p
  data_scATA_sigSimilarity[sampleid, "pair_15_c_ATA"] <- r_scATA$pair_15_c
  data_scATA_sigSimilarity[sampleid, "pair_16_p_ATA"] <- r_scATA$pair_16_p
  data_scATA_sigSimilarity[sampleid, "pair_16_c_ATA"] <- r_scATA$pair_16_c
}

df_RNA <- data_scRNA_sigSimilarity
pairs_RNA <- unique(sub("_p_RNA", "", grep("_p_RNA$", colnames(df_RNA), value = TRUE)))
for (p in pairs_RNA) {
  p_col <- paste0(p, "_p_RNA")
  c_col <- paste0(p, "_c_RNA")
  df_RNA[[c_col]][df_RNA[[p_col]] > 0.001] <- NA
}

df_ATA <- data_scATA_sigSimilarity
pairs_ATA <- unique(sub("_p_ATA", "", grep("_p_ATA$", colnames(df_ATA), value = TRUE)))
for (p in pairs_ATA) {
  p_col <- paste0(p, "_p_ATA")
  c_col <- paste0(p, "_c_ATA")
  df_ATA[[c_col]][df_ATA[[p_col]] > 0.001] <- NA
}


data_scRNA_scATA_sigSimilarity <- cbind(df_RNA[, c("pair_12_c_RNA", "pair_13_c_RNA", "pair_14_c_RNA", "pair_15_c_RNA", "pair_16_c_RNA")],
                                        df_ATA[, c("pair_12_c_ATA", "pair_13_c_ATA", "pair_14_c_ATA", "pair_15_c_ATA", "pair_16_c_ATA")])

#max_abs <- max(abs(data_scRNA_scATA_sigSimilarity), na.rm = TRUE)
#breaks <- seq(-max_abs, max_abs, length.out = 101)

#colors <- rev(colorRampPalette(brewer.pal(11, "BrBG"))(100))

col_fun <- colorRamp2(
  c(-1, 0, 1),
  c("#762a83", "white", "#8c510a")
)

mat <- data_scRNA_scATA_sigSimilarity

group <- ifelse(rownames(mat) %in% c("AML4", "AML16", "AML18"),
                "AML4/16/18", "Others")
group <- factor(group, levels = c("Others", "AML4/16/18"))

Heatmap(
  mat,
  name = "effect size(obs-median)",
  row_split = group,
  cluster_rows = TRUE,
  clustering_method_rows = "ward.D",
  cluster_columns = FALSE,
  col = col_fun,
  show_row_names = TRUE,
  row_gap = unit(4, "mm"),
  na_col = "grey90"
)

Heatmap(
  mat,
  name = "effect size(obs-median)",
  row_split = group,
  cluster_rows = TRUE,
  clustering_method_rows = "ward.D",
  cluster_columns = FALSE,
  col = col_fun,
  show_row_names = TRUE,
  row_gap = unit(4, "mm"),
  na_col = "grey90",
  cell_fun = function(j, i, x, y, w, h, fill) {
    if (is.na(mat[i, j])) {
      grid::grid.text(
        "×", x = x, y = y,
        gp = grid::gpar(fontsize = 10, col = "black")
      )
    }
  }
)

ht <- Heatmap(
  mat,
  name = "effect size(obs-median)",
  row_split = group,
  cluster_rows = TRUE,
  clustering_method_rows = "ward.D",
  cluster_columns = FALSE,
  col = col_fun,
  show_row_names = TRUE,
  row_gap = unit(4, "mm"),
  na_col = "grey90"
)

pdf("heatmap_effect_size.pdf", width = 5.8, height = 6)

draw(ht)

dev.off()

sample_selected <- "AML18"
# 1
vals_RelLym_DxMye_RNA <- sapply(
  paste0("iter_", iteration_times),
  function(x) mut_cellTypeAnn_similarity_list_RNA_main[[sample_selected]][[x]][paste(sample_selected, "REL_Lymphoid", sep = "_"), paste(sample_selected, "DX_Myeloid", sep = "_")]
)

vals_RelLym_DxMye_ATA <- sapply(
  paste0("iter_", iteration_times),
  function(x) mut_cellTypeAnn_similarity_list_ATA_main[[sample_selected]][[x]][paste(sample_selected, "REL_Lymphoid", sep = "_"), paste(sample_selected, "DX_Myeloid", sep = "_")]
)

# 2
vals_RelLym_DxLym_RNA <- sapply(
  paste0("iter_", iteration_times),
  function(x) mut_cellTypeAnn_similarity_list_RNA_main[[sample_selected]][[x]][paste(sample_selected, "REL_Lymphoid", sep = "_"), paste(sample_selected, "DX_Lymphoid", sep = "_")]
)

vals_RelLym_DxLym_ATA <- sapply(
  paste0("iter_", iteration_times),
  function(x) mut_cellTypeAnn_similarity_list_ATA_main[[sample_selected]][[x]][paste(sample_selected, "REL_Lymphoid", sep = "_"), paste(sample_selected, "DX_Lymphoid", sep = "_")]
)
# 3
vals_DxMye_DxLym_RNA<- sapply(
  paste0("iter_", iteration_times),
  function(x) mut_cellTypeAnn_similarity_list_RNA_main[[sample_selected]][[x]][paste(sample_selected, "DX_Myeloid", sep = "_"), paste(sample_selected, "DX_Lymphoid", sep = "_")]
)

vals_DxMye_DxLym_ATA<- sapply(
  paste0("iter_", iteration_times),
  function(x) mut_cellTypeAnn_similarity_list_ATA_main[[sample_selected]][[x]][paste(sample_selected, "DX_Myeloid", sep = "_"), paste(sample_selected, "DX_Lymphoid", sep = "_")]
)
# 4
vals_RelMye_RelLym_RNA <- sapply(
    paste0("iter_", iteration_times),
    function(x) mut_cellTypeAnn_similarity_list_RNA_main[[sample_selected]][[x]][paste(sample_selected, "REL_Myeloid", sep = "_"), paste(sample_selected, "REL_Lymphoid", sep = "_")]
  )
vals_RelMye_RelLym_ATA <- sapply(
    paste0("iter_", iteration_times),
    function(x) mut_cellTypeAnn_similarity_list_ATA_main[[sample_selected]][[x]][paste(sample_selected, "REL_Myeloid", sep = "_"), paste(sample_selected, "REL_Lymphoid", sep = "_")]
  )

set.seed(2025)
res_RNA12 <- unpaired_perm_test_viz(vals_RelLym_DxMye_RNA, vals_RelLym_DxLym_RNA)
res_RNA13 <- unpaired_perm_test_viz(vals_RelLym_DxMye_RNA, vals_DxMye_DxLym_RNA) 
res_RNA14 <- unpaired_perm_test_viz(vals_RelLym_DxMye_RNA, vals_RelMye_RelLym_RNA)  
  
res_ATA12 <- unpaired_perm_test_viz(vals_RelLym_DxMye_ATA, vals_RelLym_DxLym_ATA)
res_ATA13 <- unpaired_perm_test_viz(vals_RelLym_DxMye_ATA, vals_DxMye_DxLym_ATA) 
res_ATA14 <- unpaired_perm_test_viz(vals_RelLym_DxMye_ATA, vals_RelMye_RelLym_ATA)     


p_res_RNA12 <- plot_perm_test(res_RNA12)
p_res_ATA12 <- plot_perm_test(res_ATA12)
p_res_RNA13 <- plot_perm_test(res_RNA13)
p_res_ATA13 <- plot_perm_test(res_ATA13)
p_res_RNA14 <- plot_perm_test(res_RNA14)
p_res_ATA14 <- plot_perm_test(res_ATA14)

p_res_RNA12
p_res_ATA12

p_res_RNA13
p_res_ATA13

p_res_RNA14
p_res_ATA14

data_value_similarity_RNA <- list(DXMy_RELLy = vals_RelLym_DxMye_RNA,
                                  DXLy_RELLy = vals_RelLym_DxLym_RNA,
                                  DXMy_DXLy = vals_DxMye_DxLym_RNA,
                                  RELMy_RELLy = vals_RelMye_RelLym_RNA)

data_value_similarity_ATA <- list(DXMy_RELLy = vals_RelLym_DxMye_ATA,
                                  DXLy_RELLy = vals_RelLym_DxLym_ATA,
                                  DXMy_DXLy = vals_DxMye_DxLym_ATA,
                                  RELMy_RELLy = vals_RelMye_RelLym_ATA)

comparisons <- list(
  c("DXMy_RELLy", "DXLy_RELLy"),
  c("DXMy_RELLy", "DXMy_DXLy"),
  c("DXMy_RELLy", "RELMy_RELLy")
)

p_boxPlotdata_value_similarity_RNA <- ggplot(stack(data_value_similarity_RNA), aes(ind, values, fill = ind)) +
  geom_boxplot() +
  stat_compare_means(comparisons = comparisons, method = "wilcox.test") +
  theme_classic()

p_boxPlotdata_value_similarity_ATA <- ggplot(stack(data_value_similarity_ATA), aes(ind, values, fill = ind)) +
  geom_boxplot() +
  stat_compare_means(comparisons = comparisons, method = "wilcox.test") +
  theme_classic()

p_boxPlotdata_value_similarity_RNA
p_boxPlotdata_value_similarity_ATA

#ggsave("../p_boxPlotdata_value_similarity_RNA.pdf", p_boxPlotdata_value_similarity_RNA, width = 3.2, height = 1.9)
#ggsave("../p_boxPlotdata_value_similarity_ATA.pdf", p_boxPlotdata_value_similarity_ATA, width = 3.2, height = 1.9)
```

# mutations shared by different lineage
```{r sharedMutations}
cellxmut_to_01 <- function(cellxmut,
                           mapping = c(NoCall = 0L, WT = 0L, Mut = 1L),
                           default_value = NA_integer_,
                           return_matrix = TRUE) {
  if (!(is.matrix(cellxmut) || is.data.frame(cellxmut))) {
    stop("cellxmut must be a matrix or data.frame.")
  }

  rn <- rownames(cellxmut)
  cn <- colnames(cellxmut)
  if (is.null(rn) || any(rn == "")) stop("cellxmut must have non-empty rownames (cell names).")
  if (is.null(cn) || any(cn == "")) stop("cellxmut must have non-empty colnames (mutation sites).")

  # Work in character space, trim whitespace
  x <- as.matrix(cellxmut)
  x <- apply(x, c(1, 2), function(v) trimws(as.character(v)))

  # If already numeric-ish (e.g., "0"/"1"), keep them
  suppressWarnings(num <- matrix(as.integer(x), nrow = nrow(x), ncol = ncol(x)))
  is_num <- !is.na(num) & x %in% c("0", "1")
  out <- matrix(NA_integer_, nrow = nrow(x), ncol = ncol(x), dimnames = list(rn, cn))
  out[is_num] <- num[is_num]

  # Map remaining via mapping
  map_vec <- as.integer(mapping); names(map_vec) <- names(mapping)
  idx <- is.na(out)
  if (any(idx)) {
    mapped <- map_vec[x[idx]]
    mapped[is.na(mapped)] <- default_value
    out[idx] <- as.integer(mapped)
  }

  if (return_matrix) return(out)
  as.data.frame(out, check.names = FALSE)
}

celltype_matrix <- function(mat, meta, method = "sum") {
  
  # 1. clean cell names
  cells_format_mat <- sub("_L[0-9]+_", "_", rownames(mat))
  
  # 2.  mapping
  meta_map <- setNames(meta$maincelltype, meta$CellTag)
  
  # 3.  cells
  keep_idx <- cells_format_mat %in% meta$CellTag
  
  mat <- mat[keep_idx, , drop = FALSE]
  cells_use <- cells_format_mat[keep_idx]
  
  celltype <- meta_map[cells_use]
  
  # 4. attach celltype
  df <- as.data.frame(mat)
  df$maincelltype <- celltype
  
  # 5. aggregation
  if (method == "sum") {
    agg <- aggregate(. ~ maincelltype, data = df, FUN = sum)
  } else if (method == "mean") {
    agg <- aggregate(. ~ maincelltype, data = df, FUN = mean)
  } else if (method == "binary") {
    agg <- aggregate(. ~ maincelltype, data = df, FUN = function(x) as.numeric(any(x > 0)))
  } else {
    stop("method must be 'sum', 'mean' or 'binary'")
  }
  
  rownames(agg) <- agg$maincelltype
  agg$maincelltype <- NULL
  
  return(as.matrix(agg))
}

# scRNA
celltypes <- c("HSPC", "Myeloid", "Lymphoid", "Erythroid")
samples_scRNA_csm <- intersect(names(AML_scRNA_CSM_list), unique(AML_scRNA_meta$Lambo_et_al_ID))

mutation_ct_scRNA <- vector("list", length(samples_scRNA_csm))
names(mutation_ct_scRNA) <- samples_scRNA_csm
for (sample in samples_scRNA_csm){
  message(sample)
  csm_01 <- cellxmut_to_01(AML_scRNA_CSM_list[[sample]])
  csm_01_ct <- celltype_matrix(mat = csm_01, meta = AML_scRNA_meta[AML_scRNA_meta$Lambo_et_al_ID == sample,])
  csm_01_ct <- csm_01_ct[celltypes,]
  csm_01_ct_bin <- ifelse(csm_01_ct >= 3, 1, 0)
  df <- as.data.frame(t(csm_01_ct_bin)) 
  df <- df %>% rownames_to_column("mutation")
  df[celltypes] <- lapply(df[celltypes], function(x) as.logical(as.integer(x)))
  mutation_ct_scRNA[[sample]] <- df
}

# scATA
celltypes <- c("HSPC", "Myeloid", "Lymphoid", "Erythroid")
samples_scATA_csm <- intersect(names(AML_scATA_CSM_list), unique(AML_scATA_meta$SampleID))
mutation_ct_scATA <- vector("list", length(samples_scATA_csm))
names(mutation_ct_scATA) <- samples_scATA_csm
for (sample in samples_scATA_csm){
  message(sample)
  csm_01 <- cellxmut_to_01(AML_scATA_CSM_list[[sample]])
  csm_01_ct <- celltype_matrix(mat = csm_01, meta = AML_scATA_meta[AML_scATA_meta$SampleID == sample,])
  csm_01_ct <- csm_01_ct[celltypes,]
  csm_01_ct_bin <- ifelse(csm_01_ct >= 3, 1, 0)
  df <- as.data.frame(t(csm_01_ct_bin)) 
  df <- df %>% rownames_to_column("mutation")
  df[celltypes] <- lapply(df[celltypes], function(x) as.logical(as.integer(x)))
  mutation_ct_scATA[[sample]] <- df
}

all_mutation_ct_scRNA <- do.call(rbind, mutation_ct_scRNA)
all_mutation_ct_scATA <- do.call(rbind, mutation_ct_scATA)


ComplexUpset::upset(
  all_mutation_ct_scRNA,
  intersect = celltypes,
  base_annotations = list(
    'Intersection size' = intersection_size()
  )
)

ComplexUpset::upset(
  all_mutation_ct_scATA,
  intersect = celltypes,
  base_annotations = list(
    'Intersection size' = intersection_size()
  )
)

p_all_mutation_ct_scRNA <- ComplexUpset::upset(
  all_mutation_ct_scRNA,
  intersect = celltypes,
  base_annotations = list(
    'Intersection size' = intersection_size()
  )
)

p_all_mutation_ct_scATA <- ComplexUpset::upset(
  all_mutation_ct_scATA,
  intersect = celltypes,
  base_annotations = list(
    'Intersection size' = intersection_size()
  )
)

#ggsave("../p_all_mutation_ct_scRNA.pdf", p_all_mutation_ct_scRNA, width = 4.7, height = 2.9)
#ggsave("../p_all_mutation_ct_scATA.pdf", p_all_mutation_ct_scATA, width = 4.7, height = 2.9)
```

# SNV tree cell type
```{r celltypeTree}
merge_cell_mut_matrix <- function(mat1, mat2, fill = 0) {
  all_cells <- union(rownames(mat1), rownames(mat2))
  all_muts  <- union(colnames(mat1), colnames(mat2))
  
  merged_mat <- matrix(fill,
                       nrow = length(all_cells),
                       ncol = length(all_muts),
                       dimnames = list(all_cells, all_muts))
  
  merged_mat[rownames(mat1), colnames(mat1)] <- mat1
  
  merged_mat[rownames(mat2), colnames(mat2)] <- 
    merged_mat[rownames(mat2), colnames(mat2)] + mat2
  
  return(merged_mat)
}

cellxmut_to_01 <- function(cellxmut,
                           mapping = c(NoCall = 0L, WT = 0L, Mut = 1L),
                           default_value = NA_integer_,
                           return_matrix = TRUE) {
  if (!(is.matrix(cellxmut) || is.data.frame(cellxmut))) {
    stop("cellxmut must be a matrix or data.frame.")
  }

  rn <- rownames(cellxmut)
  cn <- colnames(cellxmut)
  if (is.null(rn) || any(rn == "")) stop("cellxmut must have non-empty rownames (cell names).")
  if (is.null(cn) || any(cn == "")) stop("cellxmut must have non-empty colnames (mutation sites).")

  # Work in character space, trim whitespace
  x <- as.matrix(cellxmut)
  x <- apply(x, c(1, 2), function(v) trimws(as.character(v)))

  # If already numeric-ish (e.g., "0"/"1"), keep them
  suppressWarnings(num <- matrix(as.integer(x), nrow = nrow(x), ncol = ncol(x)))
  is_num <- !is.na(num) & x %in% c("0", "1")
  out <- matrix(NA_integer_, nrow = nrow(x), ncol = ncol(x), dimnames = list(rn, cn))
  out[is_num] <- num[is_num]

  # Map remaining via mapping
  map_vec <- as.integer(mapping); names(map_vec) <- names(mapping)
  idx <- is.na(out)
  if (any(idx)) {
    mapped <- map_vec[x[idx]]
    mapped[is.na(mapped)] <- default_value
    out[idx] <- as.integer(mapped)
  }

  if (return_matrix) return(out)
  as.data.frame(out, check.names = FALSE)
}

celltype_matrix <- function(mat, meta, method = "sum") {
  common_cells <- intersect(rownames(mat), meta$CellTag)
  mat <- mat[common_cells, , drop = FALSE]
  meta_sub <- meta[match(common_cells, meta$CellTag), ]
  
  df <- as.data.frame(mat)
  df$Classified_CelltypeQ <- meta_sub$Classified_CelltypeQ
  
  if (method == "sum") {
    agg <- aggregate(. ~ Classified_CelltypeQ, data = df, FUN = sum)
  } else if (method == "mean") {
    agg <- aggregate(. ~ Classified_CelltypeQ, data = df, FUN = mean)
  } else if (method == "binary") {
    agg <- aggregate(. ~ Classified_CelltypeQ, data = df, FUN = function(x) as.numeric(sum(x) > 0))
  } else {
    stop("method must be 'sum', 'mean' or 'binary'")
  }
  
  rownames(agg) <- agg$Classified_CelltypeQ
  agg$Classified_CelltypeQ <- NULL
  
  return(as.matrix(agg))
}

celltype_matrix_state <- function(mat, meta) {

  common_cells <- intersect(rownames(mat), meta$CellTag)
  mat <- mat[common_cells, , drop = FALSE]
  ct <- meta$Classified_CelltypeQ[match(common_cells, meta$CellTag)]

  idx_list <- split(seq_along(ct), ct)

  res <- lapply(idx_list, function(idx) {

    sub_mat <- mat[idx, , drop = FALSE]

    apply(sub_mat, 2, function(x) {

      if (any(x == 1, na.rm = TRUE)) {
        return(1L)
      } else if (any(x == 0, na.rm = TRUE)) {
        return(0L)
      } else {
        return(3L)
      }

    })

  })

  res_mat <- do.call(rbind, res)

  return(res_mat)
}

AML04_CSM_scRNA_DX <- cellxmut_to_01(cellxmut = AML_scRNA_CSM_DX_REL_select_list$AML4_DX, mapping = c(NoCall = 0L, WT = 0L, Mut = 1L), default_value = 0L)
AML04_CSM_scRNA_REL <- cellxmut_to_01(cellxmut = AML_scRNA_CSM_DX_REL_select_list$AML4_REL, mapping = c(NoCall = 0L, WT = 0L, Mut = 1L), default_value = 0L)

AML04_CSM_scATA_DX <- cellxmut_to_01(cellxmut = AML_scATA_CSM_DX_REL_select_list$AML4_DX, mapping = c(NoCall = 0L, WT = 0L, Mut = 1L), default_value = 0L)
AML04_CSM_scATA_REL <- cellxmut_to_01(cellxmut = AML_scATA_CSM_DX_REL_select_list$AML4_REL, mapping = c(NoCall = 0L, WT = 0L, Mut = 1L), default_value = 0L)

AML16_CSM_scATA_DX <- cellxmut_to_01(AML_scATA_CSM_DX_REL_select_list$AML16_DX, default_value = 0L)
AML16_CSM_scATA_REL <- cellxmut_to_01(AML_scATA_CSM_DX_REL_select_list$AML16_REL, default_value = 0L)

AML18_CSM_scATA_DX <- cellxmut_to_01(AML_scATA_CSM_DX_REL_select_list$AML18_DX, default_value = 0L)
AML18_CSM_scATA_REL <- cellxmut_to_01(AML_scATA_CSM_DX_REL_select_list$AML18_REL, default_value = 0L)

AML04_CSM_scRNA <- merge_cell_mut_matrix(AML04_CSM_scRNA_DX,
                                         AML04_CSM_scRNA_REL)

AML04_CSM_scATA <- merge_cell_mut_matrix(AML04_CSM_scATA_DX,
                                         AML04_CSM_scATA_REL)

AML16_CSM_scATA <- merge_cell_mut_matrix(AML16_CSM_scATA_DX,
                                         AML16_CSM_scATA_REL)

AML18_CSM_scATA <- merge_cell_mut_matrix(AML18_CSM_scATA_DX,
                                         AML18_CSM_scATA_REL)

AML04_CSM_scRNA_ct <- celltype_matrix(mat = AML04_CSM_scRNA, meta = AML_scRNA_meta_DX_REL)
AML04_CSM_scRNA_ct_state <- celltype_matrix_state(AML04_CSM_scRNA, AML_scRNA_meta_DX_REL)
AML04_CSM_scATA_ct <- celltype_matrix(mat = AML04_CSM_scATA, meta = AML_scATA_meta_DX_REL)
AML04_CSM_scATA_ct_state <- celltype_matrix_state(AML04_CSM_scATA, AML_scATA_meta_DX_REL)
AML16_CSM_scATA_ct <- celltype_matrix(mat = AML16_CSM_scATA, meta = AML_scATA_meta_DX_REL)
AML18_CSM_scATA_ct <- celltype_matrix(mat = AML18_CSM_scATA, meta = AML_scATA_meta_DX_REL)

#write.table(x = t(AML04_CSM_scRNA_ct_state), "/Users/liqun/Downloads/AML04_CSM_scRNA_ct_state.txt", sep = " ", row.names = T, quote = F)
#write.table(x = t(AML04_CSM_scATA_ct_state), "/Users/liqun/Downloads/AML04_CSM_scATA_ct_state.txt", sep = " ", row.names = T, quote = F)

binary_mat_RNA <- (AML04_CSM_scRNA_ct != 0) * 1
binary_mat_ATA <- (AML04_CSM_scATA_ct != 0) * 1
library(vegan)
library(ape)
dist_mat_RNA <- vegdist(binary_mat_RNA, method = "jaccard")
dist_mat_ATA <- vegdist(binary_mat_ATA, method = "jaccard")
hc_RNA <- hclust(dist_mat_RNA, method = "average")  # average = UPGMA
hc_ATA <- hclust(dist_mat_ATA, method = "average") 
plot(hc_RNA, main = "Celltype phylogenetic tree (UPGMA) RNA")
plot(hc_ATA, main = "Celltype phylogenetic tree (UPGMA) ATA")
nj_tree_RNA <- nj(dist_mat_RNA)
nj_tree_ATA <- nj(dist_mat_ATA)
plot(nj_tree_RNA, main = "Celltype phylogenetic tree (NJ) RNA")
plot(nj_tree_ATA, main = "Celltype phylogenetic tree (NJ) ATA")
```

# mutratio
```{r mutratio}
ct_counts_scRNA <- table(
  AML_scRNA_meta_DX_REL$Classified_CelltypeQ[
    AML_scRNA_meta_DX_REL$SampleID == "AML4"
  ]
)

ct_counts_scATA <- table(
  AML_scATA_meta_DX_REL$Classified_CelltypeQ[
    AML_scATA_meta_DX_REL$SampleID == "AML4"
  ]
)
# 2. mutation counts rowSums
mut_counts_scRNA <- rowSums(AML04_CSM_scRNA_ct)
mut_counts_scATA <- rowSums(AML04_CSM_scATA_ct)

# 3. map
common_ct_scRNA <- intersect(names(ct_counts_scRNA), names(mut_counts_scRNA))
ct_counts_scRNA <- ct_counts_scRNA[common_ct_scRNA]
mut_counts_scRNA <- mut_counts_scRNA[common_ct_scRNA]
common_ct_scATA <- intersect(names(ct_counts_scATA), names(mut_counts_scATA))
ct_counts_scATA <- ct_counts_scATA[common_ct_scATA]
mut_counts_scATA <- mut_counts_scATA[common_ct_scATA]

# 4. ratio
ratio_scRNA <- mut_counts_scRNA / ct_counts_scRNA
ratio_scATA <- mut_counts_scATA / ct_counts_scATA

# 5. sort
ratio_sorted_scRNA <- sort(ratio_scRNA)
ratio_sorted_scATA <- sort(ratio_scATA)

ratio_sorted_scRNA
ratio_sorted_scATA

common_ct <- intersect(names(ratio_scRNA), names(ratio_scATA))

ratio_scRNA2 <- ratio_scRNA[common_ct]
ratio_scATA2 <- ratio_scATA[common_ct]
cor(ratio_scRNA2, ratio_scATA2, method = "spearman")

df_cor <- data.frame(
  celltype = common_ct,
  scRNA = ratio_scRNA2,
  scATA = ratio_scATA2
)

ggplot(df_cor, aes(scRNA.Freq, scATA.Freq, label = celltype)) +
  geom_point(size = 3) +
  geom_text(size = 3, vjust = -0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_classic()
```

# sessionInfo
```{r sessionInfo}
sessionInfo()
```



