---
title: "Figure 4 pAML reproducibility analysis"
author: "Qun Li"
date: "22 Aug 2025 at 10:12:37"
output:
  pdf_document:
    highlight: tango
    toc: true
    toc_depth: 3
    number_sections: true
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  word_document:
    toc: true
    toc_depth: '3'
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  results = "hide",
  fig.width = 8,
  fig.height = 6,
  out.width = "100%"
)

options(width = 80)
```

# Env
```{r envs}
rm(list = ls())
setwd("/Users/liqun/Desktop/scMutrace-tutorial/ReproducibilityAnalysis/Figure4/")
```

# Load libraries
```{r loadLibs}
library(FSA)
library(ineq)
library(grid)
library(vegan)
library(ArchR)
library(dplyr)
library(GO.db)
library(tidyr)
library(ggpubr)
library(Seurat)
library(ggpmisc)
library(stringr)
library(ggplot2)
library(circlize)
library(pheatmap)
library(corrplot)
library(bedtoolsr)
library(tidyverse)
library(patchwork)
library(gridExtra)
library(data.table)
library(UpSetR)
library(ggsankey)
library(RColorBrewer)
library(org.Hs.eg.db)
library(GenomicRanges)
library(ComplexHeatmap)
library(clusterProfiler)
library(networkD3)
library(htmlwidgets)
```

# define functions
```{r defineFunctions}
cosine_similarity <- function(a, b, na.rm = TRUE) {
  a <- as.numeric(a)
  b <- as.numeric(b)
  
  if (na.rm) {
    keep <- !(is.na(a) | is.na(b))
    a <- a[keep]
    b <- b[keep]
  }
  
  denom <- sqrt(sum(a^2)) * sqrt(sum(b^2))
  if (denom == 0) return(as.numeric(NA)) 
  
  sim <- sum(a * b) / denom
  
  sim <- max(min(sim, 1), -1)
  
  return(as.numeric(sim))
}

retriveCellsByLocation <- function(CSM, locList){
  CSM$CellTag <- rownames(CSM)
  CSM_tmp <- CSM[, c("CellTag",locList)]
  data_MutPerCell_CSM_tmp <- data.frame(
    CellTag = rownames(CSM_tmp),
    MutNum = rowSums(CSM_tmp == "Mut"),
    WTNum = rowSums(CSM_tmp == "WT")
  )
  cells_Total <- data_MutPerCell_CSM_tmp[rowSums(data_MutPerCell_CSM_tmp[,c("MutNum", "WTNum")]) > 0, ]$CellTag
  cells_WT <-    data_MutPerCell_CSM_tmp[data_MutPerCell_CSM_tmp[,"WTNum"] > 0,]$CellTag
  cells_Mut <-   data_MutPerCell_CSM_tmp[data_MutPerCell_CSM_tmp[,"MutNum"] > 0,]$CellTag
  cells <- list(Total = cells_Total, Mut = cells_Mut, WT = cells_WT)
  return(cells)
}

retriveLocsByCells <- function(CSM, cellList){
  CSM$CellTag <- rownames(CSM)
  CSM_tmp <- CSM[CSM$CellTag %in% cellList,]
  data_CellPerMut_CSM_tmp <- data.frame(
    Location = colnames(CSM_tmp),
    CellMutNum = colSums(CSM_tmp == "Mut"),
    CellWTNum = colSums(CSM_tmp == "WT")
  )
  pos_Total <- data_CellPerMut_CSM_tmp[rowSums(data_CellPerMut_CSM_tmp[, c("CellMutNum", "CellWTNum")]) > 0,]$Location
  pos_Mut <- data_CellPerMut_CSM_tmp[data_CellPerMut_CSM_tmp[, "CellMutNum"] > 0 ,]$Location
  pos_WT <- data_CellPerMut_CSM_tmp[data_CellPerMut_CSM_tmp[, "CellWTNum"] > 0 ,]$Location
  pos <- list(Total = pos_Total, Mut = pos_Mut, WT = pos_WT)
  return(pos)
}

get_mut_cols <- function(df, loc) {
  row <- df[df$Location == loc, ]       
  mut_cols <- names(row)[row == "Mut"] 
  mut_cols[mut_cols != "Location"]     
}
```

*AML_census.rds* and *AML_tiergenes.txt* were downloaded from the COSMIC database and curated for downstream analyses

*AML_SSM.rds* and *AML_ATA_CSM_list.rds*, *AML_RNA_CSM_list.rds* were generated using scMutrace as described in the Methods section and `0_pipelineFunctions.Rmd`(pipelines for generating single-cell somatic mutation data list).

*AML_scRNA_meta.rds* and *AML_scATA_meta.rds* contain meta data for cell type annotation similarity analysis and were generate from `0_pipelineFunctions.Rmd`(pipelines for porcessing scRNA-seq data and pipelines for processing scATAC-seq data).

*sampleInfo.txt* contains sample information including cell numbers and patient clinical data for scRNA-seq and scATAC-seq data and was downloaded from original paper.

*EGA_scATA.hg38_multianno.txt* and *EGA_scRNA.hg38_multianno.txt* contain mutation annotation information generated using ANNOVAR.

*Mutational_Matrix_SBS.txt* and *COSMIC_SBS96_Activities.txt* were generated using SigProfiler suite as described in the Methods section.


# load data
```{r loadData}
# load AML census data
AML_census <- readRDS("../Data/HumanpAML/AML_census.rds")
AML_genes_tiers <- read.table("../Data/HumanpAML/AML_tiergenes.txt", header = T, sep = "\t")

# load SSM data
AML_SSM <- readRDS("../Data/HumanpAML/AML_SSM.rds")
AML_SSM_scRNA <- AML_SSM$scRNA
AML_SSM_scATA <- AML_SSM$scATA

# CSM
AML_ATA_CSM_list <- readRDS("../Data/HumanpAML/AML_ATA_CSM_list.rds")
AML_RNA_CSM_list <- readRDS("../Data/HumanpAML/AML_RNA_CSM_list.rds")

# meta data for celltype annotation similarity
AML_scRNA_meta_list <- readRDS("../Data/HumanpAML/AML_scRNA_meta.rds")
# metaQun
AML_scATA_meta_list <- readRDS("../Data/HumanpAML/AML_scATA_meta.rds")

AML_scRNA_meta <- do.call(rbind, AML_scRNA_meta_list)
rownames(AML_scRNA_meta) <- sub("\\.", "#", rownames(AML_scRNA_meta))

AML_scATA_meta <- do.call(rbind, AML_scATA_meta_list)
AML_scATA_meta$SampleID <- sub("_.*", "", AML_scATA_meta$Sample)

rownames(AML_scATA_meta) <- sub("^AML[0-9]+\\.", "", rownames(AML_scATA_meta))

cellTypes <- unique(union(AML_scRNA_meta$Classified_Celltype, AML_scATA_meta$predictedGroup_Un))

# extract mutations and get bed format with census tag
AML_SSM_scRNA_bed <- separate(AML_SSM_scRNA, Location, into = c("chr", "pos", "ref", "alt"), sep = "_", remove = FALSE)[,1:5]
AML_SSM_scRNA_bed$end <- AML_SSM_scRNA_bed$pos
AML_SSM_scRNA_bed <- AML_SSM_scRNA_bed[,c("chr", "pos", "end", "ref", "alt", "Location")]
AML_SSM_scRNA_bed$Tag <- ifelse(AML_SSM_scRNA_bed$Location %in% AML_census$mutation$Location, "Census", "NotCensus")
AML_scRNA_NotCensus_locs <- AML_SSM_scRNA_bed %>% dplyr::filter(Tag == "NotCensus") %>% pull(Location)

AML_SSM_scATA_bed <- separate(AML_SSM_scATA, Location, into = c("chr", "pos", "ref", "alt"), sep = "_", remove = FALSE)[,1:5]
AML_SSM_scATA_bed$end <- AML_SSM_scATA_bed$pos
AML_SSM_scATA_bed <- AML_SSM_scATA_bed[,c("chr", "pos", "end", "ref", "alt", "Location")]
AML_SSM_scATA_bed$Tag <- ifelse(AML_SSM_scATA_bed$Location %in% AML_census$mutation$Location, "Census", "NotCensus")
AML_scATA_NotCensus_locs <- AML_SSM_scATA_bed %>% dplyr::filter(Tag == "NotCensus") %>% pull(Location)

# load paired sample info data
AML_sampleInfo <- read.table("../Data/HumanpAML/sampleInfo.txt", header = T, sep = "\t")
AML_sampleInfo[is.na(AML_sampleInfo)] <- 0

# AML ann
AML_scATA_ann <- read.table("../Data/HumanpAML/EGA_scATA.hg38_multianno.txt", header = T, sep = "\t")
AML_scRNA_ann <- read.table("../Data/HumanpAML/EGA_scRNA.hg38_multianno.txt", header = T, sep = "\t")

col_freq_databases <- c("gnomad40_exome_AF", "gnomad40_exome_AF_raw", "gnomad40_exome_AF_XX", 
                        "gnomad40_exome_AF_XY", "gnomad40_exome_AF_grpmax", "gnomad40_exome_faf95", 
                        "gnomad40_exome_faf99", "gnomad40_exome_fafmax_faf95_max", "gnomad40_exome_fafmax_faf99_max", 
                        "gnomad40_exome_AF_afr", "gnomad40_exome_AF_amr", "gnomad40_exome_AF_asj", 
                        "gnomad40_exome_AF_eas", "gnomad40_exome_AF_fin", "gnomad40_exome_AF_mid", 
                        "gnomad40_exome_AF_nfe", "gnomad40_exome_AF_remaining", "gnomad40_exome_AF_sas", 
                        "gnomad40_genome_AF", "gnomad40_genome_AF_raw", "gnomad40_genome_AF_XX", 
                        "gnomad40_genome_AF_XY", "gnomad40_genome_AF_grpmax", "gnomad40_genome_faf95", 
                        "gnomad40_genome_faf99", "gnomad40_genome_fafmax_faf95_max", 
                        "gnomad40_genome_fafmax_faf99_max", "gnomad40_genome_AF_afr",
                        "gnomad40_genome_AF_ami", "gnomad40_genome_AF_amr", 
                        "gnomad40_genome_AF_asj", "gnomad40_genome_AF_eas", 
                        "gnomad40_genome_AF_fin", "gnomad40_genome_AF_mid", 
                        "gnomad40_genome_AF_nfe", "gnomad40_genome_AF_remaining", "gnomad40_genome_AF_sas")

col_freq_databases %in% colnames(AML_scATA_ann)
col_freq_databases %in% colnames(AML_scRNA_ann)

AML_scATA_ann <- AML_scATA_ann %>%
  mutate(across(all_of(col_freq_databases), 
                ~ as.numeric(replace(., . == ".", 0))))

AML_scRNA_ann <- AML_scRNA_ann %>%
  mutate(across(all_of(col_freq_databases), 
                ~ as.numeric(replace(., . == ".", 0))))

# Annotate bed census and nonCensus
# scATA
AML_SSM_scATA_bed <- AML_SSM_scATA_bed %>%
  left_join(
    AML_scATA_ann %>%
      dplyr::select(Location, all_of(col_freq_databases)),
    by = "Location"
  ) %>%
  mutate(
    freq = if_else(
      if_any(all_of(col_freq_databases), ~ . > 0.0001),
      "high", "rare"
    )
  ) %>%
  dplyr::select(names(AML_SSM_scATA_bed), freq)

# scRNA
AML_SSM_scRNA_bed <- AML_SSM_scRNA_bed %>%
  left_join(
    AML_scRNA_ann %>%
      dplyr::select(Location, all_of(col_freq_databases)),
    by = "Location"
  ) %>%
  mutate(
    freq = if_else(
      if_any(all_of(col_freq_databases), ~ . > 0.0001),
      "high", "rare"
    )
  ) %>%
  dplyr::select(names(AML_SSM_scRNA_bed), freq)

# load mutation features data
# generate using sigProfiler suite
AML_mutation_features <- read.table("../Data/HumanpAML/Mutational_Matrix_SBS.txt", header = T, sep = "\t", row.names = 1)
AML_mutation_features_contribution <- read.table("../Data/HumanpAML/COSMIC_SBS96_Activities.txt", header = T, sep = "\t")
```

# extract different stage patient mutations
```{r extractStageMutations}
AML_patient_ATA_mutations <- list()
AML_patient_RNA_mutations <- list()
Stages <- c("DX", "REM", "REL")
for (stage in Stages){
  colnames_stage <- grep(paste0("_", stage, "$"), colnames(AML_SSM_scATA), value = TRUE)
  samples_stage_loc <- AML_SSM_scATA %>%
    dplyr::select(Location, colnames_stage) %>%
    dplyr::filter(if_any(all_of(colnames_stage), ~ . == "Mut")) %>%
    dplyr::pull(Location)
  AML_patient_ATA_mutations[[stage]] <- samples_stage_loc
}

for (stage in Stages){
  colnames_stage <- grep(paste0("_", stage, "$"), colnames(AML_SSM_scRNA), value = TRUE)
  samples_stage_loc <- AML_SSM_scRNA %>%
    dplyr::select(Location, colnames_stage) %>%
    dplyr::filter(if_any(all_of(colnames_stage), ~ . == "Mut")) %>%
    dplyr::pull(Location)
  AML_patient_RNA_mutations[[stage]] <- samples_stage_loc
}

upset(fromList(AML_patient_ATA_mutations),
      nsets = length(AML_patient_ATA_mutations),
      order.by = "freq",
      mainbar.y.label = "Intersection Size ATA",
      mb.ratio = c(0.6, 0.4),
      main.bar.color = "#345170",
      sets.bar.color = "#EBAC8C",
      text.scale = c(2, 2, 2, 1.5, 2, 1.5)
)

upset(fromList(AML_patient_RNA_mutations),
      nsets = length(AML_patient_RNA_mutations),
      order.by = "freq",
      mainbar.y.label = "Intersection Size RNA",
      mb.ratio = c(0.6, 0.4),
      main.bar.color = "#345170",
      sets.bar.color = "#EBAC8C",
      text.scale = c(2, 2, 2, 1.5, 2, 1.5)
)

# export csv
AML_SSM_scATA_DX <- data.frame(Location = AML_patient_ATA_mutations$DX) %>%
  separate(Location, into = c("chr", "pos", "ref", "alt"), sep = "_", remove = TRUE) %>%
  mutate(end = pos) %>% select(chr, pos, end, ref, alt, everything())

AML_SSM_scATA_REM <- data.frame(Location = AML_patient_ATA_mutations$REM) %>%
  separate(Location, into = c("chr", "pos", "ref", "alt"), sep = "_", remove = TRUE) %>%
  mutate(end = pos) %>% select(chr, pos, end, ref, alt, everything())

AML_SSM_scATA_REL <- data.frame(Location = AML_patient_ATA_mutations$REL) %>%
  separate(Location, into = c("chr", "pos", "ref", "alt"), sep = "_", remove = TRUE) %>%
  mutate(end = pos) %>% select(chr, pos, end, ref, alt, everything())

AML_SSM_scATA_all <- data.frame(Location = unique(unlist(AML_patient_ATA_mutations))) %>%
  separate(Location, into = c("chr", "pos", "ref", "alt"), sep = "_", remove = TRUE) %>%
  mutate(end = pos) %>% select(chr, pos, end, ref, alt, everything())

AML_SSM_scRNA_DX <- data.frame(Location = AML_patient_RNA_mutations$DX) %>%
  separate(Location, into = c("chr", "pos", "ref", "alt"), sep = "_", remove = TRUE) %>%
  mutate(end = pos) %>% select(chr, pos, end, ref, alt, everything())

AML_SSM_scRNA_REM <- data.frame(Location = AML_patient_RNA_mutations$REM) %>%
  separate(Location, into = c("chr", "pos", "ref", "alt"), sep = "_", remove = TRUE) %>%
  mutate(end = pos) %>% select(chr, pos, end, ref, alt, everything())

AML_SSM_scRNA_REL <- data.frame(Location = AML_patient_RNA_mutations$REL) %>%
  separate(Location, into = c("chr", "pos", "ref", "alt"), sep = "_", remove = TRUE) %>%
  mutate(end = pos) %>% select(chr, pos, end, ref, alt, everything())

AML_SSM_scRNA_all <- data.frame(Location = unique(unlist(AML_patient_RNA_mutations))) %>%
  separate(Location, into = c("chr", "pos", "ref", "alt"), sep = "_", remove = TRUE) %>%
  mutate(end = pos) %>% select(chr, pos, end, ref, alt, everything())

#write.csv(AML_SSM_scATA_DX, "../Results/AML/AML_SSM_scATA_DX.csv", row.names = FALSE)
#write.csv(AML_SSM_scATA_REM, "../Results/AML/AML_SSM_scATA_REM.csv", row.names = FALSE)
#write.csv(AML_SSM_scATA_REL, "../Results/AML/AML_SSM_scATA_REL.csv", row.names = FALSE)
#write.csv(AML_SSM_scRNA_DX, "../Results/AML/AML_SSM_scRNA_DX.csv", row.names = FALSE)
#write.csv(AML_SSM_scRNA_REM, "../Results/AML/AML_SSM_scRNA_REM.csv", row.names = FALSE)
#write.csv(AML_SSM_scRNA_REL, "../Results/AML/AML_SSM_scRNA_REL.csv", row.names = FALSE)
#write.csv(AML_SSM_scATA_all, "../Results/AML/AML_SSM_scATA_all.csv", row.names = FALSE)
#write.csv(AML_SSM_scRNA_all, "../Results/AML/AML_SSM_scRNA_all.csv", row.names = FALSE)
```

networkD3 interactive sankey plot

intersection of mutations number from different stages were manually calculated and used to generate the sankey plot below (see `# extract different stage patient mutations`).

# networkD3 Example
```{r networkD3Example, message=FALSE, warning=FALSE}
# A connection data frame is a list of flows with intensity for each flow
links <- data.frame(
  source=c("Dx","Rem", "Dx", "Dx", "Rem", "Rel"), 
  target=c("Rem","Rel", "Rel", "Dx", "Rem", "Rel"), 
  value=c(1321 + 3652, 194 + 3652, 4209 + 3652, 1236, 88, 499)
)

links <- data.frame(
  source=c("Dx","Rem", "Dx", "Dx", "Rem", "Rel"), 
  target=c("Rem","Rel", "Rel", "Dx", "Rem", "Rel"), 
  value=c(375 + 1404, 400 + 1404, 749 + 1404, 134, 122, 158)
)


links <- data.frame(
  source=c("Dx_ATA","Rem_ATA", "Dx_ATA", "Dx_ATA", "Rem_ATA", "Rel_ATA", "Dx_RNA","Rem_RNA", "Dx_RNA", "Dx_RNA", "Rem_RNA", "Rel_RNA"), 
  target=c("Rem_ATA","Rel_ATA", "Rel_ATA", "Dx_ATA", "Rem_ATA", "Rel_ATA", "Rem_RNA","Rel_RNA", "Rel_RNA", "Dx_RNA", "Rem_RNA", "Rel_RNA"), 
  value=c(1321 + 3652, 194 + 3652, 4209 + 3652, 1236, 88, 499, 375 + 1404, 400 + 1404, 749 + 1404, 134, 122, 158)
)


# merge duplicate links
links <- links %>%
  group_by(source, target) %>%
  summarise(value = sum(value), .groups = "drop")

nodes <- data.frame(
  name = unique(c(links$source, links$target))
)

links$IDsource <- match(links$source, nodes$name)-1 
links$IDtarget <- match(links$target, nodes$name)-1

# plot
p <- sankeyNetwork(
  Links = links,
  Nodes = nodes,
  Source = "IDsource",
  Target = "IDtarget",
  Value = "value",
  NodeID = "name",
  sinksRight = FALSE
)

p

#saveWidget(p, "/Users/liqun/Downloads/sankey_plot.html", selfcontained = TRUE)
```

# sampleCellsNumber
```{r sampleCellsNumber}
sample_order <- c("AML1_DX", "AML1_REM", 
                  "AML2_DX", "AML2_REM", "AML2_REL", 
                  "AML3_DX", "AML3_REM", 
                  "AML4_DX", "AML4_REL", 
                  "AML5_DX", "AML5_REM", "AML5_REL", 
                  "AML6_DX", "AML6_REM", "AML6_REL",
                  "AML7_DX", "AML7_REM", "AML7_REL", 
                  "AML8_DX", "AML8_REM", "AML8_REL", 
                  "AML9_DX", "AML9_REM", "AML9_REL", 
                  "AML10_DX", "AML10_REM", "AML10_REL",
                  "AML11_DX", "AML11_REM", "AML11_REL", 
                  "AML12_DX", "AML12_REM", "AML12_REL", 
                  "AML13_DX", "AML13_REM", "AML13_REL", 
                  "AML14_DX", "AML14_REM", 
                  "AML15_DX", "AML15_REM", "AML15_REL", 
                  "AML16_DX", "AML16_REM", "AML16_REL", 
                  "AML17_DX", "AML17_REL", 
                  "AML18_DX", "AML18_REL",
                  "AML19_DX", "AML19_REL", 
                  "AML20_DX", "AML20_REM", 
                  "AML21_DX",  "AML21_REM", "AML21_REL", 
                  "AML22_DX", "AML22_REM", "AML22_REL", 
                  "AML23_DX", "AML23_REM", "AML23_REL", 
                  "AML24_DX", "AML24_REM", "AML24_REL", 
                  "AML25_DX", "AML25_REM", "AML25_REL", 
                  "AML26_DX", "AML26_REM", "AML26_REL", 
                  "AML27_DX", "AML27_REM", "AML27_REL",
                  "AML28_REM", "AML28_REL")

AML_sampleInfo_plot <- AML_sampleInfo[, c("sampleAlias", "Cells_scRNA", "Cells_scATAC")] %>%
  pivot_longer(
    cols = starts_with("Cells"),   
    names_to = "Tag",
    values_to = "CellsNum"
  )

AML_sampleInfo_plot$sampleAlias <- factor(AML_sampleInfo_plot$sampleAlias, levels = sample_order)

p_sampleCellsNumber <- ggplot(AML_sampleInfo_plot, aes(x = sampleAlias, y = CellsNum, fill = Tag)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(yintercept = 500, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("Cells_scRNA" = "#EBAC8C", "Cells_scATAC" = "#345170")) +
  labs(title = "cells number", x = "samples", y = "cells number") +
  theme_classic() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_sampleCellsNumber
```

# mutation features similarity
```{r mutationFeaturesSimilarity}
comparisionTypes <- colnames(AML_mutation_features)

data_cosine_similarity <- as.data.frame(matrix(data = 0, nrow = length(comparisionTypes), ncol = length(comparisionTypes)))

rownames(data_cosine_similarity) <- comparisionTypes
colnames(data_cosine_similarity) <- comparisionTypes

for (i in 1:nrow(data_cosine_similarity)){
  type_i <- rownames(data_cosine_similarity)[i]
  for (j in 1:ncol(data_cosine_similarity)){
    type_j <- colnames(data_cosine_similarity)[j]
    cs <- cosine_similarity(AML_mutation_features[,type_i], AML_mutation_features[,type_j])
    data_cosine_similarity[i, j] <- cs
  }
}

mat <- as.matrix(data_cosine_similarity)
rownames(mat) <- rownames(data_cosine_similarity)
colnames(mat) <- colnames(data_cosine_similarity)
names_order <- c("scATAC_scMutrace", "scRNA_scMutrace", "scATA_scMutrace_DX", "scATA_scMutrace_REM",
                 "scATA_scMutrace_REL", "scRNA_scMutrace_DX", "scRNA_scMutrace_REM", "scRNA_scMutrace_REL",
                 "scATAC_pub", "WGS_AML", "WGS_MPN", "WGS_skin", "WGS_headSCC")
mat <- mat[names_order, names_order]
mat_upper <- mat
mat_upper[lower.tri(mat_upper)] <- NA

mat_min <- min(mat, na.rm = TRUE)
mat_max <- max(mat, na.rm = TRUE)

col_fun_2 <- colorRamp2(
  breaks = seq(mat_min, mat_max, length.out = 4),
  colors = c("#404040", "#bababa", "#f4a582", "#ca0020")
)

Heatmap(mat_upper,
        name = "Cosine\nSimilarity",
        na_col = "grey90",
        col = col_fun_2,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        cell_fun = function(j, i, x, y, width, height, fill) {
          if (!is.na(mat_upper[i, j])) {
            grid.text(sprintf("%.2f", mat_upper[i, j]), x, y, 
                      gp = gpar(fontsize = 10, col = "white"))
          }
        })

names_order_show <- c("scATAC_scMutrace", "scRNA_scMutrace",
                 "scATAC_pub", "WGS_AML", "WGS_headSCC", "WGS_skin")

mat_show <- mat[names_order_show, names_order_show]

corrplot(
  mat_show, 
  order = "original", 
  method = 'color', 
  type = 'upper', 
  col = colorRampPalette(brewer.pal(11, "Reds"))(100), 
  col.lim = c(0, 1),
  addCoef.col = "white"
)
```

# mutation features contribution
```{r mutationFeaturesContribution}
sbs_color <- c(
  "SBS1" = "#a6cee3", "SBS2" = "#33a02c", "SBS5" = "#1f78b4",
  "SBS7a" = "#e31a1c", "SBS7b" = "#fc8d59", "SBS7c" = "#fdd49e", 
  "SBS13" = "#252525", "SBS15" = "#bdbdbd", "SBS15" = "#d9d9d9",
  "SBS18" =  "#f0f0f0", "SBS19" = "#969696",  "SBS23" = "#969696",
  "SBS31" = "#737373", "SBS39" = "#525252", "SBS54" = "#252525"
)

AML_mutation_features_contribution <- read.table("../Data/HumanpAML/COSMIC_SBS96_Activities.txt", header = T, sep = "\t")
AML_mutation_features_contribution_long <- AML_mutation_features_contribution %>%
  pivot_longer(
    cols = -Samples,
    names_to = "Signature",
    values_to = "Count"
  )

AML_mutation_features_contribution_long_pct <- AML_mutation_features_contribution_long %>%
  group_by(Samples) %>%
  mutate(Percent = Count / sum(Count) * 100)

AML_mutation_features_contribution_long_pct$Samples <- factor(AML_mutation_features_contribution_long_pct$Samples, levels = c("scATAC_scMutrace", "scRNA_scMutrace", "scATA_scMutrace_DX", "scATA_scMutrace_REM",
                 "scATA_scMutrace_REL", "scRNA_scMutrace_DX", "scRNA_scMutrace_REM", "scRNA_scMutrace_REL",
                 "scATAC_pub", "WGS_AML", "WGS_MPN", "WGS_skin", "WGS_headSCC"))

p_mutation_featureDistribution <- ggplot(AML_mutation_features_contribution_long_pct, aes(x = Samples, y = Percent, fill = Signature)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = sbs_color) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylab("Percentage (%)")

p_mutation_featureDistribution
```

# census genes
```{r censusGenes}
scATA_rare_locs <- AML_SSM_scATA_bed %>% filter(freq == "rare") %>% pull(Location)
scRNA_rare_locs <- AML_SSM_scRNA_bed %>% filter(freq == "rare") %>% pull(Location)

AML_scATA_ann_tier <- AML_scATA_ann[AML_scATA_ann$Gene.refGene %in% AML_genes_tiers$Gene,] %>% filter(Location %in% scATA_rare_locs)
AML_scRNA_ann_tier <- AML_scRNA_ann[AML_scRNA_ann$Gene.refGene %in% AML_genes_tiers$Gene,] %>% filter(Location %in% scRNA_rare_locs)

AML_SSM_scATA_ann_tier <- merge(AML_SSM_scATA, AML_scATA_ann_tier, by = "Location")
AML_SSM_scRNA_ann_tier <- merge(AML_SSM_scRNA, AML_scRNA_ann_tier, by = "Location")

all_tier_genes <- unique(union(AML_scRNA_ann_tier$Gene.refGene, AML_scATA_ann_tier$Gene.refGene))

all_tier_genes_loc_scRNA <- AML_scRNA_ann_tier %>%
  dplyr::select(Location, Gene.refGene) %>%
  dplyr::filter(Gene.refGene %in% all_tier_genes)

all_tier_genes_loc_scATA <- AML_scATA_ann_tier %>%
  dplyr::select(Location, Gene.refGene) %>%
  dplyr::filter(Gene.refGene %in% all_tier_genes)

# remove samples based on cells number
sample_selected_scRNA <- c("AML1_DX", "AML1_REM", 
                  "AML2_DX", "AML2_REM", "AML2_REL", 
                  "AML3_DX", "AML3_REM", 
                  "AML4_DX", "AML4_REL", 
                  "AML5_DX", "AML5_REM", "AML5_REL", 
                  "AML6_DX", "AML6_REM", "AML6_REL",
                             "AML7_REM", "AML7_REL", 
                  "AML8_DX", "AML8_REM", "AML8_REL", 
                  "AML9_DX", "AML9_REM", "AML9_REL", 
                  "AML10_DX", "AML10_REM", "AML10_REL",
                  "AML11_DX", "AML11_REM", "AML11_REL", 
                  "AML12_DX", "AML12_REM", "AML12_REL", 
                  "AML13_DX", "AML13_REM", "AML13_REL", 
                  "AML14_DX", "AML14_REM", 
                  "AML15_DX", "AML15_REM", "AML15_REL", 
                  "AML16_DX", "AML16_REM", "AML16_REL", 
                  "AML17_DX", "AML17_REL", 
                  "AML18_DX", "AML18_REL",
                  "AML19_DX", "AML19_REL", 
                  "AML20_DX", "AML20_REM", 
                  "AML21_DX",  "AML21_REM", "AML21_REL", 
                  "AML22_DX", "AML22_REM", "AML22_REL", 
                  "AML23_DX", "AML23_REM", "AML23_REL", 
                  "AML24_DX", "AML24_REM", "AML24_REL", 
                  "AML25_DX", "AML25_REM", "AML25_REL", 
                  "AML26_DX", "AML26_REM", "AML26_REL", 
                  "AML27_DX", "AML27_REM", "AML27_REL",
                  "AML28_REM", "AML28_REL")

sample_selected_scRNA <- paste0(sample_selected_scRNA, "_scRNA")

sample_selected_scATA <- c("AML1_DX", "AML1_REM", 
                  "AML2_DX", "AML2_REM", "AML2_REL", 
                  "AML3_DX", "AML3_REM",
                  "AML4_DX", "AML4_REL",
                  "AML5_DX", "AML5_REM", "AML5_REL", 
                  "AML6_DX",             "AML6_REL",
                  "AML7_DX", "AML7_REM", "AML7_REL", 
                  "AML8_DX", "AML8_REM", "AML8_REL", 
                  
                  "AML10_DX", "AML10_REM", "AML10_REL",
                  "AML11_DX", "AML11_REM", "AML11_REL", 
                  "AML12_DX",              "AML12_REL", 
                  "AML13_DX",              "AML13_REL", 
                  "AML14_DX",  
                  "AML15_DX", "AML15_REM", "AML15_REL", 
                  "AML16_DX", "AML16_REM", "AML16_REL", 
                  "AML17_DX", "AML17_REL", 
                  "AML18_DX", "AML18_REL",
                  "AML19_DX", "AML19_REL", 
                  "AML20_DX", "AML20_REM", 
                  "AML21_DX",  "AML21_REM", "AML21_REL", 
                  "AML22_DX", "AML22_REM", "AML22_REL", 
                  
                  "AML24_DX", "AML24_REM", "AML24_REL", 
                  "AML25_DX", "AML25_REM", "AML25_REL", 
                  "AML26_DX", "AML26_REM", "AML26_REL", 
                  "AML27_DX", "AML27_REM", "AML27_REL")

# census genes group
oncogene_RAS <- c("AFF3", "CBFB", "CBL", "CSF3R", "EGFR", "ELF4", "ETS2", "ETV6", "EZH2", "FLT3", "GATA1", "HRAS", "IDH1",
                  "IKZF1", "KIT", "KRAS", "MEIS1", "MET", "MLLT1", "MLLT3", "MLLT4", "MYC", "NPM1", "NRAS", "NUP214",
                  "PLXNB1", "PTPN11", "RUNX1", "RUNX1T1", "SPI1", "SYK", "TAF1", "VIM", "WT1")

suppressor <- c("APC", "ARID1A", "ASXL1", "AXIN1", "CBFA2T2", "CBL", "CEBPA", "CREBBP", "DICER1", "ERF", "ETS2", "ETV6", 
                "EZH2", "FAT1", "FLT3", "IDH1", "IKZF1", "KDM5A", "LLGL1", "MAD1L1", "NF1", "NPM1", "PHF6", "PTEN", 
                "PTPN11", "PTPRJ", "PTPRT", "RUNX1", "RYR1", "SETD2", "SPI1", "SYK", "TET2", "TP53", "TRIM24", "VIM", "WT1")

transcription_factors <- c("AFF3", "ANKRD30A", "BCOR", "CBFB", "CEBPA", "CREBBP", "ELF1", "ELF4", "EP300", "FOXP2", "GATA1", 
                           "GATA2", "HDAC9", "HOXA5", "HOXB3", "IKZF1", "MED23", "MEIS1", "NCOR1", "NONO", "PHF2", "RUNX1", 
                           "RUNX1T1", "SIN3A", "SRCAP", "TAF1", "TAF1L", "THRAP3", "TLE2", "TRIM24", "TRPS1", "TRRAP", "VEZF1", "WT1")

epigenetic_modifier <- c("ASXL1", "ASXL2", "ATRX", "CHD4", "CREBBP", "EP300", "EZH2", "IDH1", "IDH2", "JMJD1C", "KDM5A", "MLLT1", "MLLT10", 
                         "MLLT3", "MLLT4", "MLLT6", "MYH11", "MYH2", "MYH4", "SETD2", "SRCAP", "TET2", "TRRAP")

tyrosine_kinase <- c("CNKSR1", "EGFR", "EPHA6", "FLT3", "FLT4", "KIT", "MET", "PTK6", "ROR1", "SYK", "TYK2")

nuclear_transport <- c("NPM1", "NUP205", "NUP214")

phosphatase <- c("ANKRD28", "APC", "AXIN1", "KIAA1244", "PPM1F", "PPP1R14B", "PPP1R14C", "PPP1R15A", "PPP1R15B", 
                 "PPP2R2D", "PPP4R4", "PPP6R3", "PTEN", "PTPN11", "PTPRA", "PTPRF", "PTPRH", "PTPRJ", "PTPRT", "TPTE")

structural_integrity <- c("BSN", "COL12A1", "CSPG4", "DNAH2", "DNAH9", "FLNB", "LAMA1", "MACF1", "MAP7", "NCAM1", "PCLO", "RELN", "TRO")

ETS_Family <- c("ELF1", "ELF4", "ERF", "ETS2", "ETV6", "SPI1")

cohesin <- c("RAD21", "SMC3", "STAG2")

splicing <- c("DHX15", "NONO", "SUPT5H", "U2AF1")

growth_factor <- c("CSF3R", "VEGFC")

genes_groups <- c(oncogene_RAS, suppressor, transcription_factors, epigenetic_modifier, 
                  tyrosine_kinase, nuclear_transport, phosphatase, structural_integrity,
                  ETS_Family, cohesin, splicing, growth_factor)


sample_selected_scATA <- paste0(sample_selected_scATA, "_scATA")

AML_sample_mutations <- as.data.frame(matrix(data = 0, nrow = length(sample_selected_scRNA) + length(sample_selected_scATA), ncol = length(all_tier_genes)))
colnames(AML_sample_mutations) <- all_tier_genes
rownames(AML_sample_mutations) <- c(sample_selected_scRNA, sample_selected_scATA)
AML_sample_mutations$id <- rownames(AML_sample_mutations)

AML_sample_mutations <- separate(AML_sample_mutations, col = "id", into = c("patient", "stage", "type"), sep = "_", remove = TRUE)
AML_sample_mutations$sampleID <- paste(AML_sample_mutations$patient, AML_sample_mutations$stage, sep = "_")
AML_sample_mutations <- AML_sample_mutations[,c("patient", "stage", "type", "sampleID", all_tier_genes)]

col_AML_scRNA_ann_tier <- setdiff(colnames(AML_SSM_scRNA_ann_tier), colnames(AML_SSM_scRNA_ann_tier)[grep("_L", colnames(AML_SSM_scRNA_ann_tier))])
AML_SSM_scRNA_ann_tier <- AML_SSM_scRNA_ann_tier[,col_AML_scRNA_ann_tier]

# scRNA
for (i in 1:nrow(AML_SSM_scRNA_ann_tier)){
  gene_tmp <- AML_SSM_scRNA_ann_tier[i,]$Gene.refGene
  mut_samples <- colnames(AML_SSM_scRNA_ann_tier)[AML_SSM_scRNA_ann_tier[i,] == "Mut"]
  for (mut_sample in mut_samples){
    if (paste0(mut_sample, "_scRNA") %in% rownames(AML_sample_mutations)){
       AML_sample_mutations[paste0(mut_sample, "_scRNA"), gene_tmp] <- 1
    }
  }
}

# scATA
for (i in 1:nrow(AML_SSM_scATA_ann_tier)){
  gene_tmp <- AML_SSM_scATA_ann_tier[i,]$Gene.refGene
  mut_samples <- colnames(AML_SSM_scATA_ann_tier)[AML_SSM_scATA_ann_tier[i,] == "Mut"]
  for (mut_sample in mut_samples){
    if (paste0(mut_sample, "_scATA") %in% rownames(AML_sample_mutations)){
      AML_sample_mutations[paste0(mut_sample, "_scATA"), gene_tmp] <- 1
    }
  }
}

col_fun <- c(
  "0" = "#404040",
  "1" = "#ca0020"
)

sample_order_plotMut <- c(
  "AML1_DX_scRNA", "AML1_DX_scATA", "AML1_REM_scRNA", "AML1_REM_scATA",
  "AML2_DX_scRNA", "AML2_DX_scATA", "AML2_REM_scRNA", "AML2_REM_scATA", "AML2_REL_scRNA", "AML2_REL_scATA",
  "AML3_DX_scRNA", "AML3_DX_scATA", "AML3_REM_scRNA", "AML3_REM_scATA",
  "AML4_DX_scRNA", "AML4_DX_scATA", "AML4_REL_scRNA", "AML4_REL_scATA",
  "AML5_DX_scRNA", "AML5_DX_scATA", "AML5_REM_scRNA", "AML5_REM_scATA", "AML5_REL_scRNA", "AML5_REL_scATA",
  "AML6_DX_scRNA", "AML6_DX_scATA", "AML6_REM_scRNA", "AML6_REL_scRNA", "AML6_REL_scATA",
  "AML7_DX_scATA", "AML7_REM_scRNA", "AML7_REM_scATA", "AML7_REL_scRNA", "AML7_REL_scATA",
  "AML8_DX_scRNA", "AML8_DX_scATA", "AML8_REM_scRNA", "AML8_REM_scATA", "AML8_REL_scRNA", "AML8_REL_scATA",
  "AML9_DX_scRNA", "AML9_REM_scRNA", "AML9_REL_scRNA",
  "AML10_DX_scRNA", "AML10_DX_scATA", "AML10_REM_scRNA", "AML10_REM_scATA", "AML10_REL_scRNA", "AML10_REL_scATA",
  "AML11_DX_scRNA", "AML11_DX_scATA", "AML11_REM_scRNA", "AML11_REM_scATA", "AML11_REL_scRNA", "AML11_REL_scATA",
  "AML12_DX_scRNA", "AML12_DX_scATA",  "AML12_REM_scRNA", "AML12_REL_scRNA", "AML12_REL_scATA",
  "AML13_DX_scRNA", "AML13_DX_scATA", "AML13_REM_scRNA", "AML13_REL_scRNA", "AML13_REL_scATA",
  "AML14_DX_scRNA", "AML14_DX_scATA", "AML14_REM_scRNA",
  "AML15_DX_scRNA", "AML15_DX_scATA", "AML15_REM_scRNA", "AML15_REM_scATA", "AML15_REL_scRNA", "AML15_REL_scATA",
  "AML16_DX_scRNA", "AML16_DX_scATA", "AML16_REM_scRNA", "AML16_REM_scATA", "AML16_REL_scRNA", "AML16_REL_scATA",
  "AML17_DX_scRNA", "AML17_DX_scATA", "AML17_REL_scRNA", "AML17_REL_scATA",
  "AML18_DX_scRNA", "AML18_DX_scATA", "AML18_REL_scRNA", "AML18_REL_scATA",
  "AML19_DX_scRNA", "AML19_DX_scATA", "AML19_REL_scRNA", "AML19_REL_scATA",
  "AML20_DX_scRNA", "AML20_DX_scATA", "AML20_REM_scRNA", "AML20_REM_scATA",
  "AML21_DX_scRNA", "AML21_DX_scATA", "AML21_REM_scRNA", "AML21_REM_scATA", "AML21_REL_scRNA", "AML21_REL_scATA",
  "AML22_DX_scRNA", "AML22_DX_scATA", "AML22_REM_scRNA", "AML22_REM_scATA", "AML22_REL_scRNA", "AML22_REL_scATA",
  "AML23_DX_scRNA", "AML23_REM_scRNA", "AML23_REL_scRNA",
  "AML24_DX_scRNA", "AML24_DX_scATA",  "AML24_REM_scRNA", "AML24_REM_scATA", "AML24_REL_scRNA", "AML24_REL_scATA",
  "AML25_DX_scRNA", "AML25_DX_scATA", "AML25_REM_scRNA", "AML25_REM_scATA", "AML25_REL_scRNA", "AML25_REL_scATA",
  "AML26_DX_scRNA", "AML26_DX_scATA", "AML26_REM_scRNA", "AML26_REM_scATA", "AML26_REL_scRNA", "AML26_REL_scATA",
  "AML27_DX_scRNA", "AML27_DX_scATA", "AML27_REM_scRNA", "AML27_REM_scATA", "AML27_REL_scRNA", "AML27_REL_scATA",
  "AML28_REM_scRNA", "AML28_REL_scRNA"
)

sampleID_p <- sapply(sample_order_plotMut, function(x) strsplit(x, "_")[[1]][1])
Stages_p <- sapply(sample_order_plotMut, function(x) strsplit(x, "_")[[1]][2])
Types_p <- sapply(sample_order_plotMut, function(x) strsplit(x, "_")[[1]][3])
group_p <- AML_sampleInfo$SubgroupAnnotation[match(paste(sampleID_p, Stages_p, sep = "_"), AML_sampleInfo$sampleAlias)]
risk_p <- AML_sampleInfo$RiskGroup[match(paste(sampleID_p, Stages_p, sep = "_"), AML_sampleInfo$sampleAlias)]

row_anno <- rowAnnotation(
  Sample = anno_text(sampleID_p, gp = gpar(fontsize = 5)),
  Stage = Stages_p,
  Type = Types_p,
  Group = group_p,
  Risk = risk_p,
  col = list(Stage = c("DX" = "#d9d9d9", "REM" = "#969696", "REL" = "#000000"),
             Risk = c("Low" = "#01665e", "High" = "#8c510a"),
             Type = c("scRNA" = "#EBAC8C", "scATA" = "#345170"),
             Group = c("RUNX" = "#33a02c", "MLL" = "#ff7f00", "FLT" = "#6a3d9a", "CBFB" = "#ffff99", "Other" = "#1f78b4"))
  )

mat <- as.matrix(AML_sample_mutations[sample_order_plotMut, 5:ncol(AML_sample_mutations)])
zero_count <- colSums(mat == 0)
mat_sorted <- mat[, order(zero_count, decreasing = FALSE)]

Heatmap(mat_sorted,
        name = "Mutation",
        left_annotation = row_anno,
        col = col_fun,
        show_row_names = FALSE,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        column_names_gp = gpar(fontsize = 5))
```

# calMutFraction
```{r calMutFraction}
sample_selected_scRNA <- c("AML1_DX", "AML1_REM", 
                  "AML2_DX", "AML2_REM", "AML2_REL", 
                  "AML3_DX", "AML3_REM", 
                  "AML4_DX", "AML4_REL", 
                  "AML5_DX", "AML5_REM", "AML5_REL", 
                  "AML6_DX", "AML6_REM", "AML6_REL",
                             "AML7_REM", "AML7_REL", 
                  "AML8_DX", "AML8_REM", "AML8_REL", 
                  "AML9_DX", "AML9_REM", "AML9_REL", 
                  "AML10_DX", "AML10_REM", "AML10_REL",
                  "AML11_DX", "AML11_REM", "AML11_REL", 
                  "AML12_DX", "AML12_REM", "AML12_REL", 
                  "AML13_DX", "AML13_REM", "AML13_REL", 
                  "AML14_DX", "AML14_REM", 
                  "AML15_DX", "AML15_REM", "AML15_REL", 
                  "AML16_DX", "AML16_REM", "AML16_REL", 
                  "AML17_DX", "AML17_REL", 
                  "AML18_DX", "AML18_REL",
                  "AML19_DX", "AML19_REL", 
                  "AML20_DX", "AML20_REM", 
                  "AML21_DX",  "AML21_REM", "AML21_REL", 
                  "AML22_DX", "AML22_REM", "AML22_REL", 
                  "AML23_DX", "AML23_REM", "AML23_REL", 
                  "AML24_DX", "AML24_REM", "AML24_REL", 
                  "AML25_DX", "AML25_REM", "AML25_REL", 
                  "AML26_DX", "AML26_REM", "AML26_REL", 
                  "AML27_DX", "AML27_REM", "AML27_REL",
                  "AML28_REM", "AML28_REL")

sample_selected_scATA <- c("AML1_DX", "AML1_REM", 
                  "AML2_DX", "AML2_REM", "AML2_REL", 
                  "AML3_DX", "AML3_REM",
                  "AML4_DX", "AML4_REL",
                  "AML5_DX", "AML5_REM", "AML5_REL", 
                  "AML6_DX",             "AML6_REL",
                  "AML7_DX", "AML7_REM", "AML7_REL", 
                  "AML8_DX", "AML8_REM", "AML8_REL", 
                  
                  "AML10_DX", "AML10_REM", "AML10_REL",
                  "AML11_DX", "AML11_REM", "AML11_REL", 
                  "AML12_DX",              "AML12_REL", 
                  "AML13_DX",              "AML13_REL",
                  "AML15_DX", "AML15_REM", "AML15_REL", 
                  "AML16_DX", "AML16_REM", "AML16_REL", 
                  "AML17_DX", "AML17_REL", 
                  "AML18_DX", "AML18_REL",
                  "AML19_DX", "AML19_REL", 
                  "AML20_DX", "AML20_REM", 
                  "AML21_DX",  "AML21_REM", "AML21_REL", 
                  "AML22_DX", "AML22_REM", "AML22_REL",
                  "AML24_DX", "AML24_REM", "AML24_REL", 
                  "AML25_DX", "AML25_REM", "AML25_REL", 
                  "AML26_DX", "AML26_REM", "AML26_REL", 
                  "AML27_DX", "AML27_REM", "AML27_REL")

num_cutoff <- 0
AML_mutationFraction <- data.frame(
  Location = character(),
  Gene = character(),
  Sample = character(),
  Type = character(),
  DX_Mut_Num = numeric(),
  REM_Mut_Num = numeric(),
  REL_Mut_Num = numeric(),
  DX_Total_Num = numeric(),
  REM_Total_Num = numeric(),
  REL_Total_Num = numeric(),
  DX = numeric(),
  REM = numeric(),
  REL = numeric(),
  stringsAsFactors = FALSE
)

loc_total <- unique(c(AML_scATA_NotCensus_locs, AML_scRNA_NotCensus_locs))
line_i <- 0

for (loc in loc_total){
  # message(paste0("Processing ", loc))
  # gene
  gene <- NA
  if (loc %in% AML_scATA_ann$Location){
    gene <- AML_scATA_ann %>% 
      filter(Location == loc) %>% 
      pull(Gene.refGene)
  } else if (loc %in% AML_scRNA_ann$Location){
    gene <- AML_scRNA_ann %>% 
      filter(Location == loc) %>% 
      pull(Gene.refGene)
  }
  
  # ---------- ATA ----------
  samples_ata_loc <- AML_SSM_scATA %>%
    filter(Location == loc) %>%
    select(any_of(sample_selected_scATA)) %>%
    unlist() %>%
    { names(.)[. == "Mut"] }
  
  if(length(samples_ata_loc) > 0){
    # message(paste0("  Processing scATA..."))
    sample_ids <- unique(str_extract(samples_ata_loc, "AML[0-9]+"))
    for (sample_id in sample_ids){
      line_i <- line_i + 1
      cells_all   <- retriveCellsByLocation(AML_ATA_CSM_list[[sample_id]], loc)
      cells_mut   <- cells_all$Mut
      cells_total <- cells_all$Total
      
      num_mut_dx  <- sum(grepl("DX", cells_mut))
      num_mut_rem <- sum(grepl("REM", cells_mut))
      num_mut_rel <- sum(grepl("REL", cells_mut))
      num_total_dx  <- sum(grepl("DX", cells_total))
      num_total_rem <- sum(grepl("REM", cells_total))
      num_total_rel <- sum(grepl("REL", cells_total))
      
      # QC
      if (num_total_dx >= num_cutoff & num_total_rem >= num_cutoff & num_total_rel >= num_cutoff){
    
        frac_dx  <- ifelse(num_total_dx > 0, num_mut_dx/num_total_dx, 0)
        frac_rem <- ifelse(num_total_rem > 0, num_mut_rem/num_total_rem, 0)
        frac_rel <- ifelse(num_total_rel > 0, num_mut_rel/num_total_rel, 0)
        
        AML_mutationFraction[line_i, ] <- list(
          Location = loc,
          Gene = gene,
          Sample = sample_id,
          Type = "scATA",
          DX_Mut_Num = num_mut_dx,
          REM_Mut_Num = num_mut_rem,
          REL_Mut_Num = num_mut_rel,
          DX_Total_Num = num_total_dx,
          REM_Total_Num = num_total_rem,
          REL_Total_Num = num_total_rel,
          DX = frac_dx,
          REM = frac_rem,
          REL = frac_rel
        )
      }
    }
  }
  
  # ---------- RNA ----------
  samples_rna_loc <- AML_SSM_scRNA %>%
    filter(Location == loc) %>%
    select(any_of(sample_selected_scRNA)) %>%
    unlist() %>%
    { names(.)[. == "Mut"] }
  
  if(length(samples_rna_loc) > 0){
    sample_ids <- unique(str_extract(samples_rna_loc, "AML[0-9]+"))
    # message(paste0("  Processing scRNA..."))
    for (sample_id in sample_ids){
      line_i <- line_i + 1
      cells_all   <- retriveCellsByLocation(AML_RNA_CSM_list[[sample_id]], loc)
      cells_mut   <- cells_all$Mut
      cells_total <- cells_all$Total
      
      num_mut_dx  <- sum(grepl("DX", cells_mut))
      num_mut_rem <- sum(grepl("REM", cells_mut))
      num_mut_rel <- sum(grepl("REL", cells_mut))
      num_total_dx  <- sum(grepl("DX", cells_total))
      num_total_rem <- sum(grepl("REM", cells_total))
      num_total_rel <- sum(grepl("REL", cells_total))
      
      # QC
      if (num_total_dx >= num_cutoff & num_total_rem >= num_cutoff & num_total_rel >= num_cutoff){
        
        frac_dx  <- ifelse(num_total_dx > 0, num_mut_dx/num_total_dx, 0)
        frac_rem <- ifelse(num_total_rem > 0, num_mut_rem/num_total_rem, 0)
        frac_rel <- ifelse(num_total_rel > 0, num_mut_rel/num_total_rel, 0)
        
        AML_mutationFraction[line_i, ] <- list(
          Location = loc,
          Gene = gene,
          Sample = sample_id,
          Type = "scRNA",
          DX_Mut_Num = num_mut_dx,
          REM_Mut_Num = num_mut_rem,
          REL_Mut_Num = num_mut_rel,
          DX_Total_Num = num_total_dx,
          REM_Total_Num = num_total_rem,
          REL_Total_Num = num_total_rel,
          DX = frac_dx,
          REM = frac_rem,
          REL = frac_rel
        )
      }
    }
  }
}

AML_mutationFraction <- na.omit(AML_mutationFraction)

stage_color_setting <- c(
    "DX" = "#4DAF4A",
    "REM" = "#377EB8",
    "REL"   = "#E41A1C"
  )

common_loc_scATA_scRNA <- unique(intersect(AML_SSM_scATA$Location, AML_SSM_scRNA$Location))
common_loc_scATA_scRNA_census <- intersect(AML_scATA_ann_tier$Location, AML_scRNA_ann_tier$Location)

AML_mutationFraction_commonLocs <- AML_mutationFraction %>% filter(Location %in% common_loc_scATA_scRNA)
AML_mutationFraction_commonLocs$Tag <- paste(AML_mutationFraction_commonLocs$Location, AML_mutationFraction_commonLocs$Sample, sep = "-")

AML_mutationFraction_commonLocs <- AML_mutationFraction_commonLocs %>%
  group_by(Tag) %>%
  filter(n() > 1) %>%
  ungroup()

df_fractionSimilarity_DX_plot <- AML_mutationFraction_commonLocs %>%
  #filter(DX_Num >= 5) %>%
  select(Location, Sample, Type, DX) %>%
  pivot_wider(names_from = Type, values_from = DX) %>%
  drop_na() %>%
  mutate(Stage = "DX")

df_fractionSimilarity_REM_plot <- AML_mutationFraction_commonLocs %>%
  #filter(REM_Num >= 5) %>%
  select(Location, Sample, Type, REM) %>%
  pivot_wider(names_from = Type, values_from = REM) %>%
  drop_na() %>%
  mutate(Stage = "REM")

df_fractionSimilarity_REL_plot <- AML_mutationFraction_commonLocs %>%
  #filter(REL_Num >= 5) %>%
  select(Location, Sample, Type, REL) %>%
  pivot_wider(names_from = Type, values_from = REL) %>%
  drop_na() %>%
  mutate(Stage = "REL")

df_fractionSimilarity_all <- bind_rows(
  df_fractionSimilarity_DX_plot,
  df_fractionSimilarity_REM_plot,
  df_fractionSimilarity_REL_plot
)

df_fractionSimilarity_long <- AML_mutationFraction_commonLocs %>%
  select(Location, Sample, Type, DX, REM, REL) %>%
  pivot_longer(cols = c(DX, REM, REL), names_to = "Stage", values_to = "Fraction") %>%
  pivot_wider(names_from = Type, values_from = Fraction)

df_fractionSimilarity_long$Stage <- factor(df_fractionSimilarity_long$Stage, levels = c("DX", "REM", "REL"))

ggplot(df_fractionSimilarity_all, aes(x = scATA, y = scRNA, color = Stage)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  geom_smooth(method = "lm", se = TRUE, aes(fill = Stage), alpha = 0.2) +
  stat_cor(aes(label = paste(..r.label.., ..p.label.., sep = "~`,`~"), group = Stage),
           method = "pearson",
           show.legend = FALSE) +
  scale_color_manual(values = stage_color_setting) + 
  scale_fill_manual(values = stage_color_setting) +
  labs(
    x = "Fraction (scATA)",
    y = "Fraction (scRNA)",
    title = "Comparison of mutation fractions between scATA and scRNA"
  ) +
  theme_classic()

df_wide <- df_fractionSimilarity_long %>%
  pivot_wider(
    names_from = Stage,
    values_from = c(scATA, scRNA)
  )


p_scRNA_fractionMutation <- ggplot(df_fractionSimilarity_long, aes(x = as.factor(Stage), y = scRNA, color=Stage)) + 
  labs(x="", y="Fraction of mutant cells") +
  geom_boxplot(width=0.5, lwd = 1, outliers = FALSE) +
  geom_line(aes(group=Location), color="gray" ,position = position_dodge(0.2), alpha = 0.1) +
  geom_point(aes(fill=Stage, group=Location), size = 0.1,  position = position_dodge(0.2)) +
  scale_color_manual(values = stage_color_setting) + theme_classic() +
  stat_compare_means(
    comparisons = list(
      c("DX", "REL"),
      c("DX", "REM"),
      c("REL", "REM")
    ),
    method = "wilcox.test",
    label = "p.signif"
  )

p_scATA_fractionMutation <- ggplot(df_fractionSimilarity_long, aes(x = as.factor(Stage), y = scATA, color=Stage)) + 
  labs(x="", y="Fraction of mutant cells") +
  geom_boxplot(width=0.5, lwd = 1, outliers = FALSE) +
  geom_line(aes(group=Location), color="gray" ,position = position_dodge(0.2), alpha = 0.1) +
  geom_point(aes(fill=Stage, group=Location), size = 0.1,  position = position_dodge(0.2)) +
  scale_color_manual(values = stage_color_setting) + theme_classic() +
  stat_compare_means(
    comparisons = list(
      c("DX", "REL"),
      c("DX", "REM"),
      c("REL", "REM")
    ),
    method = "wilcox.test",
    label = "p.signif"
  )

p_scRNA_fractionMutation
p_scATA_fractionMutation
```

*JointCCA_AML8.rds* contains the UMAP coordinates and metadata for the integrated scRNA-seq and scATAC-seq data of AML8 sample and was generated using ArchR package (see `0_pipelineFunctions.R`> # pipelines for processing scATAC-seq data).

*AML8_scRNA_meta.rds* contains the metadata for scRNA-seq data of AML8 sample and was generated using Seurat package (see `0_pipelineFunctions.R` > # pipelines for porcessing scRNA-seq data).

# RNAIntegration Example
```{r RNAIntegrationExample}
data_save_block <- readRDS("../Data/HumanpAML/JointCCA_AML8.rds")
meta_scRNA <- readRDS("../Data/HumanpAML/AML8_scRNA_meta.rds")


df_AML08_CCA <- data.frame(
  UMAP1 = data_save_block$UMAP1,
  UMAP2 = data_save_block$UMAP2,
  Assay = data_save_block$Assay,
  Group = data_save_block$Group
)

getSampleStage<- function(x, meta_scRNA) {
  if (grepl("_reference$", x)) {
    cell_id <- sub("_reference$", "", x)
    if (cell_id %in% rownames(meta_scRNA)) {
      return(meta_scRNA[cell_id, "orig.ident"])
    } else {
      return(NA)
    }
  } else if (grepl("_query$", x)) {
    sample <- sub("#.*", "", x)
    return(sample)
  } else {
    return(NA)
  }
}

df_AML08_CCA$stage <- sapply(rownames(df_AML08_CCA), getSampleStage, meta_scRNA = meta_scRNA)

assay_colors <- c("ATAC" = "#272E6A", "RNA" = "#D51F26")

stage_colors <- c(
    "AML8_DX" = "#d9d9d9",
    "AML8_REM" = "#969696",
    "AML8_REL"   = "#000000")

group_colors <- c("B.Cell" = "#7DB18A", "CD16.Monocytes" = "#6F4A98", "CD4.Memory" = "#4C502D",
                  "CD4.Naive" = "#201C19", "CD8.Effector" = "#2A5B31", "CD8.Memory" = "#69AD46",
                  "CD8.Naive" = "#5B521A", "cDC" = "#8587B1", "CLP" = "#B9D5AB", "Early.Basophil" = "#CD6D82",
                  "Early.Erythrocyte" = "#B12579", "GMP" = "#25316E", "HSC" = "#EAA63A", "Late.Erythrocyte" = "#EAAFC0",
                  "Monocytes" = "#406DB1", "NK" = "#4C662E", "pDC" = "#62B1E1", "Plasma" = "#D2E7EB",
                  "Pre.B.Cell" = "#45948E", "Progenitor" = "#AA872C", "Unknown" = "#EAEAEA")

p_UMAPbyAssay <- ggplot(df_AML08_CCA, aes(x = UMAP1, y = UMAP2, color = Assay)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_manual(values = assay_colors) +
  theme_classic() +
  labs(title = "UMAP by Assay")

p_UMAPbyGroup <- ggplot(df_AML08_CCA, aes(x = UMAP1, y = UMAP2, color = Group)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_manual(values = group_colors) +
  theme_classic() +
  labs(title = "UMAP by Group")

p_UMAPbyStage <- ggplot(df_AML08_CCA, aes(x = UMAP1, y = UMAP2, color = stage)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_manual(values = stage_colors) +
  theme_classic() +
  labs(title = "UMAP by Stage")

p_UMAPbyAssay
p_UMAPbyGroup
p_UMAPbyStage

# from 0_pipelineFunctions.R > # pipelines for processing scATAC-seq data
# scATAC_data <- loadArchRProject("../Data/AML/singleCellMatrix/scATACTest/")
# proj <- addPeak2GeneLinks(scATAC_data, k = 20)
# Get P2G Links
# p2g <- getPeak2GeneLinks(proj)
# p_plotPeak2GeneHeatmap <- plotPeak2GeneHeatmap(proj, groupBy = "predictedGroup_Un", palGroup = group_colors)
# ComplexHeatmap::draw(p_plotPeak2GeneHeatmap, heatmap_legend_side = "right")
```

# celltype annotation similarity
```{r celltypeAnnotationSimilarity}
data_cellTypeAnn_similarity <- as.data.frame(matrix(data = 0, nrow = length(setdiff(cellTypes, "Unknown")), 
                                                    ncol = length(setdiff(cellTypes, "Unknown"))))

rownames(data_cellTypeAnn_similarity) <- setdiff(cellTypes, "Unknown")
colnames(data_cellTypeAnn_similarity) <- setdiff(cellTypes, "Unknown")

for (i in 1:nrow(data_cellTypeAnn_similarity)){
  # row > scRN
  ct_scRNA <- rownames(data_cellTypeAnn_similarity)[i]
  cell_barcode_scRNA <- AML_scRNA_meta %>% 
    mutate(CellTag = rownames(.)) %>%  
    dplyr::filter(Classified_Celltype == ct_scRNA) %>%
    pull(CellTag)
  
  for (j in 1:ncol(data_cellTypeAnn_similarity)){
    ct_scATA <- colnames(data_cellTypeAnn_similarity)[j]
    cell_barcode_scATA <- AML_scATA_meta %>% 
      dplyr::filter(predictedGroup_Un == ct_scATA) %>%
      pull(predictedCell_RNAQun)
    
    similarity_score <- length(unique(intersect(cell_barcode_scRNA, cell_barcode_scATA)))/length(unique(union(cell_barcode_scRNA, cell_barcode_scATA)))
    data_cellTypeAnn_similarity[i, j] <- similarity_score
  }
}

breaks <- seq(min(as.matrix(data_cellTypeAnn_similarity)), 
              max(as.matrix(data_cellTypeAnn_similarity)), 
              length.out = 4)

col_fun <- colorRamp2(
  breaks = breaks,
  colors = c("#404040", "#bababa", "#f4a582", "#ca0020")
)

row_types <- rownames(as.matrix(data_cellTypeAnn_similarity))
col_types <- colnames(as.matrix(data_cellTypeAnn_similarity))

group_colors <- c("B.Cell" = "#7DB18A", "CD16.Monocytes" = "#6F4A98", "CD4.Memory" = "#4C502D",
                  "CD4.Naive" = "#201C19", "CD8.Effector" = "#2A5B31", "CD8.Memory" = "#69AD46",
                  "CD8.Naive" = "#5B521A", "cDC" = "#8587B1", "CLP" = "#B9D5AB", "Early.Basophil" = "#CD6D82",
                  "Early.Erythrocyte" = "#B12579", "GMP" = "#25316E", "HSC" = "#EAA63A", "Late.Erythrocyte" = "#EAAFC0",
                  "Monocytes" = "#406DB1", "NK" = "#4C662E", "pDC" = "#62B1E1", "Plasma" = "#D2E7EB",
                  "Pre.B.Cell" = "#45948E", "Progenitor" = "#AA872C", "Unknown" = "#EAEAEA")

order_annotation <- c("HSC", "Progenitor",
                      "GMP", "Monocytes", "CD16.Monocytes", "cDC", "pDC",
                      "CLP", "Pre.B.Cell", "B.Cell", "CD4.Naive", "CD4.Memory", "CD8.Naive", "CD8.Memory", "CD8.Effector", "NK",
                      "Early.Erythrocyte", "Late.Erythrocyte",  "Early.Basophil", "Plasma")

mat_ord <- as.matrix(data_cellTypeAnn_similarity)[order_annotation, order_annotation]
row_types_ord <- row_types[match(order_annotation, row_types)]
col_types_ord <- col_types[match(order_annotation, col_types)]

row_ha <- rowAnnotation(
  CellType = row_types_ord,
  col = list(CellType = group_colors),
  show_annotation_name = FALSE,
  annotation_width = unit(0.5, "cm")
)

col_ha <- HeatmapAnnotation(
  CellType = col_types_ord,
  col = list(CellType = group_colors),
  show_annotation_name = FALSE,
  annotation_height = unit(0.5, "cm")
)


p_cellType_similarity <- Heatmap(
  mat_ord,
  name = "Similarity",
  col = col_fun,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),
  row_title = NULL,
  column_title = NULL,
  top_annotation = col_ha,
  left_annotation = row_ha,
  rect_gp = gpar(col = "white", lwd = 1)
)

p_cellType_similarity
```

*data_commonBed_similarity_all.csv* and *data_commonBed_similarity_noCensus.csv* contain the sample stage mutation similarity matrices calculated based on common mutated genomic regions (with and without Census genes respectively) and were generated using the functions in `0_pipelineFunctions.R` > pipelines for generating sample stage mutation similarity matrix.

# sample stage mutation similarity
```{r sampleStageMutationSimilarity}
# may spent 2hrs
# generate data_commonBed_similarity_all.csv and data_commonBed_similarity_noCensus.csv 
# using 0_pipelineFunctions.R > # pipelines for generating sample stage mutation similarity matrix

sample_order_keep <- c(
                      "AML1_DX", "AML1_REM", 
                      "AML2_DX", "AML2_REM", "AML2_REL", 
                      "AML3_DX", "AML3_REM", 
                      "AML4_DX", "AML4_REL", 
                      "AML5_DX", "AML5_REM", "AML5_REL", 
                      "AML6_DX", "AML6_REL", 
                      "AML7_REM", "AML7_REL", 
                      "AML8_DX", "AML8_REM", "AML8_REL", 
                      "AML10_DX", "AML10_REM", "AML10_REL",
                      "AML11_DX", "AML11_REM", "AML11_REL", 
                      "AML12_DX", "AML12_REL", 
                      "AML13_DX", "AML13_REL", 
                      "AML15_DX", "AML15_REM", "AML15_REL", 
                      "AML16_DX", "AML16_REM", "AML16_REL", 
                      "AML17_DX", "AML17_REL", 
                      "AML18_DX", "AML18_REL",
                      "AML19_DX", "AML19_REL", 
                      "AML20_DX", "AML20_REM", 
                      "AML21_DX",  "AML21_REM", "AML21_REL", 
                      "AML22_DX", "AML22_REM", "AML22_REL", 
                      "AML24_DX", "AML24_REM", "AML24_REL", 
                      "AML26_DX", "AML26_REM", "AML26_REL", 
                      "AML27_DX", "AML27_REM", "AML27_REL")

data_commonBed_similarity_lenient <- read.csv("../Data/HumanpAML/data_commonBed_similarity_all.csv", row.names = 1)
data_commonBed_similarity_lenient <- data_commonBed_similarity_lenient[sample_order_keep, sample_order_keep]

data_commonBed_similarity_stringent <- read.csv("../Data/HumanpAML/data_commonBed_similarity_noCensus.csv", row.names = 1)
data_commonBed_similarity_stringent <- data_commonBed_similarity_stringent[sample_order_keep, sample_order_keep]

row_prefix <- sapply(rownames(data_commonBed_similarity_lenient), function(x) strsplit(x, "_")[[1]][1])
row_suffix <- sapply(rownames(data_commonBed_similarity_lenient), function(x) strsplit(x, "_")[[1]][2])
col_prefix <- sapply(colnames(data_commonBed_similarity_lenient), function(x) strsplit(x, "_")[[1]][1])
col_suffix <- sapply(colnames(data_commonBed_similarity_lenient), function(x) strsplit(x, "_")[[1]][2])

col_anno <- HeatmapAnnotation(
  Sample = anno_text(col_prefix, gp = gpar(fontsize = 8)),
  scATACStage = col_suffix,
  col = list(scATACStage = c("DX" = "#d9d9d9", "REM" = "#969696", "REL" = "#000000"))
)

row_anno <- rowAnnotation(
  Sample = anno_text(row_prefix, gp = gpar(fontsize = 8)),
  scRNAStage = row_suffix,
  col = list(scRNAStage = c("DX" = "#d9d9d9", "REM" = "#969696", "REL" = "#000000")))

col_fun <- colorRamp2(
  breaks = c(0, 0.33, 0.66, 1),
  colors = c("#404040", "#bababa", "#f4a582", "#ca0020")
)

Heatmap(data_commonBed_similarity_lenient,
        top_annotation = col_anno,
        left_annotation = row_anno,
        col = col_fun,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        row_names_gp = gpar(fontsize = 6),
        column_names_gp = gpar(fontsize = 6),
        width = unit(10, "cm"), height = unit(10, "cm"),
        layer_fun = function(j, i, x, y, width, height, fill) {
          
          grid.rect(x = x, y = y, width = width, height = height,
                    gp = gpar(col = NA, fill = fill))

          match_prefix <- row_prefix[i] == col_prefix[j]
          if (any(match_prefix)) {
            grid.rect(x = x[match_prefix], y = y[match_prefix],
                      width = width, height = height,
                      gp = gpar(col = "#000000", lwd = 0.1, fill = NA))
          }
        })

Heatmap(data_commonBed_similarity_stringent,
        top_annotation = col_anno,
        left_annotation = row_anno,
        col = col_fun,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        row_names_gp = gpar(fontsize = 6),
        column_names_gp = gpar(fontsize = 6),
        width = unit(10, "cm"), height = unit(10, "cm"),
        layer_fun = function(j, i, x, y, width, height, fill) {
          
          grid.rect(x = x, y = y, width = width, height = height,
                    gp = gpar(col = NA, fill = fill))

          match_prefix <- row_prefix[i] == col_prefix[j]
          if (any(match_prefix)) {
            grid.rect(x = x[match_prefix], y = y[match_prefix],
                      width = width, height = height,
                      gp = gpar(col = "#000000", lwd = 0.1, fill = NA))
          }
        })
```
*data_mutationCelltype_similarity_scRNA_scATAC.rds* , *data_mutationCelltype_similarity_scRNA.rds*, and *data_mutationCelltype_similarity_scATA.rds* contain the celltype mutation similarity matrices calculated based on scRNA-seq and scATAC-seq data and were generated using the functions in `0_pipelineFunctions.R` > # pipelines for generating celltype similarity (scRNA vs scATAC), # pipelines for generating celltype similarity (scRNA) and # pipelines for generating celltype similarity (scATAC)

# mutation Celltype similarity (scRNA vs scATAC)
```{r mutationCelltypeSimilarityscRNA_scATAC}
data_mutationCelltype_similarity_scRNA_scATAC <- readRDS("../Data/HumanpAML/data_mutationCelltype_similarity_scRNA_scATAC.rds")
similarity_sum_scRNA_scATAC <- Reduce("+", data_mutationCelltype_similarity_scRNA_scATAC)
similarity_mean_scRNA_scATAC <- similarity_sum_scRNA_scATAC / length(data_mutationCelltype_similarity_scRNA_scATAC)

breaks <- seq(min(as.matrix(similarity_mean_scRNA_scATAC)), 
              max(as.matrix(similarity_mean_scRNA_scATAC)), 
              length.out = 4)

col_fun <- colorRamp2(
  breaks = breaks,
  colors = c("#404040", "#bababa", "#f4a582", "#ca0020")
)

p_cellType_similarity_Mutation_scRNA_scATAC <- Heatmap(
  as.matrix(similarity_mean_scRNA_scATAC),
  name = "Similarity",
  col = col_fun,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),
  row_title = NULL,
  column_title = NULL,
  top_annotation = col_ha,
  left_annotation = row_ha,
  rect_gp = gpar(col = "white", lwd = 1)
)

p_cellType_similarity_Mutation_scRNA_scATAC
```

# mutation among celltype similarity merge all stages (scRNA)
```{r mutationCelltypeSimilarityscRNA_allStages}
data_mutationCelltype_similarity_scRNA <- readRDS("../Data/HumanpAML/data_mutationCelltype_similarity_scRNA.rds")
similarity_sum_scRNA <- Reduce("+", data_mutationCelltype_similarity_scRNA)
similarity_mean_scRNA <- similarity_sum_scRNA / length(data_mutationCelltype_similarity_scRNA)

breaks <- seq(min(as.matrix(similarity_mean_scRNA)), 
              max(as.matrix(similarity_mean_scRNA)), 
              length.out = 4)

col_fun <- colorRamp2(
  breaks = breaks,
  colors = c("#404040", "#bababa", "#f4a582", "#ca0020")
)

Heatmap(
  as.matrix(similarity_mean_scRNA),
  name = "Similarity",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),
  row_title = NULL,
  column_title = NULL,
  top_annotation = col_ha,
  left_annotation = row_ha,
  rect_gp = gpar(col = "white", lwd = 1)
)
```

# mutation among celltype similarity merge all stages (scATA)
```{r mutationCelltypeSimilarityscATA_allStages}
data_mutationCelltype_similarity_scATA <- readRDS("../Data/HumanpAML/data_mutationCelltype_similarity_scATA.rds")
similarity_sum_scATA <- Reduce("+", data_mutationCelltype_similarity_scATA)
similarity_mean_scATA <- similarity_sum_scATA / length(data_mutationCelltype_similarity_scATA)
breaks <- seq(min(as.matrix(similarity_mean_scATA)), 
              max(as.matrix(similarity_mean_scATA)), 
              length.out = 4)

col_fun <- colorRamp2(
  breaks = breaks,
  colors = c("#404040", "#bababa", "#f4a582", "#ca0020")
)
Heatmap(
  as.matrix(similarity_mean_scATA),
  name = "Similarity",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 8),
  row_title = NULL,
  column_title = NULL,
  top_annotation = col_ha,
  left_annotation = row_ha,
  rect_gp = gpar(col = "white", lwd = 1)
)
```

# lm between scRNA and scATA mutation celltype similarity
```{r lm}
mat_long_scATA <- reshape2::melt(as.matrix(similarity_mean_scATA), varnames = c("Row", "Column"), value.name = "Value")
mat_long_scRNA <- reshape2::melt(as.matrix(similarity_mean_scRNA), varnames = c("Row", "Column"), value.name = "Value")

p_correlation_celltypeMutationSimilarity <- ggplot(data = NULL, aes(x = mat_long_scATA$Value, y = mat_long_scRNA$Value)) +
  geom_point(size = 1.5, color = "#202020") +
  geom_smooth(method = "lm", se = TRUE, color = "#ca0020", linetype = "dashed", lwd = 1.2) +
  theme_classic(base_size = 16) +
  xlab("scATA celltype mutation similarity") +
  ylab("scRNA celltype mutation similarity") +
  xlim(0, 1) + ylim(0, 1) +
  annotate("text",
           x = max(mat_long_scRNA$Value) * 0.7,
           y = max(mat_long_scATA$Value) * 0.15,
           label = paste0("r = ", round(cor(mat_long_scATA$Value, mat_long_scRNA$Value), 3)),
           size = 5)

p_correlation_celltypeMutationSimilarity

cor.test(mat_long_scATA$Value, mat_long_scRNA$Value, method = "pearson")
```

*data_mutationCelltype_similarity_scRNA_DX.rds*, *data_mutationCelltype_similarity_scRNA_REM.rds* and *data_mutationCelltype_similarity_scRNA_REL.rds* contains cell type mutation similarity across different stages (see `0_pipelineFunctions.Rmd` > # pipelines for generating celltype similarity by stages(scRNA))

# mutation among celltype similarity by stages (scRNA)
```{r mutationCelltypeSimilarityscRNA_byStages}
res_scRNA_order_manu <- c("HSC", "Progenitor",
                          "GMP", "Monocytes", "CD16.Monocytes", "cDC", "pDC",
                          "CLP", "Pre.B.Cell", "B.Cell", "CD4.Naive", "CD4.Memory", "CD8.Naive", "CD8.Memory", "CD8.Effector", "NK",
                          "Early.Erythrocyte", "Late.Erythrocyte", "Early.Basophil", "Plasma")

cols <- rev(colorRampPalette(brewer.pal(11, "RdBu"))(100))

data_mutationCelltype_similarity_scRNA_DX <- readRDS("../Data/HumanpAML/data_mutationCelltype_similarity_scRNA_DX.rds")
similarity_sum_scRNA_DX <- Reduce("+", data_mutationCelltype_similarity_scRNA_DX)
similarity_mean_scRNA_DX <- similarity_sum_scRNA_DX / length(data_mutationCelltype_similarity_scRNA_DX)
corrplot(as.matrix(similarity_mean_scRNA_DX[res_scRNA_order_manu, res_scRNA_order_manu]), order = "original", method = 'color', 
         type = 'lower', col = cols, col.lim = c(0, 1))

data_mutationCelltype_similarity_scRNA_REM <- readRDS("../Data/HumanpAML/data_mutationCelltype_similarity_scRNA_REM.rds")
similarity_sum_scRNA_REM <- Reduce("+", data_mutationCelltype_similarity_scRNA_REM)
similarity_mean_scRNA_REM <- similarity_sum_scRNA_REM / length(data_mutationCelltype_similarity_scRNA_REM)
corrplot(as.matrix(similarity_mean_scRNA_REM[res_scRNA_order_manu, res_scRNA_order_manu]), order = "original", method = 'color', 
         type = 'lower', col = cols, col.lim = c(0, 1))


data_mutationCelltype_similarity_scRNA_REL <- readRDS("../Data/HumanpAML/data_mutationCelltype_similarity_scRNA_REL.rds")
similarity_sum_scRNA_REL <- Reduce("+", data_mutationCelltype_similarity_scRNA_REL)
similarity_mean_scRNA_REL <- similarity_sum_scRNA_REL / length(data_mutationCelltype_similarity_scRNA_REL)
corrplot(as.matrix(similarity_mean_scRNA_REL[res_scRNA_order_manu, res_scRNA_order_manu]), order = "original", method = 'color', 
         type = 'lower', col = cols, col.lim = c(0, 1))
```

*data_mutationCelltype_similarity_scATA_DX.rds*, *data_mutationCelltype_similarity_scATA_REM.rds* and *data_mutationCelltype_similarity_scATA_REL.rds* contains cell type mutation similarity across different stages (see `0_pipelineFunctions.Rmd` > # pipelines for generating celltype similarity by stages(scATA))

# mutation among celltype similarity by stages (scATA)
```{r mutationCelltypeSimilarityscATA_byStages}
cols <- colorRampPalette(brewer.pal(11, "RdBu"))(100)
data_mutationCelltype_similarity_scATA_DX <- readRDS("../Data/HumanpAML/data_mutationCelltype_similarity_scATA_DX.rds")
similarity_sum_scATA_DX <- Reduce("+", data_mutationCelltype_similarity_scATA_DX)
similarity_mean_scATA_DX <- similarity_sum_scATA_DX / length(data_mutationCelltype_similarity_scATA_DX)
corrplot(as.matrix(similarity_mean_scATA_DX[res_scRNA_order_manu, res_scRNA_order_manu]), order = "original", method = 'color', 
         type = 'upper', col = cols, col.lim = c(0, 1))

data_mutationCelltype_similarity_scATA_REM <- readRDS("../Data/HumanpAML/data_mutationCelltype_similarity_scATA_REM.rds")
similarity_sum_scATA_REM <- Reduce("+", data_mutationCelltype_similarity_scATA_REM)
similarity_mean_scATA_REM <- similarity_sum_scATA_REM / length(data_mutationCelltype_similarity_scATA_REM)
corrplot(as.matrix(similarity_mean_scATA_REM[res_scRNA_order_manu, res_scRNA_order_manu]), order = "original", method = 'color', 
         type = 'upper', col = cols, col.lim = c(0, 1))

data_mutationCelltype_similarity_scATA_REL <- readRDS("../Data/HumanpAML/data_mutationCelltype_similarity_scATA_REL.rds")
similarity_sum_scATA_REL <- Reduce("+", data_mutationCelltype_similarity_scATA_REL)
similarity_mean_scATA_REL <- similarity_sum_scATA_REL / length(data_mutationCelltype_similarity_scATA_REL)
corrplot(as.matrix(similarity_mean_scATA_REL[res_scRNA_order_manu, res_scRNA_order_manu]), order = "original", method = 'color', 
         type = 'upper', col = cols, col.lim = c(0, 1))
```

*data_AML_scATA_mutpeakgene.rds* and *data_AML_scRNA_mutgeneExp.rds* contain the mutation peak-gene associations from scATAC-seq data and mutation gene expression from scRNA-seq data respectively and were generated using the functions in `0_pipelineFunctions.R` > # pipelines for generating mutgeneExp and # pipelines for generating mutpeakgenedata

# stage diff mutations
```{r stageDiffMutations}
AML_ATA_mutPeakGene <- readRDS("../Data/HumanpAML/data_AML_scATA_mutpeakgene.rds")
AML_RNA_mutExprGene <- readRDS("../Data/HumanpAML/data_AML_scRNA_mutgeneExp.rds")

HSPC = c("HSC", "Progenitor")
Myeloid = c("GMP", "Monocytes", "CD16.Monocytes", "cDC", "pDC")
Lymphoid = c("CLP", "Pre.B.Cell", "B.Cell", "CD4.Memory", "CD4.Naive", "CD8.Effector", "CD8.Memory", "CD8.Naive", "NK")
Eryth = c("Early.Erythrocyte", "Late.Erythrocyte")

groups <- list(HSPC = HSPC, Myeloid = Myeloid, Lymphoid = Lymphoid, Eryth = Eryth)

AML_ATA_mutPeakGene$mainCelltype <- sapply(AML_ATA_mutPeakGene$Classified_Celltype, function(ct) {
  mainCelltype <- "Other"
  for (group_name in names(groups)) {
    if (ct %in% groups[[group_name]]) {
      mainCelltype <- group_name
      break
    }
  }
  return(mainCelltype)
})

AML_RNA_mutExprGene$mainCelltype <- sapply(AML_RNA_mutExprGene$Classified_Celltype, function(ct) {
  mainCelltype <- "Other"
  for (group_name in names(groups)) {
    if (ct %in% groups[[group_name]]) {
      mainCelltype <- group_name
      break
    }
  }
  return(mainCelltype)
})

run_fisher_stage <- function(df) {

  # pivot_wider generate wide format mut_status_col
  df_wide <- df %>%
    tidyr::pivot_wider(
      names_from = mut_status_col,
      values_from = cell_count,
      values_fill = 0
    )

  # exclude stage_id_col and mut_id_col
  status_cols <- setdiff(colnames(df_wide), c("stage_id_col", "mut_id_col"))

  # add missing column if only one status present
  if (length(status_cols) == 1) {
    missing_col <- ifelse(status_cols == "Mut", "WT", "Mut")
    df_wide[[missing_col]] <- 0
    status_cols <- c(status_cols, missing_col)
  }

  # rename status columns to "Mut" and "WT"
  names(df_wide)[match(status_cols, names(df_wide))] <- c("Mut", "WT")

  stages <- unique(df_wide$stage_id_col)

  if (length(stages) < 2) {
    return(tibble(
      stage1 = character(0),
      stage2 = character(0),
      p.value = numeric(0),
      log2_OR = numeric(0),
      estimate = numeric(0),
      table = character(0)
    ))
  }

  purrr::map_df(combn(stages, 2, simplify = FALSE), function(stg) {

    d1 <- dplyr::filter(df_wide, stage_id_col == stg[1])
    d2 <- dplyr::filter(df_wide, stage_id_col == stg[2])

    mut1 <- ifelse(length(d1$Mut) == 0, 0, d1$Mut)
    wt1  <- ifelse(length(d1$WT)  == 0, 0, d1$WT)
    mut2 <- ifelse(length(d2$Mut) == 0, 0, d2$Mut)
    wt2  <- ifelse(length(d2$WT)  == 0, 0, d2$WT)
    
    frac1 <- ifelse((mut1 + wt1) == 0, NA, mut1 / (mut1 + wt1))
    frac2 <- ifelse((mut2 + wt2) == 0, NA, mut2 / (mut2 + wt2))
    
    tab <- matrix(c(mut1, wt1, mut2, wt2), nrow = 2, byrow = TRUE)

    # if all zero, return NA
    if (sum(tab) == 0) {
      return(tibble(
        stage1 = stg[1],
        stage2 = stg[2],
        p.value = NA,
        log2_OR = NA,
        estimate = NA,
        frac_stage1 = NA,
        frac_stage2 = NA,
        table = paste(tab, collapse = ",")
      ))
    }

    ft <- fisher.test(tab)

    tibble(
      stage1 = stg[1],
      stage2 = stg[2],
      p.value = ft$p.value,
      log2_OR = log2(ft$estimate),
      estimate = ft$estimate,
      frac_stage1 = frac1,
      frac_stage2 = frac2,
      table = paste(tab, collapse = ",")
    )
  })
}

AML_scRNA_stageMutCount <- AML_RNA_mutExprGene %>%
  dplyr::group_by(mut_id_col, stage_id_col, mut_status_col, mainCelltype) %>%
  dplyr::summarise(cell_count = n(), .groups = "drop")

AML_scATA_stageMutCount <- AML_ATA_mutPeakGene %>%
  dplyr::group_by(mut_id_col, stage_id_col, mut_status_col, mainCelltype) %>%
  dplyr::summarise(cell_count = n(), .groups = "drop")

res_scRNA <- AML_scRNA_stageMutCount %>%
  dplyr::group_by(mut_id_col, mainCelltype) %>%
  dplyr::group_modify(~ run_fisher_stage(.x)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    Tag = paste0(
      gsub(".*_", "", stage1),
      "_vs_",
      gsub(".*_", "", stage2)
    )
  )

res_scATA <- AML_scATA_stageMutCount %>%
  dplyr::group_by(mut_id_col, mainCelltype) %>%
  dplyr::group_modify(~ run_fisher_stage(.x)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    Tag = paste0(
      gsub(".*_", "", stage1),
      "_vs_",
      gsub(".*_", "", stage2)
    )
  )

# GO analysis
# Dx_vs_REM
muts_scRNA_sig_DX_vs_REM <- res_scRNA %>% dplyr::filter(Tag %in% c("DX_vs_REM") & p.value <= 0.05) %>% dplyr::pull(mut_id_col) %>% unique()
gene_scRNA_sig_Dx_vs_REM <- AML_RNA_mutExprGene %>%
  dplyr::filter(mut_id_col %in% muts_scRNA_sig_DX_vs_REM) %>%
  dplyr::pull(geneSymbolMut) %>%
  unique()

muts_scATA_sig_DX_vs_REM <- res_scATA %>% dplyr::filter(Tag %in% c("DX_vs_REM") & p.value <= 0.05) %>% dplyr::pull(mut_id_col) %>% unique()
gene_scATA_sig_DX_vs_REM <- AML_ATA_mutPeakGene %>%
  dplyr::filter(mut_id_col %in% muts_scATA_sig_DX_vs_REM) %>%
  dplyr::pull(geneSymbol) %>%
  unique()

GO_RNA_enrich_DX_vs_REM <- enrichGO(gene = gene_scRNA_sig_Dx_vs_REM,
                          OrgDb = org.Hs.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.5)
GO_RNA_enrich_DX_vs_REM <- as.data.frame(GO_RNA_enrich_DX_vs_REM)

GO_ATA_enrich_DX_vs_REM <- enrichGO(gene = gene_scATA_sig_DX_vs_REM,
                          OrgDb = org.Hs.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.5)
GO_ATA_enrich_DX_vs_REM <- as.data.frame(GO_ATA_enrich_DX_vs_REM)

# REM_vs_REL
muts_scRNA_sig_REM_vs_REL <- res_scRNA %>% dplyr::filter(Tag %in% c("REL_vs_REM") & p.value <= 0.05) %>% dplyr::pull(mut_id_col) %>% unique()
gene_scRNA_sig_REM_vs_REL <- AML_RNA_mutExprGene %>%
  dplyr::filter(mut_id_col %in% muts_scRNA_sig_REM_vs_REL) %>%
  dplyr::pull(geneSymbolMut) %>%
  unique()

muts_scATA_sig_REM_vs_REL <- res_scATA %>% dplyr::filter(Tag %in% c("REL_vs_REM") & p.value <= 0.05) %>% dplyr::pull(mut_id_col) %>% unique()
gene_scATA_sig_REM_vs_REL <- AML_ATA_mutPeakGene %>%
  dplyr::filter(mut_id_col %in% muts_scATA_sig_REM_vs_REL) %>%
  dplyr::pull(geneSymbol) %>%
  unique()

GO_RNA_enrich_REM_vs_REL <- enrichGO(gene = gene_scRNA_sig_REM_vs_REL,
                          OrgDb = org.Hs.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.5)
GO_RNA_enrich_REM_vs_REL <- as.data.frame(GO_RNA_enrich_REM_vs_REL)

GO_ATA_enrich_REM_vs_REL <- enrichGO(gene = gene_scATA_sig_REM_vs_REL,
                          OrgDb = org.Hs.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.5)
GO_ATA_enrich_REM_vs_REL <- as.data.frame(GO_ATA_enrich_REM_vs_REL)

# Dx_vs_REL
muts_scRNA_sig_DX_vs_REL <- res_scRNA %>% dplyr::filter(Tag %in% c("DX_vs_REL") & p.value <= 0.05) %>% dplyr::pull(mut_id_col) %>% unique()
gene_scRNA_sig_DX_vs_REL <- AML_RNA_mutExprGene %>%
  dplyr::filter(mut_id_col %in% muts_scRNA_sig_DX_vs_REL) %>%
  dplyr::pull(geneSymbolMut) %>%
  unique()

muts_scATA_sig_DX_vs_REL <- res_scATA %>% dplyr::filter(Tag %in% c("DX_vs_REL") & p.value <= 0.05) %>% dplyr::pull(mut_id_col) %>% unique()
gene_scATA_sig_DX_vs_REL <- AML_ATA_mutPeakGene %>%
  dplyr::filter(mut_id_col %in% muts_scATA_sig_DX_vs_REL) %>%
  dplyr::pull(geneSymbol) %>%
  unique()

GO_RNA_enrich_DX_vs_REL <- enrichGO(gene = gene_scRNA_sig_DX_vs_REL,
                          OrgDb = org.Hs.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.5)
GO_RNA_enrich_DX_vs_REL <- as.data.frame(GO_RNA_enrich_DX_vs_REL)

GO_ATA_enrich_DX_vs_REL <- enrichGO(gene = gene_scATA_sig_DX_vs_REL,
                          OrgDb = org.Hs.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.5)
GO_ATA_enrich_DX_vs_REL <- as.data.frame(GO_ATA_enrich_DX_vs_REL)

# all
muts_scRNA_sig_ALL <- res_scRNA %>% dplyr::filter(p.value <= 0.05) %>% dplyr::pull(mut_id_col) %>% unique()
gene_scRNA_sig_ALL <- AML_RNA_mutExprGene %>%
  dplyr::filter(mut_id_col %in% muts_scRNA_sig_ALL) %>%
  dplyr::pull(geneSymbolMut) %>%
  unique()

muts_scATA_sig_ALL <- res_scATA %>% dplyr::filter(p.value <= 0.05) %>% dplyr::pull(mut_id_col) %>% unique()
gene_scATA_sig_ALL <- AML_ATA_mutPeakGene %>%
  dplyr::filter(mut_id_col %in% muts_scATA_sig_ALL) %>%
  dplyr::pull(geneSymbol) %>%
  unique()

GO_RNA_enrich_ALL <- enrichGO(gene = gene_scRNA_sig_ALL,
                          OrgDb = org.Hs.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.5)
GO_RNA_enrich_ALL <- as.data.frame(GO_RNA_enrich_ALL)

GO_ATA_enrich_ALL <- enrichGO(gene = gene_scATA_sig_ALL,
                          OrgDb = org.Hs.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.5)
GO_ATA_enrich_ALL <- as.data.frame(GO_ATA_enrich_ALL)

# combine
GO_RNA_enrich_DX_vs_REM$Comparison <- "DX_vs_REM"
GO_RNA_enrich_REM_vs_REL$Comparison <- "REM_vs_REL"
GO_RNA_enrich_DX_vs_REL$Comparison <- "DX_vs_REL"
GO_RNA_enrich_ALL$Comparison <- "ALL"
GO_RNA_enrich <- rbind(GO_RNA_enrich_DX_vs_REM, GO_RNA_enrich_REM_vs_REL, GO_RNA_enrich_DX_vs_REL, GO_RNA_enrich_ALL)
GO_ATA_enrich_DX_vs_REM$Comparison <- "DX_vs_REM"
GO_ATA_enrich_REM_vs_REL$Comparison <- "REM_vs_REL"
GO_ATA_enrich_DX_vs_REL$Comparison <- "DX_vs_REL"
GO_ATA_enrich_ALL$Comparison <- "ALL"
GO_ATA_enrich <- rbind(GO_ATA_enrich_DX_vs_REM, GO_ATA_enrich_REM_vs_REL, GO_ATA_enrich_DX_vs_REL, GO_ATA_enrich_ALL)
```

*GO_RNA_enrich.csv* and *GO_ATA_enrich.csv* contain the GO enrichment results for stage differential mutations from scRNA-seq and scATAC-seq data respectively and were generated using the code in the previous chunk

# GO for stage diff mutations
```{r GOstageDiffMutations}
GO_RNA_enrich <- read.csv("../Data/HumanpAML/GO_RNA_enrich.csv")
GO_ATA_enrich <- read.csv("../Data/HumanpAML/GO_ATA_enrich.csv")
# remove duplicated GO terms
GO_ID_RNA_ex <- c("GO:0030099", "GO:0051056")
GO_ID_ATA_ex <- c("GO:0015669", "GO:0015670", "GO:0042744", "GO:0098869", "GO:0019755",
                  "GO:0042743", "GO:0044403")

GO_RNA_enrich_selectAll <- GO_RNA_enrich %>%
  dplyr::filter(Comparison == "ALL") %>%
  dplyr::filter(!ID %in% GO_ID_RNA_ex) %>%
  dplyr::arrange(pvalue) %>% 
  dplyr::slice_head(n = 10) %>%
  dplyr::mutate(
    nlog2p = -log2(pvalue),
    Description = factor(Description, levels = Description[order(nlog2p)]) 
  )

p_GO_RNA_enrich_all_top10 <- ggplot(GO_RNA_enrich_selectAll, aes(x = nlog2p, y = Description)) +
  geom_col(fill = "#F8B191") +
  theme_classic() +
  labs(x = "-log2(p-value)", y = "GO Term") +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

GO_ATA_enrich_selectAll <- GO_ATA_enrich %>%
  dplyr::filter(Comparison == "ALL") %>%
  dplyr::filter(!ID %in% GO_ID_ATA_ex) %>%
  dplyr::arrange(pvalue) %>% 
  dplyr::slice_head(n = 10) %>%
  dplyr::mutate(
    nlog2p = -log2(pvalue),
    Description = factor(Description, levels = Description[order(nlog2p)]) 
  )

p_GO_ATA_enrich_all_top10 <-ggplot(GO_ATA_enrich_selectAll, aes(x = nlog2p, y = Description)) +
  geom_col(fill = "#B9D4E2") +
  theme_classic() +
  labs(x = "-log2(p-value)", y = "GO Term") +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

# combine plot
GO_ATA_enrich_selectAll$Type <- "scATAC"
GO_RNA_enrich_selectAll$Type <- "scRNA"
GO_enrich_selectAll_combined <- rbind(GO_RNA_enrich_selectAll, GO_ATA_enrich_selectAll)
GO_enrich_selectAll_combined <- GO_enrich_selectAll_combined %>%
  arrange(Type, nlog2p) %>%
  group_by(Type) %>%
  mutate(
    Description = factor(Description, levels = unique(Description[order(nlog2p)]))
  )

p_GO_combined <- ggplot(GO_enrich_selectAll_combined,
                        aes(x = nlog2p, y = Description, fill = Type)) +
  geom_col() +
  facet_wrap(~ Type, ncol = 1, scales = "free_y") +
  theme_classic() +
  scale_fill_manual(values = c("scRNA" = "#F8B191", "scATAC" = "#B9D4E2")) +
  labs(x = "-log2(p-value)", y = "GO Term") +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "none"
  )

p_GO_RNA_enrich_all_top10
p_GO_ATA_enrich_all_top10
p_GO_combined
```

# percentage celltype per sample across stages
```{r percentageCelltypePerSampleAcrossStages}
AML_RNA_meta_merge <- AML_scRNA_meta
AML_ATA_meta_merge <- AML_scATA_meta

AML_ATA_meta_merge$SampleID <- sub("_.*", "", AML_ATA_meta_merge$Sample)

rownames(AML_ATA_meta_merge) <- sub("^AML[0-9]+\\.", "", rownames(AML_ATA_meta_merge))

colnames(AML_ATA_meta_merge) <- c("Library_ID", "TSSEnrichment", "ReadsInTSS", "ReadsInPromoter", 
                "ReadsInBlacklist", "PromoterRatio", "PassQC", "NucleosomeRatio", 
                "nMultiFrags", "nMonoFrags", "nFrags", "nDiFrags", "BlacklistRatio", 
                "DoubletScore", "DoubletEnrichment", "Clusters", "ReadsInPeaks", 
                "FRIP", "predictedCell_Un", "Classified_Celltype", 
                "predictedScore_Un", "predictedCell_RNAQun", "Lambo_et_al_ID")


# ATA
AML_ATA_celltype_percentage <- AML_ATA_meta_merge %>%
  dplyr::count(Library_ID, Classified_Celltype, name = "cell_count") %>%  
  dplyr::group_by(Library_ID) %>%                                       
  dplyr::mutate(
    total_cells = sum(cell_count),                                
    percentage  = cell_count / total_cells * 100
  ) %>%
  dplyr::ungroup()


AML_RNA_celltype_percentage <- AML_RNA_meta_merge %>%
  dplyr::count(Library_ID, Classified_Celltype, name = "cell_count") %>%  
  dplyr::group_by(Library_ID) %>%                                       
  dplyr::mutate(
    total_cells = sum(cell_count),                                
    percentage  = cell_count / total_cells * 100
  ) %>%
  dplyr::ungroup()


cellTypes_order <- c(
  "HSC", "Progenitor", "GMP", "Monocytes", "CD16.Monocytes", "cDC", "pDC",
  "CLP", "Pre.B.Cell", "B.Cell", "CD4.Memory", "CD4.Naive", "CD8.Effector", "CD8.Memory", "CD8.Naive", "NK",
  "Early.Erythrocyte", "Late.Erythrocyte", "Unkwnown"
)

AML_RNA_celltype_percentage$Classified_Celltype <- factor(
  AML_RNA_celltype_percentage$Classified_Celltype,
  levels = cellTypes_order
)

AML_ATA_celltype_percentage$Classified_Celltype <- factor(
  AML_ATA_celltype_percentage$Classified_Celltype,
  levels = cellTypes_order
)


ggplot(AML_RNA_celltype_percentage, aes(x = Classified_Celltype, y = Library_ID, fill = percentage)) +
  geom_tile() +
  scale_fill_gradient(name = "Percentage", low = "white", high = "red") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.text.y = element_text(size = 10)
  )

ggplot(AML_ATA_celltype_percentage, aes(x = Classified_Celltype, y = Library_ID, fill = percentage)) +
  geom_tile() +
  scale_fill_gradient(name = "Percentage", low = "white", high = "red") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.text.y = element_text(size = 10)
  )


# main
HSPC = c("HSC", "Progenitor")
Myeloid = c("GMP", "Monocytes", "CD16.Monocytes", "cDC", "pDC")
Lymphoid = c("CLP", "Pre.B.Cell", "B.Cell", "CD4.Memory", "CD4.Naive", "CD8.Effector", "CD8.Memory", "CD8.Naive", "NK")
Eryth = c("Early.Erythrocyte", "Late.Erythrocyte")

AML_ATA_meta_merge_main <- AML_ATA_meta_merge %>%
  mutate(
    Main_celltype = case_when(
      Classified_Celltype %in% HSPC ~ "HSPC",
      Classified_Celltype %in% Myeloid ~ "Myeloid",
      Classified_Celltype %in% Lymphoid ~ "Lymphoid",
      Classified_Celltype %in% Eryth ~ "Eryth",
      TRUE ~ "Other"  # Other
    )
  )


AML_RNA_meta_merge_main <- AML_RNA_meta_merge %>%
  mutate(
    Main_celltype = case_when(
      Classified_Celltype %in% HSPC ~ "HSPC",
      Classified_Celltype %in% Myeloid ~ "Myeloid",
      Classified_Celltype %in% Lymphoid ~ "Lymphoid",
      Classified_Celltype %in% Eryth ~ "Eryth",
      TRUE ~ "Other"
    )
  )

AML_ATA_Maincelltype_percentage <- AML_ATA_meta_merge_main %>%
  dplyr::count(Library_ID, Main_celltype, name = "cell_count") %>%  
  dplyr::group_by(Library_ID) %>%                                       
  dplyr::mutate(
    total_cells = sum(cell_count),                                
    percentage  = cell_count / total_cells * 100
  ) %>%
  dplyr::ungroup()


AML_RNA_Maincelltype_percentage <- AML_RNA_meta_merge_main %>%
  dplyr::count(Library_ID, Main_celltype, name = "cell_count") %>%  
  dplyr::group_by(Library_ID) %>%                                       
  dplyr::mutate(
    total_cells = sum(cell_count),                                
    percentage  = cell_count / total_cells * 100
  ) %>%
  dplyr::ungroup()

mainCellTypes_order <- c("HSPC", "Myeloid", "Lymphoid", "Eryth", "Other")
AML_RNA_Maincelltype_percentage$Main_celltype <- factor(
  AML_RNA_Maincelltype_percentage$Main_celltype,
  levels = mainCellTypes_order
)
AML_ATA_Maincelltype_percentage$Main_celltype <- factor(
  AML_ATA_Maincelltype_percentage$Main_celltype,
  levels = mainCellTypes_order
)

ggplot(AML_RNA_Maincelltype_percentage, aes(x = Main_celltype, y = Library_ID, fill = percentage)) +
  geom_tile() +
  scale_fill_gradient(name = "Percentage", low = "white", high = "blue") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.text.y = element_text(size = 10)
  )
ggplot(AML_ATA_Maincelltype_percentage, aes(x = Main_celltype, y = Library_ID, fill = percentage)) +
  geom_tile() +
  scale_fill_gradient(name = "Percentage", low = "white", high = "blue") +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.text.y = element_text(size = 10)
  )

# Main cell type proportion barplot
AML_RNA_Maincelltype_percentage_data <- AML_RNA_Maincelltype_percentage %>% mutate(Main_celltype = ifelse(is.na(Main_celltype), "Unknown", as.character(Main_celltype)))
AML_ATA_Maincelltype_percentage_data <- AML_ATA_Maincelltype_percentage %>% mutate(Main_celltype = ifelse(is.na(Main_celltype), "Unknown", as.character(Main_celltype)))

# keep DX and REL pair samples only
# RNA 
sample_pair_RNA <- c("AML10", "AML11", "AML12", "AML13", "AML15", "AML16", 
                     "AML17", "AML18", "AML19", "AML21", "AML22", "AML23",
                     "AML24", "AML25", "AML26", "AML27", "AML2", "AML4",
                     "AML5", "AML6", "AML8", "AML9")

sample_pair_ATA <- c("AML10", "AML11", "AML12", "AML13", "AML15", "AML16", "AML17", "AML18",
                     "AML19", "AML21", "AML22", "AML24", "AML25", "AML26",
                     "AML27", "AML2", "AML4", "AML5", "AML6", "AML8")

sample_pair_RNA_ATA <- intersect(sample_pair_RNA, sample_pair_ATA)


AML_RNA_Maincelltype_percentage_data_filtered <- AML_RNA_Maincelltype_percentage_data %>%
  dplyr::filter(gsub("_.*", "", Library_ID) %in% sample_pair_RNA_ATA) %>%
  dplyr::filter(Main_celltype != "Other") %>%
  tidyr::separate(col = Library_ID, into = c("Sample", "Stage"), sep = "_", remove = FALSE) %>%
  dplyr::filter(Stage %in% c("DX", "REL"))

AML_RNA_Maincelltype_percentage_data_filtered$Stage <- factor(
  AML_RNA_Maincelltype_percentage_data_filtered$Stage,
  levels = c("DX", "REL")
)


# ATA
sample_pair_ATA <- c("AML10", "AML11", "AML12", "AML13", "AML15", "AML16", "AML17", "AML18",
                     "AML19", "AML21", "AML22", "AML24", "AML25", "AML26",
                     "AML27", "AML2", "AML4", "AML5", "AML6", "AML8")

AML_ATA_Maincelltype_percentage_data_filtered <- AML_ATA_Maincelltype_percentage_data %>%
  dplyr::filter(gsub("_.*", "", Library_ID) %in% sample_pair_RNA_ATA) %>%
  dplyr::filter(Main_celltype != "Other") %>%
  tidyr::separate(col = Library_ID, into = c("Sample", "Stage"), sep = "_", remove = FALSE) %>%
  dplyr::filter(Stage %in% c("DX", "REL"))

AML_ATA_Maincelltype_percentage_data_filtered$Stage <- factor(
  AML_ATA_Maincelltype_percentage_data_filtered$Stage,
  levels = c("DX", "REL")
)

AML_RNA_Maincelltype_percentage_data_filtered$Main_celltype <- factor(
  AML_RNA_Maincelltype_percentage_data_filtered$Main_celltype,
  levels = c("HSPC", "Myeloid", "Lymphoid", "Eryth")
)

RNA_sample_level <- c("AML4", "AML16",  "AML11", "AML17", "AML18", "AML24", 
    "AML10", "AML12", "AML13", "AML15", "AML19", "AML21", "AML22", "AML25", 
    "AML26", "AML27", "AML2", "AML5", "AML6", "AML8")

AML_RNA_Maincelltype_percentage_data_filtered$Sample <- factor(
  AML_RNA_Maincelltype_percentage_data_filtered$Sample,
  levels = RNA_sample_level
)

AML_ATA_Maincelltype_percentage_data_filtered <-
  AML_ATA_Maincelltype_percentage_data_filtered %>%
  dplyr::add_row(
    Library_ID   = "AML11_REL",
    Sample       = "AML11",
    Stage        = "REL",
    Main_celltype= "Eryth",
    cell_count   = 0L,
    total_cells  = 3686L,
    percentage   = 0
  )

AML_ATA_Maincelltype_percentage_data_filtered$Main_celltype <- factor(
  AML_ATA_Maincelltype_percentage_data_filtered$Main_celltype,
  levels = c("HSPC", "Myeloid", "Lymphoid", "Eryth")
)

ATA_sample_level <- c("AML4",  "AML16", "AML11", "AML17", "AML18", "AML24", 
                      "AML10", "AML12", "AML13", "AML15", "AML19", "AML21", "AML22", 
                      "AML25", "AML26", "AML27", "AML2", "AML5", 
                      "AML6", "AML8")

AML_ATA_Maincelltype_percentage_data_filtered$Sample <- factor(
  AML_ATA_Maincelltype_percentage_data_filtered$Sample,
  levels = ATA_sample_level
)

myColors <- colorRampPalette(brewer.pal(11, "RdBu"))(100)

p_RNA_percentage <- ggplot(AML_RNA_Maincelltype_percentage_data_filtered,
  aes(x = Main_celltype, y = Sample, fill = percentage)) +
  geom_tile(color = "white", linewidth = 0.2) +
  facet_wrap(~ Stage) +
  scale_y_discrete(limits = RNA_sample_level) +
  scale_fill_gradientn(colors = rev(myColors)[50:100], name = "Percentage") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank())

p_ATA_percentage <- ggplot(AML_ATA_Maincelltype_percentage_data_filtered,
  aes(x = Main_celltype, y = Sample, fill = percentage)) +
  geom_tile(color = "white", linewidth = 0.2) +
  facet_wrap(~ Stage) +
  scale_y_discrete(limits = ATA_sample_level) +
  scale_fill_gradientn(colors = myColors[50:100], name = "Percentage") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank())

p_RNA_percentage
p_ATA_percentage

AML_ATA_Maincelltype_percentage_data_filtered$Tag <- paste(AML_ATA_Maincelltype_percentage_data_filtered$Library_ID, 
                                                           AML_ATA_Maincelltype_percentage_data_filtered$Main_celltype,
                                                           sep = "_")

AML_RNA_Maincelltype_percentage_data_filtered$Tag <- paste(AML_RNA_Maincelltype_percentage_data_filtered$Library_ID,
                                                          AML_RNA_Maincelltype_percentage_data_filtered$Main_celltype,
                                                          sep = "_")

# merge
merged_percentage_data <- merge(
  AML_RNA_Maincelltype_percentage_data_filtered,
  AML_ATA_Maincelltype_percentage_data_filtered,
  by = "Tag",
  suffixes = c("_RNA", "_ATA")
)

p_similarity_celltypePercentage <- ggplot(data = merged_percentage_data, 
       aes(x = percentage_ATA, y = percentage_RNA)) +
  geom_point(size = 1.5, color = "#202020") +
  geom_smooth(method = "lm", se = TRUE, color = "#ca0020", linetype = "dashed", size = 1.2) +
  theme_classic() +
  xlab("scATA celltype percentage") +
  ylab("scRNA celltype percentage") +
  xlim(0, 100) + ylim(0, 100) +
  annotate("text", x = 10, y = 90,
           label = paste0("r = ", round(cor(merged_percentage_data$percentage_ATA, merged_percentage_data$percentage_RNA), 3)),
           size = 5)

p_similarity_celltypePercentage

cor.test(merged_percentage_data$percentage_ATA, merged_percentage_data$percentage_RNA, method = "pearson")
```

*p8_RDS_GIM2_sampleID_M.rds* and *p8_RDS_MM2_sampleID_M.rds* contain the GeneIntegrationMatrix and MotifMatrix data for transcrption factor lineage analysis respectively and were generated using `0_pipelineFunctions.Rmd` > # pipelines for processing scATAC-seq data and # pipelines for trajectory analysis of scATAC-seq data

GIM, Motif and geneExp data can be accessed at zenodo ()

# transcrption factor lineage analysis
```{r TFlineageAnalysis, eval=FALSE}
plot_list_lineageSwitching <- function(GIMPath, MMPath, colOrder = NA){
  message("Function to plot lineage switching heatmaps for GIM and MM data.")
  if (all(is.na(colOrder))) {
    colOrder <- seq_len(ncol(GIM_data_mat))
  }
  
  message("Reading GIM data from: ", GIMPath)
  GIM_data <- readRDS(GIMPath)
  GIM_data_mat <- GIM_data$assayMat
  #GIM_data_mat <- pmin(pmax(GIM_data_mat, -1.5), 1.5)
  rownames(GIM_data_mat) <- sub("chr[0-9XY]+:", "", rownames(GIM_data_mat))
  colnames(GIM_data_mat) <- paste(GIM_data$assayColData$label, 1:length(GIM_data$assayColData$label), sep = "_")
  message("Reading MM data from: ", MMPath)
  MM_data <- readRDS(MMPath)
  MM_data_mat <- MM_data$assayMat
  #MM_data_mat <- pmin(pmax(MM_data_mat, -1.5), 1.5)
  rownames(MM_data_mat) <-  sub("^z:([A-Za-z0-9]+)_.*$", "\\1", rownames(MM_data_mat))
  colnames(MM_data_mat) <- paste(MM_data$assayColData$label, 1:length(MM_data$assayColData$label), sep = "_")
  
  final_row_order <- rownames(GIM_data_mat)[row_order(Heatmap(GIM_data_mat))]
  
  GIM_data_mat <- GIM_data_mat[, colOrder]
  MM_data_mat <- MM_data_mat[, colOrder]
  
  col_GIM_stage <- sub("^.*_([^_]+)_.*$", "\\1", colnames(GIM_data_mat))
  ha_col <- HeatmapAnnotation(
    SampleStage = col_GIM_stage,
    col = list(
      SampleStage = c(
        "DX"  = "#33a02c",
        "REL" = "#E31A1C"
      )
    )
  )
  
  ht_GIM_final <- Heatmap(GIM_data_mat[final_row_order,],
                 cluster_rows = FALSE,
                 cluster_columns = FALSE,
                 top_annotation = ha_col,
                 col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100))[50:100])
  
  ht_MM_final <- Heatmap(MM_data_mat[final_row_order,],
                 cluster_rows = FALSE,
                 cluster_columns = FALSE,
                 top_annotation = ha_col,
                 col = colorRampPalette(brewer.pal(11, "RdBu"))(100)[50:100])

  return(list(ht_GIM_final = ht_GIM_final,
              ht_MM_final = ht_MM_final,
              GIM_mat = GIM_data_mat,
              MM_mat = MM_data_mat,
              genes = final_row_order))
}

rowMeans_no_outlier <- function(mat) {
  apply(mat, 1, function(x) {
    q <- quantile(x, probs = c(0.25, 0.75), na.rm = TRUE)
    iqr <- q[2] - q[1]
    x[x >= (q[1] - 1.5 * iqr) & x <= (q[2] + 1.5 * iqr)] |> 
      mean(na.rm = TRUE)
  })
}


generate_marker_matrix <- function(marker_gene, matMatrix, iteration_times = 1000, iteration_sizef  = 0.5, dx_pattern  = "DX", rel_pattern = "REL"){
  matMatrix_marker <- matMatrix[marker_gene, ]
  cells_DX  <- grep(dx_pattern,  colnames(matMatrix_marker), value = TRUE)
  cells_REL <- grep(rel_pattern, colnames(matMatrix_marker), value = TRUE)
  
  set.seed(2025)
  iteration_DX <- list()
  iteration_REL <- list()
  for (i in 1:iteration_times) {
    sampled_cells_DX <- sample(cells_DX, size = round(length(cells_DX) * iteration_sizef), replace = TRUE)
    iteration_DX[[i]] <- rowMeans_no_outlier(matMatrix_marker[,sampled_cells_DX, drop = FALSE])
    sampled_cells_REL <- sample(cells_REL, size = round(length(cells_REL) * iteration_sizef), replace = TRUE)
    iteration_REL[[i]] <- rowMeans_no_outlier(matMatrix_marker[,sampled_cells_REL, drop = FALSE])
  }
  
  return(list(DX = iteration_DX,
              REL = iteration_REL))
}

range_scale <- function(x, min_val = -1.5, max_val = 1.5) {
  x_min <- min(x, na.rm = TRUE)
  x_max <- max(x, na.rm = TRUE)

  (x - x_min) / (x_max - x_min) * (max_val - min_val) + min_val
}


# paths
GIM_AML4_Path <- "../Data/HumanpAML/p8_RDS_GIM2_AML4_M.rds"
MM_AML4_Path <- "../Data/HumanpAML/p8_RDS_MM2_AML4_M.rds"
GIM_AML16_Path <- "../Data/HumanpAML/p8_RDS_GIM2_AML16_M.rds"
MM_AML16_Path <- "../Data/HumanpAML/p8_RDS_MM2_AML16_M.rds"
GIM_AML11_Path <- "../Data/HumanpAML/p8_RDS_GIM2_AML18_M.rds"
MM_AML11_Path <- "../Data/HumanpAML/p8_RDS_MM2_AML18_M.rds"
GIM_AML17_Path <- "../Data/HumanpAML/p8_RDS_GIM2_AML17_M.rds"
MM_AML17_Path <- "../Data/HumanpAML/p8_RDS_MM2_AML17_M.rds"
GIM_AML18_Path <- "../Data/HumanpAML/p8_RDS_GIM2_AML18_M.rds"
MM_AML18_Path <- "../Data/HumanpAML/p8_RDS_MM2_AML18_M.rds"
GIM_AML24_Path <- "../Data/HumanpAML/p8_RDS_GIM2_AML24_M.rds"
MM_AML24_Path <- "../Data/HumanpAML/p8_RDS_MM2_AML24_M.rds"

colOrderQ <- c(1:100)
lineage_AML4 <- plot_list_lineageSwitching(GIM_AML4_Path, MM_AML4_Path, colOrderQ)

colOrderQ <- c(1:100)
lineage_AML16 <- plot_list_lineageSwitching(GIM_AML16_Path, MM_AML16_Path, colOrderQ)

colOrderQ <- c(1:100)
lineage_AML11 <- plot_list_lineageSwitching(GIM_AML11_Path, MM_AML11_Path, colOrderQ)

colOrderQ <- c(1:100)
lineage_AML17 <- plot_list_lineageSwitching(GIM_AML17_Path, MM_AML17_Path, colOrderQ)

colOrderQ <- c(1:100)
lineage_AML18 <- plot_list_lineageSwitching(GIM_AML18_Path, MM_AML18_Path, colOrderQ)

colOrderQ <- c(1:100)
lineage_AML24 <- plot_list_lineageSwitching(GIM_AML24_Path, MM_AML24_Path, colOrderQ)

lineage_AML4$ht_GIM_final
lineage_AML4$ht_MM_final

lineage_AML16$ht_GIM_final
lineage_AML16$ht_MM_final

lineage_AML11$ht_GIM_final
lineage_AML11$ht_MM_final

lineage_AML17$ht_GIM_final
lineage_AML17$ht_MM_final

lineage_AML18$ht_GIM_final
lineage_AML18$ht_MM_final

lineage_AML24$ht_GIM_final
lineage_AML24$ht_MM_final


# select genes showing lineage switching
lineageSwitching_genes_H <-unique(c("LMO2", "RUNX1", "GATA2", "TAL1", "ERG", "FLI1",
                                    "HOXA9"))

lineageSwitching_genes_M <- unique(c("STAT2", "CREM", "SPI1", "KLF2", "BACH1", "ATF4", "IRF1", "ATF3",
    "JDP2", "NR4A1", "KLF3", "BHLHE40", "CEBPD", "NR4A3", "KLF4", "NR4A2", "DDIT3", "RXRA", "REL",
    "MAFF", "CEBPA", "NFKB1", "TFEB", "CEBPB", "KLF11", "NR1D2", "JUNB", "ETV3", "CREB5", "FOSL2",
    "ELF4", "RARA", "FOS", "KLF13", "JUN", "MLX", "NFIL3", "FOSB"))

lineageSwitching_genes_L <- unique(c("USF2", "PAX5", "POU2F1", "IRF7", "XBP1", "FOXO1", "GFI1", "TCF4",
    "FOXP1", "ETS1", "ID3", "MEF2C", "FOXN3", "IRF4", "EBF1", "TCF7L2", "IRF2", "TCFL5", "TCF3", "LEF1",
    "SMARCC2", "KLF6", "TCF12", "IRF3", "SP3"))

HSC_marker <- c("HOPX", "SPINK2")
Progenitor_marker <- c("CD34", "NUCB2", "C1QTNF4","CDK6", "IGFBP7")
GMP_Myeloid_marker <- c("PRTN3", "ELANE", "MPO", "AZU1", "CTSG")
Monocyte_Myeloid_marker <- c("TYROBP", "FCN1", "VCAN")
B_Lymphoid_marker <- c("CD79A", "CD19", "DNTT", "VPREB3")
T_Lymphoid_marker <- c("IL7R", "LTB", "CD7")

tf_genes <- c(lineageSwitching_genes_H, lineageSwitching_genes_M, lineageSwitching_genes_L)
marker_genes <- c(HSC_marker, Progenitor_marker, GMP_Myeloid_marker, Monocyte_Myeloid_marker,
                  B_Lymphoid_marker, T_Lymphoid_marker)

gene_lineage <- data.frame(
  Gene = c(lineageSwitching_genes_H, lineageSwitching_genes_M, lineageSwitching_genes_L, HSC_marker, Progenitor_marker,
           GMP_Myeloid_marker, Monocyte_Myeloid_marker, B_Lymphoid_marker, T_Lymphoid_marker),
  Lineage = c(rep("HSPC_TF", length(lineageSwitching_genes_H)),
              rep("Myeloid_TF", length(lineageSwitching_genes_M)),
              rep("Lymphoid_TF", length(lineageSwitching_genes_L)),
              rep("HSC_Marker", length(HSC_marker)),
              rep("Progenitor_Marker", length(Progenitor_marker)),
              rep("Myeloid_Marker_GMP", length(GMP_Myeloid_marker)),
              rep("Myeloid_Marker_Monocyte", length(Monocyte_Myeloid_marker)),
              rep("Lymphoid_Marker_B", length(B_Lymphoid_marker)),
              rep("Lymphoid_Marker_T", length(T_Lymphoid_marker)))
)

lineageSwitching_genes <- unique(c(lineageSwitching_genes_H, lineageSwitching_genes_M, lineageSwitching_genes_L, 
                                   HSC_marker, Progenitor_marker, GMP_Myeloid_marker,
           Monocyte_Myeloid_marker, B_Lymphoid_marker, T_Lymphoid_marker))


include_genes <- c(lineageSwitching_genes_H, "NR4A1", "CEBPA", "KLF13", "NFIL3", "PAX5", "POU2F1", "TCF4", 
                   "MEF2C", "EBF1", "TCFL5", "TCF3", "LEF1", "IRF3", "PRTN3", 
                   "ELANE", "MPO", "AZU1", "CTSG", "TYROBP", "IGLL1", "CD79A", 
                   "DNTT", "VPREB3", "IL7R", "LTB", marker_genes)

lineageSwitching_genes <- lineageSwitching_genes[which(lineageSwitching_genes %in% include_genes)]

sample_names_lineage <- c("AML8", "AML6", "AML5", "AML2", "AML27", "AML26", 
                          "AML25", "AML22", "AML21", "AML19", "AML15", "AML13", 
                          "AML12", "AML10", "AML24", "AML18", "AML17", "AML11", 
                          "AML16", "AML4")

modality_names_lineage <- c("RNA", "ATA")
stage_names_lineage <- c("DX", "REL")
iteration_times <- 100
# lineage switching gene matrix
colnames_lineageSwitching_matrix <- unlist(
  lapply(sample_names_lineage, function(sample) {
    unlist(
      lapply(modality_names_lineage, function(modality) {
        unlist(
          lapply(stage_names_lineage, function(stage) {
            paste(sample, stage, 1:iteration_times, modality, sep = "_")
          })
        )
      })
    )
  })
)

lineageSwitching_gene_matrix <- as.data.frame(
  matrix(
    NA,
    nrow = length(lineageSwitching_genes),
    ncol = length(colnames_lineageSwitching_matrix)
  )
)

rownames(lineageSwitching_gene_matrix) <- lineageSwitching_genes
colnames(lineageSwitching_gene_matrix) <- colnames_lineageSwitching_matrix

for (sample in sample_names_lineage) {
  ATA_motif_path <- paste0("../Data/HumanpAML/", sample, "_cellMotif.rds")
  RNA_motif_path <- paste0("../Data/HumanpAML/", sample, "_geneExp.rds")
  ATA_motif <- readRDS(ATA_motif_path)
  rownames(ATA_motif) <- sub("_.*$", "", rownames(ATA_motif))
  RNA_exp <- readRDS(RNA_motif_path)
  for (gene in lineageSwitching_genes) {
    message("Processing gene: ", gene, " in sample: ", sample)
    # ATA
    message("  Processing ATA data...")
    res_ATA <- generate_marker_matrix(gene, ATA_motif, iteration_times = iteration_times, iteration_sizef  = 0.5, dx_pattern  = "DX", rel_pattern = "REL")
    col_ATA <- paste(sample, c(paste(rep(stage_names_lineage[1], iteration_times), 1:iteration_times, sep = "_"), paste(rep(stage_names_lineage[2], iteration_times), 1:iteration_times, sep = "_")), "ATA", sep = "_")
    col_ATA_DX <- grep("DX", col_ATA, value = TRUE)
    col_ATA_REL <- grep("REL", col_ATA, value = TRUE)
    ATA_value <- c(unlist(res_ATA$DX), unlist(res_ATA$REL))
    ATA_value_scaled <- range_scale(ATA_value, min_val = -1.5, max_val = 1.5)
    lineageSwitching_gene_matrix[gene, col_ATA_DX] <- ATA_value_scaled[1:iteration_times]
    lineageSwitching_gene_matrix[gene, col_ATA_REL] <- ATA_value_scaled[(iteration_times + 1):(2 * iteration_times)]
    # RNA
    message("  Processing RNA data...")
    res_RNA <- generate_marker_matrix(gene, RNA_exp, iteration_times = iteration_times, iteration_sizef  = 0.5, dx_pattern  = "DX", rel_pattern = "REL")
    col_RNA <- paste(sample, c(paste(rep(stage_names_lineage[1], iteration_times), 1:iteration_times, sep = "_"), paste(rep(stage_names_lineage[2], iteration_times), 1:iteration_times, sep = "_")), "RNA", sep = "_")
    col_RNA_DX <- grep("DX", col_RNA, value = TRUE)
    col_RNA_REL <- grep("REL", col_RNA, value = TRUE)
    RNA_value <- c(unlist(res_RNA$DX), unlist(res_RNA$REL))
    RNA_value_scaled <- range_scale(RNA_value, min_val = -1.5, max_val = 1.5)
    lineageSwitching_gene_matrix[gene, col_RNA_DX] <- RNA_value_scaled[1:iteration_times]
    lineageSwitching_gene_matrix[gene, col_RNA_REL] <- RNA_value_scaled[(iteration_times + 1):(2 * iteration_times)]
  }
}

lineageSwitching_gene_matrix_rmna <- na.omit(lineageSwitching_gene_matrix)

gene_lineage <- gene_lineage[gene_lineage$Gene %in% include_genes, ]

ha_row <- rowAnnotation(
  Lineage = gene_lineage$Lineage,
  col = list(
    Lineage = c(
      HSPC_TF  = "#662506",
      HSC_Marker = "#cc4c02",
      Progenitor_Marker = "#fec44f",
      Myeloid_TF  = "#053061",
      Myeloid_Marker_GMP = "#2166AC",
      Myeloid_Marker_Monocyte = "#4393C3",
      Lymphoid_TF = "#00441b",
      Lymphoid_Marker_B = "#238b45",
      Lymphoid_Marker_T = "#66c2a4"
    )
  )
)

Heatmap(lineageSwitching_gene_matrix,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        left_annotation = ha_row,
        col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100))[50:100])

#saveRDS(lineageSwitching_gene_matrix, "../Data/AML/AML_lineageSwitching_gene_matrix.rds")
#saveRDS(gene_lineage, "../Data/AML/AML_lineageSwitching_gene_meta.rds")
```

*AML_lineageSwitching_gene_matrix.rds* and *AML_lineageSwitching_gene_meta.rds* contain the lineage switching gene matrix and gene lineage metadata used for plotting transcription factor lineage plots and were generated using the code above.

# transcription factor lineage plots
```{r TFlineagePlots}
lineageSwitching_gene_matrix <- readRDS("../Data/HumanpAML/AML_lineageSwitching_gene_matrix.rds")
gene_lineage <- readRDS("../Data/HumanpAML/AML_lineageSwitching_gene_meta.rds")
gene_lineage_tf <- gene_lineage[gene_lineage$Lineage %in% c("Myeloid_TF", "Lymphoid_TF"),]
gene_lineage_marker <- gene_lineage[gene_lineage$Lineage %in% c("HSC_Marker", "Progenitor_Marker",
                                                            "Myeloid_Marker_GMP", "Myeloid_Marker_Monocyte",
                                                            "Lymphoid_Marker_B", "Lymphoid_Marker_T"),]

gene_tf <- gene_lineage$Gene[gene_lineage$Lineage %in% c("Myeloid_TF", "Lymphoid_TF")]
gene_marker <- gene_lineage$Gene[gene_lineage$Lineage %in% c("HSC_Marker", "Progenitor_Marker",
                                                            "Myeloid_Marker_GMP", "Myeloid_Marker_Monocyte",
                                                            "Lymphoid_Marker_B", "Lymphoid_Marker_T")]

lineageSwitching_gene_matrix_tf <- lineageSwitching_gene_matrix[rownames(lineageSwitching_gene_matrix) %in% gene_tf, ]
lineageSwitching_gene_matrix_tf_RNA <- lineageSwitching_gene_matrix_tf[, grep("RNA", colnames(lineageSwitching_gene_matrix_tf))]
lineageSwitching_gene_matrix_tf_ATA <- lineageSwitching_gene_matrix_tf[, grep("ATA", colnames(lineageSwitching_gene_matrix_tf))]

lineageSwitching_gene_matrix_marker <- lineageSwitching_gene_matrix[rownames(lineageSwitching_gene_matrix) %in% gene_marker, ]
lineageSwitching_gene_matrix_marker_RNA <- lineageSwitching_gene_matrix_marker[, grep("RNA", colnames(lineageSwitching_gene_matrix_marker))]

ha_row_tf <- rowAnnotation(
  Lineage = gene_lineage_tf$Lineage,
  col = list(
    Lineage = c(
      Myeloid_TF  = "#053061",
      Lymphoid_TF = "#00441b"
    )
  )
)

ha_row_marker <- rowAnnotation(
  Lineage = gene_lineage_marker$Lineage,
  col = list(
    Lineage = c(
      HSC_Marker = "#7f2704",
      Progenitor_Marker = "#d94801",
      Myeloid_Marker_GMP = "#2166AC",
      Myeloid_Marker_Monocyte = "#4393C3",
      Lymphoid_Marker_B = "#238b45",
      Lymphoid_Marker_T = "#66c2a4"
    )
  )
)

TF_motif_order_byletter <- c("CEBPA", "KLF13", "NFIL3", "NR4A1", "EBF1", "IRF3",
                             "LEF1", "MEF2C", "PAX5", "POU2F1", "TCF3", "TCF4", "TCFL5")

marker_gene_order_byletter <- c("C1QTNF4", "CD34", "CDK6", "HOPX", "IGFBP7", "NUCB2", "SPINK2", 
                                "AZU1", "CTSG", "ELANE", "FCN1", "MPO", "PRTN3", "TYROBP", "VCAN", 
                                "CD19", "CD79A", "DNTT", "IL7R", "LTB", "VPREB3", "CD7")

# TF_EXP
Heatmap(lineageSwitching_gene_matrix_tf_RNA[TF_motif_order_byletter,],
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        left_annotation = ha_row_tf,
        col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100))[50:100])

# TF_ATA
Heatmap(lineageSwitching_gene_matrix_tf_ATA[TF_motif_order_byletter, ],
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        left_annotation = ha_row_tf,
        col = colorRampPalette(brewer.pal(11, "RdBu"))(100)[50:100])


# Marker_EXP
Heatmap(lineageSwitching_gene_matrix_marker_RNA[marker_gene_order_byletter, ],
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        left_annotation = ha_row_marker,
        col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100))[50:100])
```
# merge figures for TF lineage switching
```{r mergeTFLineageSwitching}
iteration_times <- 100
lineageSwitching_gene_matrix <- readRDS("../Data/HumanpAML/AML_lineageSwitching_gene_matrix.rds")
gene_lineage <- readRDS("../Data/HumanpAML/AML_lineageSwitching_gene_meta.rds")
gene_lineage_tf <- gene_lineage[gene_lineage$Lineage %in% c("Myeloid_TF", "Lymphoid_TF"),]
gene_lineage_marker <- gene_lineage[gene_lineage$Lineage %in% c("HSC_Marker", "Progenitor_Marker",
                                                            "Myeloid_Marker_GMP", "Myeloid_Marker_Monocyte",
                                                            "Lymphoid_Marker_B", "Lymphoid_Marker_T"),]

gene_tf <- gene_lineage$Gene[gene_lineage$Lineage %in% c("Myeloid_TF", "Lymphoid_TF")]
gene_marker <- gene_lineage$Gene[gene_lineage$Lineage %in% c("HSC_Marker", "Progenitor_Marker",
                                                            "Myeloid_Marker_GMP", "Myeloid_Marker_Monocyte",
                                                            "Lymphoid_Marker_B", "Lymphoid_Marker_T")]

lineageSwitching_gene_matrix_tf <- lineageSwitching_gene_matrix[rownames(lineageSwitching_gene_matrix) %in% gene_tf, ]
lineageSwitching_gene_matrix_tf_RNA <- lineageSwitching_gene_matrix_tf[, grep("RNA", colnames(lineageSwitching_gene_matrix_tf))]
lineageSwitching_gene_matrix_tf_ATA <- lineageSwitching_gene_matrix_tf[, grep("ATA", colnames(lineageSwitching_gene_matrix_tf))]

lineageSwitching_gene_matrix_marker <- lineageSwitching_gene_matrix[rownames(lineageSwitching_gene_matrix) %in% gene_marker, ]
lineageSwitching_gene_matrix_marker_RNA <- lineageSwitching_gene_matrix_marker[, grep("RNA", colnames(lineageSwitching_gene_matrix_marker))]

# --------
lineageSwitching_gene_matrix_MEF2C_CEBPA <- lineageSwitching_gene_matrix[rownames(lineageSwitching_gene_matrix) %in% c("MEF2C", "CEBPA"),]
col_info <- data.frame(col = colnames(lineageSwitching_gene_matrix_MEF2C_CEBPA))

col_info$sample <- sub("_(DX|REL)_.*", "", col_info$col)
col_info$stage  <- sub(".*_(DX|REL)_.*", "\\1", col_info$col)

col_info$stage <- factor(col_info$stage, levels = c("DX", "REL"))
col_info <- col_info[order(col_info$sample, col_info$stage), ]
lineageSwitching_gene_matrix_MEF2C_CEBPA_sorted <- lineageSwitching_gene_matrix_MEF2C_CEBPA[, col_info$col]

colnames_df <- data.frame(col = colnames(lineageSwitching_gene_matrix_MEF2C_CEBPA_sorted))

colnames_df$sample <- sub("_(DX|REL)_.*", "", colnames_df$col)
colnames_df$stage  <- sub(".*_(DX|REL)_.*", "\\1", colnames_df$col)
colnames_df$mod    <- sub(".*_(RNA|ATA)$", "\\1", colnames_df$col)
colnames_df$iter   <- as.numeric(sub(".*_(DX|REL)_([0-9]+)_.*", "\\2", colnames_df$col))

rownames(colnames_df) <- colnames_df$col
ann_colors <- list(
  stage = c(DX = "#1f78b4", REL = "#e31a1c"),
  mod   = c(RNA = "#33a02c", ATA = "#ff7f00")
)

pheatmap(
  lineageSwitching_gene_matrix_MEF2C_CEBPA_sorted,
  cluster_cols = FALSE,
  cluster_rows = TRUE,
  annotation_col = colnames_df[, c("sample", "stage", "mod")],
  annotation_colors = ann_colors,
  show_colnames = FALSE,
  scale = "row"
)

samples_unique <- unique(colnames_df$sample)
gene_mef2c_cebpa <- c("MEF2C", "CEBPA")
allStages <- c("DX", "REL")
colnames_sample_gene_dxrel <- c(paste0("MEF2C", "_DX_", 1:iteration_times, "_RNA"), paste0("MEF2C", "_REL_", 1:iteration_times, "_RNA"),
                                paste0("MEF2C", "_DX_", 1:iteration_times, "_ATA"), paste0("MEF2C", "_REL_", 1:iteration_times, "_ATA"),
                                paste0("CEBPA", "_DX_", 1:iteration_times, "_RNA"), paste0("CEBPA", "_REL_", 1:iteration_times, "_RNA"),
                                paste0("CEBPA", "_DX_", 1:iteration_times, "_ATA"), paste0("CEBPA", "_REL_", 1:iteration_times, "_ATA"))
sample_gene_dxrel <- as.data.frame(matrix(data = 0, nrow = length(samples_unique), ncol = length(colnames_sample_gene_dxrel)))
rownames(sample_gene_dxrel) <- samples_unique 
colnames(sample_gene_dxrel) <- colnames_sample_gene_dxrel

for (stage in allStages){
  for (sample in samples_unique){
    for (gene in gene_mef2c_cebpa){
      for (iter_n in 1:iteration_times){
        # ori
        col_ori_RNA <- paste(sample, stage, iter_n, "RNA", sep = "_")
        col_ori_ATA <- paste(sample, stage, iter_n, "ATA", sep = "_")
        print(col_ori_RNA)
        item_ori_RNA <- lineageSwitching_gene_matrix_MEF2C_CEBPA_sorted[gene, col_ori_RNA]
        item_ori_ATA <- lineageSwitching_gene_matrix_MEF2C_CEBPA_sorted[gene, col_ori_ATA]
        # new
        col_new_RNA <- paste(gene, stage, iter_n, "RNA", sep = "_")
        col_new_ATA <- paste(gene, stage, iter_n, "ATA", sep = "_")
        sample_gene_dxrel[sample, col_new_RNA] <- item_ori_RNA
        sample_gene_dxrel[sample, col_new_ATA] <- item_ori_ATA
      }
    }
  }
}

sample_gene_dxrel_RNA <- sample_gene_dxrel[,grep("RNA", colnames(sample_gene_dxrel), value = T)]
sample_gene_dxrel_ATA <- sample_gene_dxrel[,grep("ATA", colnames(sample_gene_dxrel), value = T)]
samples_order <- c("AML24", "AML19", "AML11", "AML17", "AML18", "AML2", "AML15",
                   "AML5", "AML10", "AML12", "AML27", "AML6", "AML4", "AML8",
                   "AML16", "AML21", "AML22", "AML25", "AML26", "AML13")

Heatmap(sample_gene_dxrel,
        cluster_columns = FALSE,
        col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100))[50:100])


Heatmap(sample_gene_dxrel_RNA[samples_order,],
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        #left_annotation = ha_row_tf,
        col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100))[50:100])

Heatmap(sample_gene_dxrel_ATA[samples_order,],
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        #left_annotation = ha_row_tf,
        col = colorRampPalette(brewer.pal(11, "RdBu"))(100)[50:100])
# --------

ha_row_tf <- rowAnnotation(
  Lineage = gene_lineage_tf$Lineage,
  col = list(
    Lineage = c(
      Myeloid_TF  = "#053061",
      Lymphoid_TF = "#00441b"
    )
  )
)

ha_row_marker <- rowAnnotation(
  Lineage = gene_lineage_marker$Lineage,
  col = list(
    Lineage = c(
      HSC_Marker = "#7f2704",
      Progenitor_Marker = "#d94801",
      Myeloid_Marker_GMP = "#2166AC",
      Myeloid_Marker_Monocyte = "#4393C3",
      Lymphoid_Marker_B = "#238b45",
      Lymphoid_Marker_T = "#66c2a4"
    )
  )
)

TF_motif_order_byletter <- c("CEBPA", "KLF13", "NFIL3", "NR4A1", "EBF1", "IRF3",
                             "LEF1", "MEF2C", "PAX5", "POU2F1", "TCF3", "TCF4", "TCFL5")

marker_gene_order_byletter <- c("C1QTNF4", "CD34", "CDK6", "HOPX", "IGFBP7", "NUCB2", "SPINK2", 
                                "AZU1", "CTSG", "ELANE", "FCN1", "MPO", "PRTN3", "TYROBP", "VCAN", 
                                "CD19", "CD79A", "DNTT", "IL7R", "LTB", "VPREB3", "CD7")

# TF_EXP
Heatmap(lineageSwitching_gene_matrix_tf_RNA[TF_motif_order_byletter,],
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        left_annotation = ha_row_tf,
        col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100))[50:100])

# TF_ATA
Heatmap(lineageSwitching_gene_matrix_tf_ATA[TF_motif_order_byletter, ],
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        left_annotation = ha_row_tf,
        col = colorRampPalette(brewer.pal(11, "RdBu"))(100)[50:100])


# Marker_EXP
Heatmap(lineageSwitching_gene_matrix_marker_RNA[marker_gene_order_byletter, ],
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        left_annotation = ha_row_marker,
        col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100))[50:100])
```

# mutation Similarity DXREL for lineage switching
```{r mutationSimilarityDXREL}
get_scRNA_mutSites_byStages <- function(CSM_list, meta_data){
  common_sample_pairs <- unique(names(CSM_list))
  ct_forComparision <- unique(meta_data$Classified_Celltype)
  ct_forComparision <- setdiff(ct_forComparision, "Unknown")
  print(ct_forComparision)
  data_mutationCelltype_sites_list <- list()
  for (sample in common_sample_pairs){
    message(sample)
    csm_tmp <- CSM_list[[sample]]
    for (i in ct_forComparision){
      ct_i <- i
      cell_barcode_meta_i <- meta_data %>% 
        mutate(CellTag = rownames(.)) %>%  
        dplyr::filter(Classified_Celltype == ct_i & Library_ID == sample) %>%
        pull(CellTag)
      cell_barcode_csm_format_all_i <- rownames(csm_tmp) %>%
        sub("_L[0-9]+_", "_", .) %>%
        sub("_(.*)_", "_\\1.", .) %>%
        paste0(.,"-1") %>%
        sub("\\.", "#", x = .)
      cell_barcode_csm_selected_i <- rownames(csm_tmp)[cell_barcode_csm_format_all_i %in% cell_barcode_meta_i]
      mut_site_i <- retriveLocsByCells(csm_tmp, cell_barcode_csm_selected_i)$Mut
      if (length(cell_barcode_meta_i) <= 10){
        mut_site_i <- "Empty"
      }
      data_mutationCelltype_sites_list[[sample]][[ct_i]] <- mut_site_i
    }
  }
  return(data_mutationCelltype_sites_list)
}

get_scATA_mutSites_byStages <- function(CSM_list, meta_data){
  common_sample_pairs <- unique(names(CSM_list))
  ct_forComparision <- unique(meta_data$Classified_Celltype)
  ct_forComparision <- setdiff(ct_forComparision, "Unknown")
  print(ct_forComparision)
  data_mutationCelltype_sites_list <- list()
  for (sample in common_sample_pairs){
    message(sample)
    csm_tmp <- CSM_list[[sample]]
    for (i in ct_forComparision){
      ct_i <- i
      cell_barcode_meta_i <- meta_data %>% 
        mutate(CellTag = rownames(.)) %>%  
        dplyr::filter(Classified_Celltype == ct_i & Sample == sample) %>%
        pull(CellTag)
      cell_barcode_csm_format_all_i <- rownames(csm_tmp) %>%
        sub("_L[0-9]+_", "_", .) %>%
        sub("_(.*)_", "_\\1.", .) %>%
        paste0(.,"-1") %>%
        sub("\\.", "#", x = .)
      cell_barcode_csm_selected_i <- rownames(csm_tmp)[cell_barcode_csm_format_all_i %in% cell_barcode_meta_i]
      mut_site_i <- retriveLocsByCells(csm_tmp, cell_barcode_csm_selected_i)$Mut
      if (length(cell_barcode_meta_i) <= 10){
        mut_site_i <- "Empty"
      }
      data_mutationCelltype_sites_list[[sample]][[ct_i]] <- mut_site_i
    }
  }
  return(data_mutationCelltype_sites_list)
}

AML_scRNA_CSM_list <- AML_RNA_CSM_list
AML_scATA_CSM_list <- AML_ATA_CSM_list

# scRNA
# AML_CSM_DX_list
AML_scRNA_CSM_DX_list <- lapply(AML_scRNA_CSM_list, function(df) {
  df[grep("DX", rownames(df)), , drop = FALSE]
})
AML_scRNA_CSM_DX_list <- Filter(function(x) nrow(x) > 0, AML_scRNA_CSM_DX_list)

# AML_CSM_REM_list
AML_scRNA_CSM_REM_list <- lapply(AML_scRNA_CSM_list, function(df) {
  df[grep("REM", rownames(df)), , drop = FALSE]
})
AML_scRNA_CSM_REM_list <- Filter(function(x) nrow(x) > 0, AML_scRNA_CSM_REM_list)

# AML_CSM_REL_list
AML_scRNA_CSM_REL_list <- lapply(AML_scRNA_CSM_list, function(df) {
  df[grep("REL", rownames(df)), , drop = FALSE]
})
AML_scRNA_CSM_REL_list <- Filter(function(x) nrow(x) > 0, AML_scRNA_CSM_REL_list)

# scATA
# AML_CSM_DX_list
AML_scATA_CSM_DX_list <- lapply(AML_scATA_CSM_list, function(df) {
  df[grep("DX", rownames(df)), , drop = FALSE]
})
AML_scATA_CSM_DX_list <- Filter(function(x) nrow(x) > 0, AML_scATA_CSM_DX_list)

# AML_CSM_REM_list
AML_scATA_CSM_REM_list <- lapply(AML_scATA_CSM_list, function(df) {
  df[grep("REM", rownames(df)), , drop = FALSE]
})
AML_scATA_CSM_REM_list <- Filter(function(x) nrow(x) > 0, AML_scATA_CSM_REM_list)

# AML_CSM_REL_list
AML_scATA_CSM_REL_list <- lapply(AML_scATA_CSM_list, function(df) {
  df[grep("REL", rownames(df)), , drop = FALSE]
})
AML_scATA_CSM_REL_list <- Filter(function(x) nrow(x) > 0, AML_scATA_CSM_REL_list)

HSPC <- c("HSC", "Progenitor")
Myeloid <- c("GMP", "Monocytes", "CD16.Monocytes", "cDC", "pDC")
Lymphoid <- c("CLP", "Pre.B.Cell", "B.Cell", "CD4.Naive", "CD4.Memory", "CD8.Naive", "CD8.Memory", "CD8.Effector", "NK")
Erythroid_Basophil <- c("Early.Erythrocyte", "Late.Erythrocyte")
Other <- c("Early.Basophil", "Plasma")

AML_scRNA_meta <- AML_scRNA_meta %>%
  dplyr::mutate(
    maincelltype = case_when(
      Classified_Celltype %in% HSPC ~ "HSPC",
      Classified_Celltype %in% Myeloid ~ "Myeloid",
      Classified_Celltype %in% Lymphoid ~ "Lymphoid",
      Classified_Celltype %in% Erythroid_Basophil ~ "Erythroid",
      Classified_Celltype %in% Other ~ "Other",
      TRUE ~ NA_character_
    )
  )

AML_scATA_meta <- AML_scATA_meta %>%
  dplyr::mutate(
    maincelltype = case_when(
      predictedGroup_Un %in% HSPC ~ "HSPC",
      predictedGroup_Un %in% Myeloid ~ "Myeloid",
      predictedGroup_Un %in% Lymphoid ~ "Lymphoid",
      predictedGroup_Un %in% Erythroid_Basophil ~ "Erythroid",
      predictedGroup_Un %in% Other ~ "Other",
      TRUE ~ NA_character_
    )
  )

AML_scATA_meta$Classified_Celltype <- AML_scATA_meta$predictedGroup_Un

maincelltypes <- c("HSPC", "Myeloid", "Lymphoid", "Erythroid", "Other")
subcelltypes <- c(HSPC, Myeloid, Lymphoid, Erythroid_Basophil, Other)

AML_scRNA_meta_DX <- AML_scRNA_meta %>%
  dplyr::filter(grepl("DX", Library_ID))

AML_scRNA_meta_REL <- AML_scRNA_meta %>%
  dplyr::filter(grepl("REL", Library_ID))

AML_scATA_meta_DX <- AML_scATA_meta %>%
  dplyr::filter(grepl("DX", Sample))

AML_scATA_meta_REL <- AML_scATA_meta %>%
  dplyr::filter(grepl("REL", Sample))

AML_scRNA_meta_DX$Classified_CelltypeQ <- paste(AML_scRNA_meta_DX$Classified_Celltype, "DX", sep = "_")
AML_scRNA_meta_REL$Classified_CelltypeQ <- paste(AML_scRNA_meta_REL$Classified_Celltype, "REL", sep = "_")
AML_scRNA_meta_DX$maincelltypeQ <- paste(AML_scRNA_meta_DX$maincelltype, "DX", sep = "_")
AML_scRNA_meta_REL$maincelltypeQ <- paste(AML_scRNA_meta_REL$maincelltype, "REL", sep = "_")

AML_scATA_meta_DX$Classified_CelltypeQ <- paste(AML_scATA_meta_DX$Classified_Celltype, "DX", sep = "_")
AML_scATA_meta_REL$Classified_CelltypeQ <- paste(AML_scATA_meta_REL$Classified_Celltype, "REL", sep = "_")
AML_scATA_meta_DX$maincelltypeQ <- paste(AML_scATA_meta_DX$maincelltype, "DX", sep = "_")
AML_scATA_meta_REL$maincelltypeQ <- paste(AML_scATA_meta_REL$maincelltype, "REL", sep = "_")

AML_scRNA_meta_DX_REL <- rbind(AML_scRNA_meta_DX, AML_scRNA_meta_REL)
AML_scATA_meta_DX_REL <- rbind(AML_scATA_meta_DX, AML_scATA_meta_REL)

# samples for lineage switching
selected_samples_id <- c("AML4","AML11", "AML16",  "AML17", "AML18", "AML24")
# scRNA
AML_scRNA_CSM_DX_select_list <- AML_scRNA_CSM_DX_list[names(AML_scRNA_CSM_DX_list) %in% selected_samples_id]
names(AML_scRNA_CSM_DX_select_list) <- paste(selected_samples_id, "DX", sep = "_")
AML_scRNA_CSM_REL_select_list <- AML_scRNA_CSM_REL_list[names(AML_scRNA_CSM_REL_list) %in% selected_samples_id]
names(AML_scRNA_CSM_REL_select_list) <- paste(selected_samples_id, "REL", sep = "_")
AML_scRNA_CSM_DX_REL_select_list <- c(AML_scRNA_CSM_DX_select_list, AML_scRNA_CSM_REL_select_list)

# scATA
AML_scATA_CSM_DX_select_list <- AML_scATA_CSM_DX_list[names(AML_scATA_CSM_DX_list) %in% selected_samples_id]
names(AML_scATA_CSM_DX_select_list) <- paste(selected_samples_id, "DX", sep = "_")
AML_scATA_CSM_REL_select_list <- AML_scATA_CSM_REL_list[names(AML_scATA_CSM_REL_list) %in% selected_samples_id]
names(AML_scATA_CSM_REL_select_list) <- paste(selected_samples_id, "REL", sep = "_")
AML_scATA_CSM_DX_REL_select_list <- c(AML_scATA_CSM_DX_select_list, AML_scATA_CSM_REL_select_list)

data_mutationCelltype_mutSites_scRNA_byStage <- get_scRNA_mutSites_byStages(AML_scRNA_CSM_DX_REL_select_list, AML_scRNA_meta_DX_REL)
data_mutationCelltype_mutSites_scATA_byStage <- get_scATA_mutSites_byStages(AML_scATA_CSM_DX_REL_select_list, AML_scATA_meta_DX_REL)

data_mutationCelltype_similarity_scRNA_byStage_list <- list()
data_mutationCelltype_similarity_scATA_byStage_list <- list()
common_sample_pairs <- selected_samples_id
common_sample_stages <- c("DX", "REL")
ct_forComparision <- subcelltypes

for (sample in common_sample_pairs){
  message(sample)
  row_col_data_Names <- c(paste(ct_forComparision, "DX", sep = "_"), paste(ct_forComparision, "REL", sep = "_"))
  mut_cellTypeAnn_similarity <- as.data.frame(matrix(data = 0, nrow = length(row_col_data_Names), 
                                                    ncol = length(row_col_data_Names)))
  rownames(mut_cellTypeAnn_similarity) <- row_col_data_Names
  colnames(mut_cellTypeAnn_similarity) <- row_col_data_Names
  for (i in 1:nrow(mut_cellTypeAnn_similarity)){
    ct_stage_i <- rownames(mut_cellTypeAnn_similarity)[i]
    ct_i <- strsplit(ct_stage_i, "_")[[1]][1]
    stage_i <- strsplit(ct_stage_i, "_")[[1]][2]
    mut_site_i <- data_mutationCelltype_mutSites_scRNA_byStage[[paste(sample, stage_i, sep = "_")]][[ct_i]]
    for (j in 1:ncol(mut_cellTypeAnn_similarity)){
      ct_stage_j <- colnames(mut_cellTypeAnn_similarity)[j]
      ct_j <- strsplit(ct_stage_j, "_")[[1]][1]
      stage_j <- strsplit(ct_stage_j, "_")[[1]][2]
      mut_site_j <- data_mutationCelltype_mutSites_scRNA_byStage[[paste(sample, stage_j, sep = "_")]][[ct_j]]
      similarity_score <- ifelse(
        length(unique(c(mut_site_i, mut_site_j))) == 0,
        0,
        length(unique(intersect(mut_site_i, mut_site_j))) / length(unique(c(mut_site_i, mut_site_j)))
      )
      if (length(mut_site_i) == 1 | length(mut_site_j) == 1){
        similarity_score <- 0
      }
      
      mut_cellTypeAnn_similarity[i, j] <- similarity_score
    }
  }
  data_mutationCelltype_similarity_scRNA_byStage_list[[sample]] <- mut_cellTypeAnn_similarity
}

for (sample in common_sample_pairs){
  message(sample)
  row_col_data_Names <- c(paste(ct_forComparision, "DX", sep = "_"), paste(ct_forComparision, "REL", sep = "_"))
  mut_cellTypeAnn_similarity <- as.data.frame(matrix(data = 0, nrow = length(row_col_data_Names), 
                                                    ncol = length(row_col_data_Names)))
  rownames(mut_cellTypeAnn_similarity) <- row_col_data_Names
  colnames(mut_cellTypeAnn_similarity) <- row_col_data_Names
  for (i in 1:nrow(mut_cellTypeAnn_similarity)){
    ct_stage_i <- rownames(mut_cellTypeAnn_similarity)[i]
    ct_i <- strsplit(ct_stage_i, "_")[[1]][1]
    stage_i <- strsplit(ct_stage_i, "_")[[1]][2]
    mut_site_i <- data_mutationCelltype_mutSites_scATA_byStage[[paste(sample, stage_i, sep = "_")]][[ct_i]]
    for (j in 1:ncol(mut_cellTypeAnn_similarity)){
      ct_stage_j <- colnames(mut_cellTypeAnn_similarity)[j]
      ct_j <- strsplit(ct_stage_j, "_")[[1]][1]
      stage_j <- strsplit(ct_stage_j, "_")[[1]][2]
      mut_site_j <- data_mutationCelltype_mutSites_scATA_byStage[[paste(sample, stage_j, sep = "_")]][[ct_j]]
      similarity_score <- ifelse(
        length(unique(c(mut_site_i, mut_site_j))) == 0,
        0,
        length(unique(intersect(mut_site_i, mut_site_j))) / length(unique(c(mut_site_i, mut_site_j)))
      )
      if (length(mut_site_i) == 1 | length(mut_site_j) == 1){
        similarity_score <- 0
      }
      mut_cellTypeAnn_similarity[i, j] <- similarity_score
    }
  }
  data_mutationCelltype_similarity_scATA_byStage_list[[sample]] <- mut_cellTypeAnn_similarity
}


corrplot(as.matrix(data_mutationCelltype_similarity_scRNA_byStage_list$AML4), order = "original", type = 'lower',
         method = 'color',  col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), title = "AML4")

corrplot(as.matrix(data_mutationCelltype_similarity_scRNA_byStage_list$AML16), order = "original", type = 'lower',
         method = 'color',  col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), title = "AML16")

corrplot(as.matrix(data_mutationCelltype_similarity_scATA_byStage_list$AML4), order = "original", type = 'upper',
         method = 'color',  col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), title = "AML4")

corrplot(as.matrix(data_mutationCelltype_similarity_scATA_byStage_list$AML16), order = "original", type = 'upper',
         method = 'color',  col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), title = "AML16")

corrplot(as.matrix(data_mutationCelltype_similarity_scRNA_byStage_list$AML17), order = "original", type = 'lower',
         method = 'color',  col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), title = "AML17")

corrplot(as.matrix(data_mutationCelltype_similarity_scRNA_byStage_list$AML11), order = "original", type = 'lower',
         method = 'color',  col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), title = "AML11")

corrplot(as.matrix(data_mutationCelltype_similarity_scRNA_byStage_list$AML18), order = "original", type = 'lower',
         method = 'color',  col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), title = "AML18")

corrplot(as.matrix(data_mutationCelltype_similarity_scRNA_byStage_list$AML24), order = "original", type = 'lower',
         method = 'color',  col = rev(colorRampPalette(brewer.pal(11, "RdBu"))(100)), col.lim = c(0, 1), title = "AML24")

corrplot(as.matrix(data_mutationCelltype_similarity_scATA_byStage_list$AML17), order = "original", type = 'upper',
         method = 'color',  col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), title = "AML17")

corrplot(as.matrix(data_mutationCelltype_similarity_scATA_byStage_list$AML11), order = "original", type = 'upper',
         method = 'color',  col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), title = "AML11")

corrplot(as.matrix(data_mutationCelltype_similarity_scATA_byStage_list$AML18), order = "original", type = 'upper',
         method = 'color',  col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), title = "AML18")

corrplot(as.matrix(data_mutationCelltype_similarity_scATA_byStage_list$AML24), order = "original", type = 'upper',
         method = 'color',  col = colorRampPalette(brewer.pal(11, "RdBu"))(100), col.lim = c(0, 1), title = "AML24")
```

# sessionInfo
```{r sessionInfo}
sessionInfo()
```



